### 2016-10-05 manuscript/manuscript.tex
very initial ideas

%\documentclass[manuscript]{geophysics}
%\usepackage[nomarkers]{endfloat}

% For a two-column format that looks like the final product, uncomment bellow
% and comment the two lines above
\documentclass[paper,twocolumn,twoside]{geophysics}

\usepackage{graphicx}

\usepackage{amsmath}
\usepackage{array}

\usepackage{booktabs}
\usepackage{color}
\usepackage{epstopdf}

\newcommand{\vect}[1]{\mathbf{#1}}
\newcommand{\mat}[1]{\mathbf{#1}}
\newcommand{\comp}[1]{#1^{\alpha\beta}}
\newcommand{\norm}[1]{\left|\left|#1\right|\right|}

\begin{document}

\title{Gravitational disturbance}

% manuscript number
\ms{GEO-2015-XXXX}

\address{
    \footnotemark[1]
    Observat\'{o}rio Nacional, Rio de Janeiro, Brazil \\
    \footnotemark[2]
    Universidade do Estado do Rio de Janeiro, Rio de Janeiro, Brazil \\
    e-mails: vandscoelho@gmail.com, valcris@on.br

}

\author{Vanderlei C. Oliveira Jr.\footnotemark[1],
Leonardo Uieda\footnotemark[2] and
Val\'{e}ria C. F. Barbosa\footnotemark[1]}

\lefthead{Oliveira Jr., Uieda and Barbosa}
\righthead{Gravitational disturbance}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{abstract}

Gravity anomalies have long been used by geophysicists for the 
purpose of determining density distributions in subsurface.

However, gravity anomalies

In this paper, we discuss the fundamental concepts

\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}

The resultant of gravitational force and centrifugal force acting 
on a body at rest on the Earth's surface is called gravity vector
and its intensity is called simply gravity 
\citep{hofmann-wellenhof-moritz2005}.
In the case of gravimetry on moving platforms (e.g., airplanes,
helicopters, marine vessels), there are additional
non-gravitational accelerations due to the vehicle motion, 
such as Coriolis acceleration and high-frequency vibrations 
\citep{glennie-etal2000, nabighian-etal2005-grav, baumann_etal2012}.
Gravity is the most commonly measured quantity in
geophysical surveying.
Geophysicists use gravity for estimating the Earth's 
internal density distribution whereas geodesists use
gravity to estimate the geoid \citep{li2001}
Hence, geophysicists are usually interested 
in the gravitational component of the observed gravity 
which is produced by the Earth's internal 
density distribution.
The first step of the procedure for isolating the
gravitational component of the observed gravity 
consists in removing the non-gravitational effects due 
to the vehicle motion and also the time variations
such as Earth tides, instrumental drift and barometric 
pressure changes, for example.
If these effects are properly removed, the resultant 
gravity data can be considered as the sum of a 
centrifugal component due to the Earth's rotation and
a gravitational component produced by the whole Earth's
internal density distribution.
%For simplicity, we will henceforth call this 
%resultant gravity corrected from non-gravitational 
%effects due to the vehicle motion and also time variations
%as simply gravity.

In applied geophysics, gravity surveys are usually conducted 
over small areas on the Earth's surface for the purpose of 
characterizing geological structures located within the
crust and upper mantle.
Consequently, geophysicists are only interested in the
particular gravitational component of gravity which 
is produced by these target geological structures.
The isolation of this particular gravitational component 
and its subsequent use for estimating density 
distributions related to geological structures in subsurface 
are the main goals in applied geophysics 
\citep{blakely1996}.

In this paper, we discuss the theoretical definition
of the gravitational disturbance based on the 
well-established concepts of normal Earth, normal gravity, 
gravity disturbance and gravity anomalies.
The concept of gravitational disturbance plays an
important role in the discussions about 
the use of gravity disturbance or gravity anomalies
in geophysical applications.
It seems that this theoretical issue has been 
debated within the scientific community from a 
more geodetic than geophysical point of view
\citep{lafehr1991, chapin1996, li2001, fairhead2003,
hackney-featherstone2003, hinze2005}.
Our reasoning suggests that the gravity 
disturbance is conceptually more appropriated than 
gravity anomalies for approximating
the gravitational disturbance and use it for estimating 
density distributions in subsurface.

\section{Normal Earth and normal gravity}

Traditionally, the Earth's gravity field is approximated 
by the gravity field produced by a geocentric and rigid ellipsoid 
of revolution which has the minor axis $b$ 
coincident with the mean rotating axis of the Earth $Z$, the 
same total mass (including the atmosphere) and also the
same angular velocity of the Earth (Figure \ref{fig:fig1}).
Another characteristic of this model is that its
limiting surface coincides with a particular equipotential 
of its own gravity field.
This model is called as normal Earth and its gravity
field is called normal gravity field
\citep{vanicek1987, hofmann-wellenhof-moritz2005, torge2014}.

It is worth noting that, although the normal Earth has the
same total mass (including the atmosphere) of the Earth,
its internal density distribution is unknown.

The only condition imposed on its internal density
distribution is that it produces a normal gravity field
having a particular equipotential which coincides
with its limiting surface.

For convenience, we denote any density distribution 
satisfying this condition as a normal density distribution.

The normal Earth gives rise to the geodetic coordinate system.
In this coordinate system, the position of a point $P$
is defined by the geometric height $h$, geodetic latitude
$\varphi$ and longitude $\lambda$ (Figure \ref{fig:fig1}).
Geodetic coordinates $(h, \varphi, \lambda)$ can be easily 
converted into geocentric Cartesian coordinates $(X, Y, Z)$
(Figure \ref{fig:fig1}).
The plane containing the point $P$, the axis $Z$ and
the origin $O$ of this geocentric Cartesian coordinate system
is called meridian plane (gray plane in Figure \ref{fig:fig1}).

Similarly to the gravity vector and gravity, 
the resultant of the virtual 
gravitational and centrifugal forces exerted by the normal
Earth on a body at rest at a point $P$ is called 
normal gravity vector and its intensity is called 
simply normal gravity.

\section{Gravity disturbance}

It is worth noting that, by definition, 
the centrifugal component of the normal gravity field is
equal to the centrifugal component of the Earth's gravity
field if they are evaluated at the same point.

Then, the differences between the gravity vector
(corrected from non-gravitational effects due to the vehicle motion) 
and the normal gravity vector, at the same point, represents a purely 
gravitational and consequently harmonic disturbing field.
For convenience, we denote this disturbing field as
gravitational disturbance.

It seems logical to expected that the gravitational disturbance 
is caused by contrasts between the actual internal 
density distribution of the Earth and the internal density 
distribution of the normal Earth.
In applied geophysics, these density differences are generally 
called ``anomalous masses" (e.g., \citeauthor{hammer1945}, 
\citeyear{hammer1945}; \citeauthor{lafehr1965}, \citeyear{lafehr1965})
or ``gravity sources" (e.g., \citeauthor{blakely1996}, 
\citeyear{blakely1996}). Here, we opted for using the second term.

The difference between the observed gravity and the
normal gravity, at the same point, is called gravity disturbance
\citep{hofmann-wellenhof-moritz2005}.
Notice that the gravity disturbance is not equivalent to the
magnitude of the difference between the gravity vector
and the normal gravity vector at the same point.
As properly pointed out by \citet{hackney-featherstone2003},
the gravity disturbance is a very-well established quantity in geodesy,
but appears to be less well known in geophysics.

The gravity anomaly is defined as the difference
between the gravity at the geoid and the normal gravity at the ellipsoid
and is the commonly used quantity in applied geophysics.
Different gravity anomalies can be calculated, depending on the
corrections applied to them \citep{blakely1996, hofmann-wellenhof-moritz2005}.
These corrections are usually called gravity reductions.
For example, the Free-air anomaly is an approximation of the
gravity disturbance whereas the Bouguer anomaly
is an approximation of the terrain corrected gravity disturbance.
The last one is commonly used by geophysicists as the
gravitational effect produced by the gravity sources.
Although this approximation is valid for most practical applications,
it is important to bear in mind not only the terminology 
changes, but also the conceptual assumptions.

There was a certain lack of comprehension regarding the
geophysical meaning of gravity anomalies until the
mid 90's.
As properly pointed out by \citet{chapin1996} at that time, 
``although the corrections which bring about a Bouguer 
gravity anomaly are well established, the reasons for doing
them are not well understood. One cause of this common 
misunderstanding is that the subject has been poorly presented in
many of the basic texts".
In his seminal book, \citet{blakely1996} brought some light
on the geophysical meaning of gravity anomalies from the 
perspective of applied geophysics. \citet{blakely1996} correctly
defined gravity sources as density contrasts between the actual
internal density distribution of the Earth and the internal density
distribution of the normal Earth.
However, he did not stress that, by removing the normal gravity 
evaluated on the ellipsoid from the gravity measured 
on the Earth's surface, the remaining disturbing field will reflect 
not only the effect produced by the gravity sources, but also a
small combination of gravitational and centrifugal effects.
This additional, non-harmonic and undesired effect is 
due to the calculation of the normal gravity at the surface of the ellipsoid
instead of at the same points where the gravity is measured.

- Mencionar tambem que anomalias de gravidade requerem o
valor da gravidade dentro das fontes

\section{Mathematical description of the gravity disturbance in a local coordinate system}

In a local- or regional-gravity study, 
geophysicists commonly use a topocentric Cartesian coordinate system
with the $x$ axis pointing to North, the $y$ axis pointing to East and 
the $z$ axis with the same direction as the normal $\hat{\mathbf{n}}_{i}$
to the ellipsoid, but pointing downward (Figure \ref{fig:fig1}).
In this coordinate system, the observed gravity vector
$\mathbf{g}^{o}_{i}$, at the point $(x_{i}, y_{i}, z_{i})$, 
$i = 1, ..., N$, can be represented by
\begin{equation}
\mathbf{g}^{o}_{i} = \boldsymbol{\gamma}_{i} + \Delta \mathbf{g}^{o}_{i} \: ,
\label{eq:gravity-vector}
\end{equation}
where $\boldsymbol{\gamma}_{i}$ and $\Delta \mathbf{g}^{o}_{i}$
are, respectively, the normal gravity vector and a
disturbing gravitational attraction produced by the anomalous 
masses at the point $(x_{i}, y_{i}, z_{i})$.

By definition, the gravity disturbance $\delta g^{o}_{i}$,
at the point $(x_{i}, y_{i}, z_{i})$, is given by
\begin{equation}
\delta g^{o}_{i} =  \| \mathbf{\delta g}^{o}_{i} \| - \| \boldsymbol{\gamma}_{i} \| \: ,
\label{eq:gravity-disturbance}
\end{equation}
where $\| \mathbf{\delta g}^{o}_{i} \|$ and $\| \boldsymbol{\gamma}_{i} \|$ are,
respectively, the observed gravity and the normal gravity at the
point $(x_{i}, y_{i}, z_{i})$.
Fortunately, the condition $\| \boldsymbol{\gamma}_{i} \| \gg \| \Delta \mathbf{g}^{o}_{i} \|$
is met at all points located above or on the Earth's surface.
By combining this condition and the definition of observed gravity vector
given in equation \ref{eq:gravity-vector}, we can approximate the observed gravity
$\| \mathbf{\delta g}^{o}_{i} \|$ (equation \ref{eq:gravity-disturbance})
by a first order Taylor's expansion as follows:
\begin{equation}
\| \mathbf{\delta g}^{o}_{i} \| \approx \| \boldsymbol{\gamma}_{i} \| + 
\hat{\boldsymbol{\gamma}}_{i}^{\top} \Delta \mathbf{g}^{o}_{i} \: ,
\label{eq:gobs-approx}
\end{equation}
where $\hat{\boldsymbol{\gamma}}_{i}$ is a unit vector with the same 
direction as the normal gravity vector $\boldsymbol{\gamma}_{i}$ at
the point $(x_{i}, y_{i}, z_{i})$.
This approximation is largely used in applied geophysics for representing
total-field anomalies (e.g., \citeauthor{blakely1996}, \citeyear{blakely1996}).
Notice that, local- or regional-gravity studies, the unit
vector $\hat{\boldsymbol{\gamma}}_{i}$ (equation \ref{eq:gobs-approx}) 
coincides with the $z$ axis
of the local Cartesian coordinate system defined
at the beginning of this section. Consequently,
by using the approximation defined in equation \ref{eq:gobs-approx}, 
the gravity disturbance (equation \ref{eq:gravity-disturbance}) can be 
rewritten as follows \begin{equation}
\delta g^{o}_{i} \approx \hat{\mathbf{z}}^{\top} \Delta \mathbf{g}^{o}_{i} \: ,
\label{eq:gravity-disturbance-approx}
\end{equation}
where $\hat{\mathbf{z}}^{\top} = \left[ \, 0 \; 0 \; 1 \right]$.
According to equation \ref{eq:gravity-disturbance-approx},
the gravity disturbance $\delta g^{o}_{i}$ (equation \ref{eq:gravity-disturbance})
represents the vertical component of the gravitational attraction exerted by
the gravity sources at the point $(x_{i}, y_{i}, z_{i})$.
As a consequence, the gravity disturbance produced by
a homogeneous gravity source can be represented 
by the following harmonic function
\begin{equation}
d^{o}_{i} = k_{g} \, G \, \rho \, \partial_{z} \phi_{i} \: ,
\label{eq:gz-local}
\end{equation}
where $G$ is the Newtonian constant of gravitation
(in $m^{3} / (kg \, s^{2})$),
$k_{g} = 10^{5}$ is a constant factor
transforming from $m/s^{2}$ to milligal (mGal),
and $\partial_{z} \phi_{i}$ is a harmonic function representing the first 
derivative, evaluated at the observation point $(x_{i},y_{i},z_{i})$, 
$i = 1, ..., N$, of the function 
\begin{equation}
\phi(x,y,z) = \int\int\limits_{v}\int \frac{1}{\sqrt{(x - x^{\prime})^{2} + 
(y - y^{\prime})^{2} + (z - z^{\prime})^{2}}} \: dv
\label{eq:phi}
\end{equation}
with respect to the variable $z$. The 
integral is conducted over the coordinates $x^{\prime}$, $y^{\prime}$ 
and $z^{\prime}$ within the volume $v$ of the gravity source.
This equation can be easily generalized for the case
of multiple gravity sources.

\section{Geophysical arguments for using the gravity disturbance}

Almost all interpretation techniques assume, implicitly or
directly, that the gravity data is harmonic (e.g., 
upward/downward continuation, 
data processing with equivalent layer,
conversions between gravity and magnetic data,
computation of vertical derivatives via Fourier and Hilbert
transforms).

Consequently, they implicitly or directly assume that
the gravity data approximates the gravitational disturbance.

Almost all forward modelling techniques compute
the vertical component of the gravitational attraction 
exerted by the geological bodies at the observation points.

Hence, almost all geophysicists implicitly compute 
the vertical component of the gravitational disturbance.

The gravity anomaly is defined as the difference between 
the gravity $\| \mathbf{g}_{P} \|$, at a point $P$ on the
Geoid, and the normal gravity $\| \boldsymbol{\gamma}_{Q} \|$,
on the reference elipsoid, where $P$ and $Q$ have the same
geodetic latitude and longitude.

Consequently, the gravity anomaly is a function of
the geodetic latitude and longitude only and cannot
be calculated at arbitrary heights.

Gravity anomalies require the computation of gravity
within the topographic masses, where the gravitational
disturbance is not harmonic.

% The bibliography
\bibliographystyle{seg}
\bibliography{bib-file}

\clearpage

% Tables
%\begin{table}
%    \centering
%    \renewcommand{\arraystretch}{1.5}
%    \begin{tabular}{>{\centering\arraybackslash}m{1.1in} 
%                    >{\centering\arraybackslash}m{1.1in} 
%                    >{\centering\arraybackslash}m{1.1in}
%                    >{\centering\arraybackslash}m{1.1in}}
%        \toprule        
%        \textbf{Author(s)} & \textbf{Equivalent sources} & 
%            \textbf{Harmonic function} $g_{ij}$ &
%            \textbf{Observed data}\\
%        \midrule
%        \citet{dampney1969} & point masses & 
%            vertical component of the gravitational attraction &
%            Bouguer anomaly \\ \hline
%        \citet{emilia1973} & lines of dipoles & total-field anomaly &
%            total-field anomaly \\ \hline
%        \citet{hansen-miyazaki1984} & uniformly-magnetized plane faces & 
%            magnetic scalar potential &
%            total-field anomaly \\ \hline
%        \citet{silva1986} & doublets & total-field anomaly &
%        total-field anomaly \\ \hline
%        \citet{cordell1992} & point masses & gravitational potential &
%            Bouguer anomaly \\ \hline
%        \citet{guspi-novara2009} & generic point sources producing a higher-order 
%            potential & second-directional derivative of the higher-order 
%            potential & total-field anomaly \\ \hline
%        \citet{barnes-lumley2011} & rectangular prisms & second derivatives
%            of the gravitational potential & gravity gradient \\ \hline
%        \citet{oliveirajr-etal2013} & dipoles & total-field anomaly & 
%            total-field anomaly \\ \hline
%        \citet{li-nabighian-oldenburg2014} & rectangular prisms & total-field 
%            anomaly & total-field anomaly \\
%        \bottomrule
%    \end{tabular}
%    \caption{Observed potential-field data and the
%    mathematical/physical meaning of the harmonic functions $g_{ij}$
%    (equation \ref{eq:dpi}) describing the potential field produced 
%    by the equivalent sources used by several authors.}
%    \label{tab:equivalent-sources-gij-literature}
%\end{table}
%\clearpage

% Figures generated by using Inkscape

\begin{figure}
    \centering
    \includegraphics[width=0.3\textwidth]{figures/geodetic_coordinates.png}
    \caption{Schematic representation of the geodetic coordinate
    system.}
  \label{fig:fig1}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.3\textwidth]{figures/gravitational_disturbance.png}
    \caption{Schematic representation of the geodetic coordinate
    system.}
  \label{fig:fig2}
\end{figure}

\end{document}

### 2017-03-10 manuscript/manuscript.tex
new references
include the reference of Forsberg 1984 and the term density anomalies

@@
 In this paper, we discuss the fundamental concepts
+
+URGENTE: \citep{marussi1974}, \citep{torge2012}, section 4.2.1
 
@@
 field is called normal gravity field
-\citep{vanicek1987, hofmann-wellenhof-moritz2005, torge2014}.
+\citep{vanicek1987, hofmann-wellenhof-moritz2005, torge2012}.
 
@@
 called ``anomalous masses" (e.g., \citeauthor{hammer1945}, 
-\citeyear{hammer1945}; \citeauthor{lafehr1965}, \citeyear{lafehr1965})
+\citeyear{hammer1945}; \citeauthor{lafehr1965}, \citeyear{lafehr1965}),
+``density anomalies" (e.g., \citeauthor{forsberg1984}, \citeyear{forsberg1984})
 or ``gravity sources" (e.g., \citeauthor{blakely1996}, 
-\citeyear{blakely1996}). Here, we opted for using the second term.
+\citeyear{blakely1996}). Here, we opted for using the last term.
 

### 2018-03-23 manuscript/manuscript.tex
remove the term gravitational disturbance

@@
-%\documentclass[manuscript]{geophysics}
-%\usepackage[nomarkers]{endfloat}
+\documentclass[manuscript]{geophysics}
+\usepackage[nomarkers]{endfloat}
 
@@
 % and comment the two lines above
-\documentclass[paper,twocolumn,twoside]{geophysics}
+%\documentclass[paper,twocolumn,twoside]{geophysics}
 
@@
 \citep{hofmann-wellenhof-moritz2005}.
+
 In the case of gravimetry on moving platforms (e.g., airplanes,
@@
 \citep{glennie-etal2000, nabighian-etal2005-grav, baumann_etal2012}.
-Gravity is the most commonly measured quantity in
-geophysical surveying.
+
+Gravity is the most commonly measured quantity in geophysical surveying.
 Geophysicists use gravity for estimating the Earth's 
 internal density distribution whereas geodesists use
-gravity to estimate the geoid \citep{li2001}
+gravity to estimate the geoid \citep{li2001}.
+
 Hence, geophysicists are usually interested 
@@
 density distribution.
-The first step of the procedure for isolating the
-gravitational component of the observed gravity 
-consists in removing the non-gravitational effects due 
+
+The first step of the procedure for isolating this 
+gravitational component consists in removing the non-gravitational effects due 
 to the vehicle motion and also the time variations
@@
 pressure changes, for example.
+
 If these effects are properly removed, the resultant 
@@
 internal density distribution.
+
 %For simplicity, we will henceforth call this 
@@
 crust and upper mantle.
+
 Consequently, geophysicists are only interested in the
@@
 is produced by these target geological structures.
+
 The isolation of this particular gravitational component 
@@
 
-In this paper, we discuss the theoretical definition
-of the gravitational disturbance based on the 
-well-established concepts of normal Earth, normal gravity, 
-gravity disturbance and gravity anomalies.
-The concept of gravitational disturbance plays an
-important role in the discussions about 
-the use of gravity disturbance or gravity anomalies
-in geophysical applications.
+In this paper, we present a discussion based on the
+well-established concepts of normal Earth, normal gravity, gravity disturbance and gravity anomalies aiming at bringing some
+light to the following question: in geophysical applications,
+should we use gravity disturbance or gravity anomaly?
+
 It seems that this theoretical issue has been 
@@
 hackney-featherstone2003, hinze2005}.
+
 Our reasoning suggests that the gravity 

### 2018-04-06 manuscript/manuscript.tex
review up to the normal gravity section

@@
 The resultant of gravitational force and centrifugal force acting 
-on a body at rest on the Earth's surface is called gravity vector
-and its intensity is called simply gravity 
-\citep{hofmann-wellenhof-moritz2005}.
+on a body at rest on the Earth's surface is called 
+\textit{gravity vector} and its intensity is called simply 
+\textit{gravity} \citep{hofmann-wellenhof-moritz2005}.
 
@@
 
-Gravity is the most commonly measured quantity in geophysical surveying.
 Geophysicists use gravity for estimating the Earth's 
@@
 
+%In applied geophysics, gravity surveys are usually conducted 
+%over local or regional areas on the Earth's surface for the purpose of 
+%characterizing geological structures located within the
+%crust and upper mantle.
+%
+%Consequently, geophysicists are only interested in the
+%particular gravitational component of gravity which 
+%is produced by these target geological structures.
+
 The first step of the procedure for isolating this 
-gravitational component consists in removing the non-gravitational effects due 
-to the vehicle motion and also the time variations
+gravitational component consists in removing the non-gravitational 
+effects due to the vehicle motion and also the time variations 
 such as Earth tides, instrumental drift and barometric 
@@
 internal density distribution.
-
-%For simplicity, we will henceforth call this 
-%resultant gravity corrected from non-gravitational 
-%effects due to the vehicle motion and also time variations
-%as simply gravity.
-
-In applied geophysics, gravity surveys are usually conducted 
-over small areas on the Earth's surface for the purpose of 
-characterizing geological structures located within the
-crust and upper mantle.
-
-Consequently, geophysicists are only interested in the
-particular gravitational component of gravity which 
-is produced by these target geological structures.
 
@@
 of its own gravity field.
-This model is called as normal Earth and its gravity
-field is called normal gravity field
+This model is called as \textit{normal Earth} 
 \citep{vanicek1987, hofmann-wellenhof-moritz2005, torge2012}.
@@
 The only condition imposed on its internal density
-distribution is that it produces a normal gravity field
+distribution is that it produces a gravity field
 having a particular equipotential which coincides
@@
 For convenience, we denote any density distribution 
-satisfying this condition as a normal density distribution.
-
-The normal Earth gives rise to the geodetic coordinate system.
-In this coordinate system, the position of a point $P$
-is defined by the geometric height $h$, geodetic latitude
-$\varphi$ and longitude $\lambda$ (Figure \ref{fig:fig1}).
-Geodetic coordinates $(h, \varphi, \lambda)$ can be easily 
-converted into geocentric Cartesian coordinates $(X, Y, Z)$
-(Figure \ref{fig:fig1}).
-The plane containing the point $P$, the axis $Z$ and
-the origin $O$ of this geocentric Cartesian coordinate system
-is called meridian plane (gray plane in Figure \ref{fig:fig1}).
+satisfying this condition as a \textit{normal density distribution}.
 
@@
 Earth on a body at rest at a point $P$ is called 
-normal gravity vector and its intensity is called 
-simply normal gravity.
+\textit{normal gravity vector} and its intensity is called 
+simply \textit{normal gravity}.
+
+The normal Earth gives rise to the \textit{geodetic coordinate system}.
+In this coordinate system, the position of a point $P$
+is defined by the \textit{geometric height} $h$, 
+\textit{geodetic latitude}
+$\varphi$ and \textit{longitude} $\lambda$ (Figure \ref{fig:fig1}).
+Geodetic coordinates $(h, \varphi, \lambda)$ can be easily 
+converted into geocentric Cartesian coordinates $(X, Y, Z)$
+(Figure \ref{fig:fig1}).
+
+The plane containing the point $P$, the axis $Z$ and
+the origin $O$ of this \textit{geocentric Cartesian coordinate system}
+is called meridian plane (gray plane in Figure \ref{fig:fig1}).
 
@@
 corrections applied to them \citep{blakely1996, hofmann-wellenhof-moritz2005}.
+
 These corrections are usually called gravity reductions.
@@
 By definition, the gravity disturbance $\delta g^{o}_{i}$,
-at the point $(x_{i}, y_{i}, z_{i})$, is given by
+at the point $(x_{i}, y_{i}, z_{i})$, is given by \citep{hofmann-wellenhof-moritz2005}:
 \begin{equation}
@@
     \centering
-    \includegraphics[width=0.3\textwidth]{figures/geodetic_coordinates.png}
+    \includegraphics[width=0.7\textwidth]{figures/geodetic_coordinates.png}
     \caption{Schematic representation of the geodetic coordinate
-    system.}
+    system defined by an oblate ellipsoid with semi-minor axis $b$, 
+    coincident with the mean Earth's rotation axis, and a semi-major
+    axis $a$. In this coordinate system, the position of a point is
+    determined by the geometric height $h$, the geodetic latitude 
+    $\varphi$ and longitude $\lambda$. The Earth's origin is represented 
+    by $O$, $P$ represents a point $(h, \varphi, \lambda)$ and 
+    $P^{\prime}$ represents the projection of $P$ on the plane $XY$ 
+    (Equatorial plane). The plane containing $O$, $P$ and 
+    $P^{\prime}$ is represented by the gray triangle in (a) and (b). 
+    In (b), the point $Q$ represents the projection of $P$ on the 
+    ellipsoid surface.}
   \label{fig:fig1}
@@
     \centering
-    \includegraphics[width=0.3\textwidth]{figures/gravitational_disturbance.png}
+    \includegraphics[width=0.7\textwidth]{figures/gravitational_disturbance.png}
     \caption{Schematic representation of the geodetic coordinate

### 2018-04-10 manuscript/manuscript.tex
review intro and detect problems with refs

@@
 \textit{gravity} \citep{hofmann-wellenhof-moritz2005}.
-
 In the case of gravimetry on moving platforms (e.g., airplanes,
@@
 \citep{glennie-etal2000, nabighian-etal2005-grav, baumann_etal2012}.
-
 Geophysicists use gravity for estimating the Earth's 
@@
 gravity to estimate the geoid \citep{li2001}.
-
 Hence, geophysicists are usually interested 
-in the gravitational component of the observed gravity 
+in the gravitational component of the observed gravity, 
 which is produced by the Earth's internal 
@@
 pressure changes, for example.
-
 If these effects are properly removed, the resultant 
@@
 internal density distribution.
-
 The isolation of this particular gravitational component 
@@
 
-In this paper, we present a discussion based on the
-well-established concepts of normal Earth, normal gravity, gravity disturbance and gravity anomalies aiming at bringing some
+Based on well-established concepts of the literature,
+we present a discussion aiming at bringing some
 light to the following question: in geophysical applications,
-should we use gravity disturbance or gravity anomaly?
-
+should we use the gravity disturbance or gravity anomaly?
 It seems that this theoretical issue has been 
@@
 hackney-featherstone2003, hinze2005}.
-
 Our reasoning suggests that the gravity 
-disturbance is conceptually more appropriated than 
-gravity anomalies for approximating
-the gravitational disturbance and use it for estimating 
-density distributions in subsurface.
+disturbance is more appropriated than gravity anomalies for 
+approximating the gravitational effect produced by the Earth's 
+internal density distribution.
 
@@
 by the gravity field produced by a geocentric and rigid ellipsoid 
-of revolution which has the minor axis $b$ 
+of revolution, which has the minor axis $b$ 
 coincident with the mean rotating axis of the Earth $Z$, the 
@@
 This model is called as \textit{normal Earth} 
-\citep{vanicek1987, hofmann-wellenhof-moritz2005, torge2012}.
+\citep{heiskanen_moritz1967, vanicek1987, hofmann-wellenhof-moritz2005,
+torge2012}.
 

### 2018-04-13 manuscript/geophysics_example.tex
replace the template of geophysics

%\documentclass[paper]{geophysics}
%\documentclass[manuscript,revised]{geophysics}
\documentclass[manuscript]{geophysics}

% An example of defining macros
\newcommand{\rs}[1]{\mathstrut\mbox{\scriptsize\rm #1}}
\newcommand{\rr}[1]{\mbox{\rm #1}}

\begin{document}

\title{An example \emph{Geophysics} article, \\ with a two-line title}

\renewcommand{\thefootnote}{\fnsymbol{footnote}}

\ms{GEO-Example} % manuscript number

\address{
\footnotemark[1]BP UTG, \\
200 Westlake Park Blvd, \\
Houston, TX, 77079 \\
\footnotemark[2]Bureau of Economic Geology, \\
John A. and Katherine G. Jackson School of Geosciences \\
The University of Texas at Austin \\
University Station, Box X \\
Austin, TX 78713-8924}
\author{Joe Dellinger\footnotemark[1] and Sergey Fomel\footnotemark[2]}

\footer{Example}
\lefthead{Dellinger \& Fomel}
\righthead{\emph{Geophysics} example}

\maketitle

\begin{abstract}
  This is an example of using \textsf{geophysics.cls} for writing
  \emph{Geophysics} papers.
\end{abstract}

\section{Introduction}

This is an introduction. \LaTeX\ is a \new{powerful} document
typesetting system \cite[]{lamport}. An excellent reference is
\cite[]{kopka}. The new \textsf{geophysics.cls} class complies with
the \LaTeX2e\ standard. \old{We had something else written here but
  decided to get rid of it}.

\section*{Theory}

This is another section.

\subsection{Equations}

Section headings should be capitalized. Subsection headings should
only have the first letter of the first word capitalized.

Here are examples of equations involving vectors and tensors:
\begin{equation}
\tensor{R} = 
\pmatrix{R_{\rs{XX}} & R_{\rs{YX}} \cr R_{\rs{XY}} & R_{\rs{YY}}} 
=
\tensor{P}_{M\rightarrow R} \; \tensor{D} \; \tensor{P}_{S\rightarrow M}
\;\;\; \tensor{S} \ \ \  ,
\label{SVD}
\end{equation}
and
\begin{equation}
R_{j,m}(\omega) =
\sum_{n=1}^{N} \, \,
P_{j}^{(n)}(\mathbf{x}_R) \, \,
D^{(n)}(\omega) \, \,
P_{m}^{(n)}(\mathbf{x}_S) \ \ \ .
\label{SVDray}
\end{equation}

Note that the macro for the \verb#\tensor# command has been changed to
force tensors to be bold uppercase, in compliance with current SEG
submission standards. This is so that documents typeset to the old
standards will print out according to the new ones: e.g., tensor
$\tensor{t}$ (note converted to uppercase).

\subsection*{Figures}
\renewcommand{\figdir}{Fig} % figure directory

Figure~\ref{fig:waves} shows what it is about.

\plot{waves}{width=\textwidth}
{This figure is specified in the document by \texttt{
    $\backslash$plot\{waves\}\{width=$\backslash$textwidth\}\{This caption.\}}.
}

\subsubsection{Multiplot}

Sometimes it is convenient to put two or more figures from different
files in an array (see Figure~\ref{fig:exph,exgr}). Individual plots
are Figures~\ref{fig:exph} and~\ref{fig:exgr}.

\multiplot{2}{exph,exgr}{width=0.4\textwidth}
{This figure is specified in the document by \newline \texttt{
    $\backslash$multiplot\{2\}\{exph,exgr\}\{width=0.4$\backslash$textwidth\}\{This caption.\}}.
}

The first argument of the \texttt{multiplot} command specifies the
number of plots per row.

\subsection{Tables}

The discussion is summarized in Table~\ref{tbl:example}.

\tabl{example}{This table is specified in the document by \texttt{
    $\backslash$tabl\{example\}\{This caption.\}\{\ldots\}}.
}{
  \begin{center}
    \begin{tabular}{|c|c|c|}
      \hline
      \multicolumn{3}{|c|}{Table Example} \\
      \hline
      migration\rule[-2ex]{0ex}{5ex} & 
      $\omega \rightarrow k_z$ & 
      $k_y^2+k-z^2\cos^2 \psi=4\omega^2/v^2$ \\
      \hline
      \parbox{1in}{zero-offset\\diffraction}\rule[-4ex]{0ex}{8ex} &
      $k_z\rightarrow\omega_0$ &
      $k_y^2+k_z^2=4\omega_0^2/v^2$ \\
      \hline
      DMO+NMO\rule[-2ex]{0in}{5ex} & $\omega\rightarrow\omega_0$ & 
      $\frac{1}{4}
      v^2k_y^2\sin^2\psi+\omega_0^2\cos^2\psi=\omega^2$ \\
      \hline
      radial DMO\rule[-2ex]{0in}{5ex} & $\omega\rightarrow\omega_s$ &
      $\frac{1}{4}v^2k_y^2\sin^2\psi+\omega_s^2=\omega^2$\\
      \hline
      radial NMO\rule[-2ex]{0in}{5ex} & $\omega_s\rightarrow\omega_0$ &
      $\omega_0\cos\psi=\omega_s$\\
      \hline
    \end{tabular}
  \end{center}
}

\section{ACKNOWLEDGMENTS}

I wish to thank Ivan P\v{s}en\v{c}\'{\i}k and Fr\'ed\'eric Billette
for having names with non-English letters in them.  I wish to thank
\cite{Cerveny} for providing an example of how to make a bib file that
includes an author whose name begins with a non-English character and
\cite{forgues96} for providing both an example of referencing a Ph.D.
thesis and yet more non-English characters.

\append{Appendix example}
\label{example}

According to the new SEG standard, appendices come before references.

\begin{equation}
\frac{\partial U}{\partial z} = 
\left\{
  \sqrt{\frac{1}{v^2} - \left[\frac{\partial t}{\partial g}\right]^2} +
  \sqrt{\frac{1}{v^2} - \left[\frac{\partial t}{\partial s}\right]^2}
\right\}
\frac{\partial U}{\partial t}
\label{eqn:partial}
\end{equation}
It is important to get equation~\ref{eqn:partial} right. See also
Appendix~\ref{equations}.

\append[equations]{Another appendix}

\begin{equation}
\frac{\partial U}{\partial z} = 
\left\{
  \sqrt{\frac{1}{v^2} - \left[\frac{\partial t}{\partial g}\right]^2} +
  \sqrt{\frac{1}{v^2} - \left[\frac{\partial t}{\partial s}\right]^2}
\right\}
\frac{\partial U}{\partial t}
\label{eqn:partial2}
\end{equation}
Too lazy to type a different equation but note the numeration.

The error comparison is provided in Figure~\ref{fig:errgrp}.

\sideplot{errgrp}{width=0.8\textwidth}
{This figure is specified in the document by \texttt{
    $\backslash$sideplot\{errgrp\}\{width=0.8$\backslash$text\-width\}\{This caption.\}}.
}

\append{The source of this document}

\verbatiminput{geophysics_example.tex}

\append{The source of the bibliography}

\verbatiminput{example.bib}

\newpage

\bibliographystyle{seg}  % style file is seg.bst
\bibliography{example}

\end{document}

### 2018-04-15 manuscript/letter.tex
solve problems with gji template

--- 2018-04-10 manuscript/manuscript.tex
@@
-\documentclass[manuscript]{geophysics}
-\usepackage[nomarkers]{endfloat}
+% gjilguid2e.tex
+% V2.0 released 1998 December 18
+% V2.1 released 2003 October 7 -- Gregor Hutton, updated the web address for the style files.
 
-% For a two-column format that looks like the final product, uncomment bellow
-% and comment the two lines above
-%\documentclass[paper,twocolumn,twoside]{geophysics}
+%\documentclass[extra,mreferee]{gji}
+\documentclass[extra]{gji}
+\usepackage{timet}
 
+%Extra packages
 \usepackage{graphicx}
 
-\usepackage{amsmath}
-\usepackage{array}
+\title[Gravity disturbance or gravity anomaly?]
+      {Gravity disturbance or gravity anomaly?}
 
-\usepackage{booktabs}
-\usepackage{color}
-\usepackage{epstopdf}
+\author[Oliveira Jr, Barbosa and Uieda]
+{Vanderlei C. Oliveira Jr$^1$\thanks{Sei la, Nacional, Brazil}, Val\'{e}ria C. F. Barbosa$^1$ and Leonardo Uieda$^2$ \\
+	$^1$ Department of Geophysics, Observat\'{o}rio Nacional, Rio de Janeiro, Brazil \\
+	$^2$ Department of Geology and Geophysics, University of Hawaii, Manoa, USA
+}
 
-\newcommand{\vect}[1]{\mathbf{#1}}
-\newcommand{\mat}[1]{\mathbf{#1}}
-\newcommand{\comp}[1]{#1^{\alpha\beta}}
-\newcommand{\norm}[1]{\left|\left|#1\right|\right|}
+\date{Received 2018 Month XX; in original form 2018 Month XX}
+\pagerange{\pageref{firstpage}--\pageref{lastpage}}
+\volume{XXX}
+\pubyear{2018}
+
+%\def\LaTeX{L\kern-.36em\raise.3ex\hbox{{\small A}}\kern-.15em
+%    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}
+%\def\LATeX{L\kern-.36em\raise.3ex\hbox{{\Large A}}\kern-.15em
+%    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}
+% Authors with AMS fonts and mssymb.tex can comment out the following
+% line to get the correct symbol for Geophysical Journal International.
+\let\leqslant=\leq
+
+\newtheorem{theorem}{Theorem}[section]
 
@@
 
-\title{Gravitational disturbance}
+\label{firstpage}
 
-% manuscript number
-\ms{GEO-2015-XXXX}
+\maketitle
 
-\address{
-    \footnotemark[1]
-    Observat\'{o}rio Nacional, Rio de Janeiro, Brazil \\
-    \footnotemark[2]
-    Universidade do Estado do Rio de Janeiro, Rio de Janeiro, Brazil \\
-    e-mails: vandscoelho@gmail.com, valcris@on.br
-
-}
-
-\author{Vanderlei C. Oliveira Jr.\footnotemark[1],
-Leonardo Uieda\footnotemark[2] and
-Val\'{e}ria C. F. Barbosa\footnotemark[1]}
-
-\lefthead{Oliveira Jr., Uieda and Barbosa}
-\righthead{Gravitational disturbance}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{abstract}
-
+\begin{summary}
  Gravity anomalies have long been used by geophysicists for the 
@@
  URGENTE: \citep{marussi1974}, \citep{torge2012}, section 4.2.1
+\end{summary}
 
-\end{abstract}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{keywords}
+ potential fields -- gravity disturbance -- gravity anomaly -- gravity modeling.
+\end{keywords}
 
@@
 such as Coriolis acceleration and high-frequency vibrations 
-\citep{glennie-etal2000, nabighian-etal2005-grav, baumann_etal2012}.
+\citep{glennie-etal2000,nabighian-etal2005-grav,baumann-etal2012}.
 Geophysicists use gravity for estimating the Earth's 
@@
 density distribution.
-
 %In applied geophysics, gravity surveys are usually conducted 
@@
 %is produced by these target geological structures.
-
 The first step of the procedure for isolating this 
@@
 more geodetic than geophysical point of view
-\citep{lafehr1991, chapin1996, li2001, fairhead2003,
-hackney-featherstone2003, hinze2005}.
+\citep{lafehr1991,chapin1996,li2001,fairhead2003,
+hackney-featherstone2003,hinze2005}.
 Our reasoning suggests that the gravity 
@@
 This model is called as \textit{normal Earth} 
-\citep{heiskanen_moritz1967, vanicek1987, hofmann-wellenhof-moritz2005,
+\citep{heiskanen-moritz1967,vanicek1987,hofmann-wellenhof-moritz2005,
 torge2012}.
@@
 
-\section{Gravity disturbance}
+\begin{acknowledgments}
+The authors would like to thank the editor and all the
+reviewers for their criticisms and corrections.
+\end{acknowledgments}
 
-It is worth noting that, by definition, 
-the centrifugal component of the normal gravity field is
-equal to the centrifugal component of the Earth's gravity
-field if they are evaluated at the same point.
-
-Then, the differences between the gravity vector
-(corrected from non-gravitational effects due to the vehicle motion) 
-and the normal gravity vector, at the same point, represents a purely 
-gravitational and consequently harmonic disturbing field.
-For convenience, we denote this disturbing field as
-gravitational disturbance.
-
-It seems logical to expected that the gravitational disturbance 
-is caused by contrasts between the actual internal 
-density distribution of the Earth and the internal density 
-distribution of the normal Earth.
-In applied geophysics, these density differences are generally 
-called ``anomalous masses" (e.g., \citeauthor{hammer1945}, 
-\citeyear{hammer1945}; \citeauthor{lafehr1965}, \citeyear{lafehr1965}),
-``density anomalies" (e.g., \citeauthor{forsberg1984}, \citeyear{forsberg1984})
-or ``gravity sources" (e.g., \citeauthor{blakely1996}, 
-\citeyear{blakely1996}). Here, we opted for using the last term.
-
-The difference between the observed gravity and the
-normal gravity, at the same point, is called gravity disturbance
-\citep{hofmann-wellenhof-moritz2005}.
-Notice that the gravity disturbance is not equivalent to the
-magnitude of the difference between the gravity vector
-and the normal gravity vector at the same point.
-As properly pointed out by \citet{hackney-featherstone2003},
-the gravity disturbance is a very-well established quantity in geodesy,
-but appears to be less well known in geophysics.
-
-The gravity anomaly is defined as the difference
-between the gravity at the geoid and the normal gravity at the ellipsoid
-and is the commonly used quantity in applied geophysics.
-Different gravity anomalies can be calculated, depending on the
-corrections applied to them \citep{blakely1996, hofmann-wellenhof-moritz2005}.
-
-These corrections are usually called gravity reductions.
-For example, the Free-air anomaly is an approximation of the
-gravity disturbance whereas the Bouguer anomaly
-is an approximation of the terrain corrected gravity disturbance.
-The last one is commonly used by geophysicists as the
-gravitational effect produced by the gravity sources.
-Although this approximation is valid for most practical applications,
-it is important to bear in mind not only the terminology 
-changes, but also the conceptual assumptions.
-
-There was a certain lack of comprehension regarding the
-geophysical meaning of gravity anomalies until the
-mid 90's.
-As properly pointed out by \citet{chapin1996} at that time, 
-``although the corrections which bring about a Bouguer 
-gravity anomaly are well established, the reasons for doing
-them are not well understood. One cause of this common 
-misunderstanding is that the subject has been poorly presented in
-many of the basic texts".
-In his seminal book, \citet{blakely1996} brought some light
-on the geophysical meaning of gravity anomalies from the 
-perspective of applied geophysics. \citet{blakely1996} correctly
-defined gravity sources as density contrasts between the actual
-internal density distribution of the Earth and the internal density
-distribution of the normal Earth.
-However, he did not stress that, by removing the normal gravity 
-evaluated on the ellipsoid from the gravity measured 
-on the Earth's surface, the remaining disturbing field will reflect 
-not only the effect produced by the gravity sources, but also a
-small combination of gravitational and centrifugal effects.
-This additional, non-harmonic and undesired effect is 
-due to the calculation of the normal gravity at the surface of the ellipsoid
-instead of at the same points where the gravity is measured.
-
-- Mencionar tambem que anomalias de gravidade requerem o
-valor da gravidade dentro das fontes
-
-\section{Mathematical description of the gravity disturbance in a local coordinate system}
-
-In a local- or regional-gravity study, 
-geophysicists commonly use a topocentric Cartesian coordinate system
-with the $x$ axis pointing to North, the $y$ axis pointing to East and 
-the $z$ axis with the same direction as the normal $\hat{\mathbf{n}}_{i}$
-to the ellipsoid, but pointing downward (Figure \ref{fig:fig1}).
-In this coordinate system, the observed gravity vector
-$\mathbf{g}^{o}_{i}$, at the point $(x_{i}, y_{i}, z_{i})$, 
-$i = 1, ..., N$, can be represented by
-\begin{equation}
-\mathbf{g}^{o}_{i} = \boldsymbol{\gamma}_{i} + \Delta \mathbf{g}^{o}_{i} \: ,
-\label{eq:gravity-vector}
-\end{equation}
-where $\boldsymbol{\gamma}_{i}$ and $\Delta \mathbf{g}^{o}_{i}$
-are, respectively, the normal gravity vector and a
-disturbing gravitational attraction produced by the anomalous 
-masses at the point $(x_{i}, y_{i}, z_{i})$.
-
-By definition, the gravity disturbance $\delta g^{o}_{i}$,
-at the point $(x_{i}, y_{i}, z_{i})$, is given by \citep{hofmann-wellenhof-moritz2005}:
-\begin{equation}
-\delta g^{o}_{i} =  \| \mathbf{\delta g}^{o}_{i} \| - \| \boldsymbol{\gamma}_{i} \| \: ,
-\label{eq:gravity-disturbance}
-\end{equation}
-where $\| \mathbf{\delta g}^{o}_{i} \|$ and $\| \boldsymbol{\gamma}_{i} \|$ are,
-respectively, the observed gravity and the normal gravity at the
-point $(x_{i}, y_{i}, z_{i})$.
-Fortunately, the condition $\| \boldsymbol{\gamma}_{i} \| \gg \| \Delta \mathbf{g}^{o}_{i} \|$
-is met at all points located above or on the Earth's surface.
-By combining this condition and the definition of observed gravity vector
-given in equation \ref{eq:gravity-vector}, we can approximate the observed gravity
-$\| \mathbf{\delta g}^{o}_{i} \|$ (equation \ref{eq:gravity-disturbance})
-by a first order Taylor's expansion as follows:
-\begin{equation}
-\| \mathbf{\delta g}^{o}_{i} \| \approx \| \boldsymbol{\gamma}_{i} \| + 
-\hat{\boldsymbol{\gamma}}_{i}^{\top} \Delta \mathbf{g}^{o}_{i} \: ,
-\label{eq:gobs-approx}
-\end{equation}
-where $\hat{\boldsymbol{\gamma}}_{i}$ is a unit vector with the same 
-direction as the normal gravity vector $\boldsymbol{\gamma}_{i}$ at
-the point $(x_{i}, y_{i}, z_{i})$.
-This approximation is largely used in applied geophysics for representing
-total-field anomalies (e.g., \citeauthor{blakely1996}, \citeyear{blakely1996}).
-Notice that, local- or regional-gravity studies, the unit
-vector $\hat{\boldsymbol{\gamma}}_{i}$ (equation \ref{eq:gobs-approx}) 
-coincides with the $z$ axis
-of the local Cartesian coordinate system defined
-at the beginning of this section. Consequently,
-by using the approximation defined in equation \ref{eq:gobs-approx}, 
-the gravity disturbance (equation \ref{eq:gravity-disturbance}) can be 
-rewritten as follows \begin{equation}
-\delta g^{o}_{i} \approx \hat{\mathbf{z}}^{\top} \Delta \mathbf{g}^{o}_{i} \: ,
-\label{eq:gravity-disturbance-approx}
-\end{equation}
-where $\hat{\mathbf{z}}^{\top} = \left[ \, 0 \; 0 \; 1 \right]$.
-According to equation \ref{eq:gravity-disturbance-approx},
-the gravity disturbance $\delta g^{o}_{i}$ (equation \ref{eq:gravity-disturbance})
-represents the vertical component of the gravitational attraction exerted by
-the gravity sources at the point $(x_{i}, y_{i}, z_{i})$.
-As a consequence, the gravity disturbance produced by
-a homogeneous gravity source can be represented 
-by the following harmonic function
-\begin{equation}
-d^{o}_{i} = k_{g} \, G \, \rho \, \partial_{z} \phi_{i} \: ,
-\label{eq:gz-local}
-\end{equation}
-where $G$ is the Newtonian constant of gravitation
-(in $m^{3} / (kg \, s^{2})$),
-$k_{g} = 10^{5}$ is a constant factor
-transforming from $m/s^{2}$ to milligal (mGal),
-and $\partial_{z} \phi_{i}$ is a harmonic function representing the first 
-derivative, evaluated at the observation point $(x_{i},y_{i},z_{i})$, 
-$i = 1, ..., N$, of the function 
-\begin{equation}
-\phi(x,y,z) = \int\int\limits_{v}\int \frac{1}{\sqrt{(x - x^{\prime})^{2} + 
-(y - y^{\prime})^{2} + (z - z^{\prime})^{2}}} \: dv
-\label{eq:phi}
-\end{equation}
-with respect to the variable $z$. The 
-integral is conducted over the coordinates $x^{\prime}$, $y^{\prime}$ 
-and $z^{\prime}$ within the volume $v$ of the gravity source.
-This equation can be easily generalized for the case
-of multiple gravity sources.
-
-\section{Geophysical arguments for using the gravity disturbance}
-
-Almost all interpretation techniques assume, implicitly or
-directly, that the gravity data is harmonic (e.g., 
-upward/downward continuation, 
-data processing with equivalent layer,
-conversions between gravity and magnetic data,
-computation of vertical derivatives via Fourier and Hilbert
-transforms).
-
-Consequently, they implicitly or directly assume that
-the gravity data approximates the gravitational disturbance.
-
-Almost all forward modelling techniques compute
-the vertical component of the gravitational attraction 
-exerted by the geological bodies at the observation points.
-
-Hence, almost all geophysicists implicitly compute 
-the vertical component of the gravitational disturbance.
-
-The gravity anomaly is defined as the difference between 
-the gravity $\| \mathbf{g}_{P} \|$, at a point $P$ on the
-Geoid, and the normal gravity $\| \boldsymbol{\gamma}_{Q} \|$,
-on the reference elipsoid, where $P$ and $Q$ have the same
-geodetic latitude and longitude.
-
-Consequently, the gravity anomaly is a function of
-the geodetic latitude and longitude only and cannot
-be calculated at arbitrary heights.
-
-Gravity anomalies require the computation of gravity
-within the topographic masses, where the gravitational
-disturbance is not harmonic.
-
-% The bibliography
-\bibliographystyle{seg}
+\bibliographystyle{gji}
 \bibliography{bib-file}
 
-\clearpage
-
-% Tables
-%\begin{table}
-%    \centering
-%    \renewcommand{\arraystretch}{1.5}
-%    \begin{tabular}{>{\centering\arraybackslash}m{1.1in} 
-%                    >{\centering\arraybackslash}m{1.1in} 
-%                    >{\centering\arraybackslash}m{1.1in}
-%                    >{\centering\arraybackslash}m{1.1in}}
-%        \toprule        
-%        \textbf{Author(s)} & \textbf{Equivalent sources} & 
-%            \textbf{Harmonic function} $g_{ij}$ &
-%            \textbf{Observed data}\\
-%        \midrule
-%        \citet{dampney1969} & point masses & 
-%            vertical component of the gravitational attraction &
-%            Bouguer anomaly \\ \hline
-%        \citet{emilia1973} & lines of dipoles & total-field anomaly &
-%            total-field anomaly \\ \hline
-%        \citet{hansen-miyazaki1984} & uniformly-magnetized plane faces & 
-%            magnetic scalar potential &
-%            total-field anomaly \\ \hline
-%        \citet{silva1986} & doublets & total-field anomaly &
-%        total-field anomaly \\ \hline
-%        \citet{cordell1992} & point masses & gravitational potential &
-%            Bouguer anomaly \\ \hline
-%        \citet{guspi-novara2009} & generic point sources producing a higher-order 
-%            potential & second-directional derivative of the higher-order 
-%            potential & total-field anomaly \\ \hline
-%        \citet{barnes-lumley2011} & rectangular prisms & second derivatives
-%            of the gravitational potential & gravity gradient \\ \hline
-%        \citet{oliveirajr-etal2013} & dipoles & total-field anomaly & 
-%            total-field anomaly \\ \hline
-%        \citet{li-nabighian-oldenburg2014} & rectangular prisms & total-field 
-%            anomaly & total-field anomaly \\
-%        \bottomrule
-%    \end{tabular}
-%    \caption{Observed potential-field data and the
-%    mathematical/physical meaning of the harmonic functions $g_{ij}$
-%    (equation \ref{eq:dpi}) describing the potential field produced 
-%    by the equivalent sources used by several authors.}
-%    \label{tab:equivalent-sources-gij-literature}
-%\end{table}
-%\clearpage
-
-% Figures generated by using Inkscape
+\appendix
+%\section{For authors}
+%
+%Table~\ref{authors} is a list of design macros which are unique to GJI. The
+%list displays each macro's name and description.
+%
+%\section{For editors}
+%
+%The additional features shown in Table~\ref{editors} may be used for
+%production purposes.
+%
+%\bsp % ``This paper has been produced using the Blackwell
+%     %   Publishing GJI \LaTeXe\ class file.''
 
 \begin{figure}
-    \centering
-    \includegraphics[width=0.7\textwidth]{figures/geodetic_coordinates.png}
+%    \vspace{5.5cm}
+    \includegraphics{figures/geodetic_coordinates.png}
     \caption{Schematic representation of the geodetic coordinate
@@
 
-\begin{figure}
-    \centering
-    \includegraphics[width=0.7\textwidth]{figures/gravitational_disturbance.png}
-    \caption{Schematic representation of the geodetic coordinate
-    system.}
-  \label{fig:fig2}
-\end{figure}
+\label{lastpage}
 

### 2018-04-16 manuscript/letter.tex
include new figures and define the unit vectors

@@
 \usepackage{graphicx}
+\usepackage{amsmath}
 
@@
 same total mass (including the atmosphere) and also the
-same angular velocity of the Earth (Figure \ref{fig:fig1}).
+same angular velocity of the Earth \citep{heiskanen-moritz1967,
+vanicek1987,hofmann-wellenhof-moritz2005,torge2012}.
 Another characteristic of this model is that its
@@
 of its own gravity field.
-This model is called as \textit{normal Earth} 
-\citep{heiskanen-moritz1967,vanicek1987,hofmann-wellenhof-moritz2005,
-torge2012}.
-
-It is worth noting that, although the normal Earth has the
-same total mass (including the atmosphere) of the Earth,
-its internal density distribution is unknown.
-
-The only condition imposed on its internal density
-distribution is that it produces a gravity field
-having a particular equipotential which coincides
-with its limiting surface.
-
-For convenience, we denote any density distribution 
-satisfying this condition as a \textit{normal density distribution}.
+Here, we follow \citep{torge2012} and call this model as
+\textit{normal Earth}.
 
@@
 simply \textit{normal gravity}.
+In geodesy, any model used to represent the normal gravity field
+can be arbitrarily defined for the only purpose
+of keeping the difference from the actual gravity field as small 
+as possible \citep{vanicek1987}.
 
-The normal Earth gives rise to the \textit{geodetic coordinate system}.
+It is worth noting that, although the normal Earth has the
+same total mass (including the atmosphere) of the Earth,
+its internal density distribution is unknown.
+The search for physically meaningful mass distributions 
+that generate a required normal gravity field
+has geophysical rather than geodetic motives \citep{marussi1974}.
+The only condition imposed on its internal density
+distribution is that it produces a gravity field
+having a particular equipotential which coincides
+with its limiting surface.
+For convenience, we denote any density distribution 
+satisfying this condition as a \textit{normal density distribution}.
+
+The normal Earth gives rise to the \textit{geodetic coordinate system}
+\citep{heiskanen-moritz1967, soler1976, torge2012, bouman_etal2013}.
 In this coordinate system, the position of a point $P$
@@
 (Figure \ref{fig:fig1}).
-
 The plane containing the point $P$, the axis $Z$ and
@@
 is called meridian plane (gray plane in Figure \ref{fig:fig1}).
+
+At a given point $(h, \varphi, \lambda)$, there are three 
+unit vectors (Figure \ref{fig:fig1}) given by \citep{soler1976}:
+\begin{equation}
+\begin{split}
+\hat{\mitbf{u}} &= 
+\begin{bmatrix}
+\cos\varphi \, \cos\lambda \\
+\cos\varphi \, \sin\lambda \\
+\sin\varphi
+\end{bmatrix} \\
+\hat{\mitbf{v}} &= 
+\begin{bmatrix}
+-\sin\varphi \, \cos\lambda \\
+-\sin\varphi \, \sin\lambda \\
+\cos\varphi
+\end{bmatrix} \\
+\hat{\mitbf{w}} &= 
+\begin{bmatrix}
+-\sin\lambda \\
+\cos\lambda \\
+0
+\end{bmatrix}
+\end{split} \: ,
+\label{eq:unit-vectors}
+\end{equation}
+Notice that these unit vectors are mutually orthogonal.
+The normal gravity vector $\mitbf{\gamma}_{P}$ at a point $P$
+is opposite to the unit vector $\hat{\mitbf{u}}_{P}$ at $P$.
+
+\section{Gravity disturbance}
+
+It is worth noting that, by definition, 
+the centrifugal component of the normal gravity field is
+equal to the centrifugal component of the Earth's gravity
+field if they are evaluated at the same point.
+Then, the differences between the gravity vector
+(corrected from non-gravitational effects) 
+and the normal gravity vector, at the same point, represents a purely 
+gravitational and consequently harmonic disturbing field.
+This gravitational disturbance 
+is caused by contrasts between the actual internal 
+density distribution of the Earth and the internal density 
+distribution of the normal Earth.
+In applied geophysics, these density differences are generally 
+called ``anomalous masses" (e.g., \citeauthor{hammer1945}, 
+\citeyear{hammer1945}; \citeauthor{lafehr1965}, \citeyear{lafehr1965}),
+``density anomalies" (e.g., \citeauthor{forsberg1984}, \citeyear{forsberg1984})
+or ``gravity sources" (e.g., \citeauthor{blakely1996}, 
+\citeyear{blakely1996}). Here, we opted for using the last term.
+
+The difference between the observed gravity and the
+normal gravity, at the same point, is called gravity disturbance
+\citep{hofmann-wellenhof-moritz2005}.
+Notice that the gravity disturbance is not equivalent to the
+magnitude of the difference between the gravity vector
+and the normal gravity vector at the same point.
+As properly pointed out by \citet{hackney-featherstone2003},
+the gravity disturbance is a very-well established quantity in geodesy,
+but appears to be less well known in geophysics.
+
+The gravity anomaly is the commonly used quantity in applied 
+geophysics. It is defined as the difference
+between the gravity on the geoid and the normal gravity on the ellipsoid,
+both at the same geodetic latitude and longitude.
+Notice that, by definition, the gravity anomaly depends on 
+longitude and latitude only and is not a function in space 
+\citep{barthelmes2013}.
+Different gravity anomalies can be calculated, depending on the
+corrections applied to them \citep{blakely1996, hofmann-wellenhof-moritz2005}.
+These corrections are usually called gravity reductions.
+For example, the Free-air anomaly is an approximation of the
+gravity disturbance whereas the Bouguer anomaly
+is an approximation of the terrain corrected gravity disturbance.
+The last one is commonly used by geophysicists as the
+gravitational effect produced by the gravity sources.
+Although this approximation is valid for most practical applications,
+it is important to bear in mind not only the terminology 
+changes, but also the conceptual assumptions.
+
+There was a certain lack of comprehension regarding the
+geophysical meaning of gravity anomalies until the
+mid 90's.
+As properly pointed out by \citet{chapin1996} at that time, 
+``although the corrections which bring about a Bouguer 
+gravity anomaly are well established, the reasons for doing
+them are not well understood. One cause of this common 
+misunderstanding is that the subject has been poorly presented in
+many of the basic texts".
+In his seminal book, \citet{blakely1996} brought some light
+on the geophysical meaning of gravity anomalies from the 
+perspective of applied geophysics. \citet{blakely1996} correctly
+defined gravity sources as density contrasts between the actual
+internal density distribution of the Earth and the internal density
+distribution of the normal Earth.
+However, he did not stress that, by removing the normal gravity 
+evaluated on the ellipsoid from the gravity measured 
+on the Earth's surface, the remaining disturbing field will reflect 
+not only the effect produced by the gravity sources, but also a
+small combination of gravitational and centrifugal effects.
+This additional, non-harmonic and undesired effect is 
+simply due to the calculation of the normal gravity at a point
+other than that were the gravity is measured.
+
+\section{Mathematical description of the gravity disturbance in a local coordinate system}
+
+In a local- or regional-gravity study, 
+geophysicists commonly use a topocentric Cartesian coordinate system
+with origin at a point $P$ and axes $x$, $y$ and $z$ defined
+by the unit vectors $\hat{\mitbf{v}}_{P}$,
+$\hat{\mitbf{w}}_{P}$ and $-\hat{\mitbf{u}}_{P}$ 
+(equation \ref{eq:unit-vectors}), respectively
+(Figures \ref{fig:fig1} and \ref{fig:fig2}).
+In this coordinate system, the observed gravity vector
+$\mitbf{g}_{i}$, at a point $(x_{i}, y_{i}, z_{i})$, 
+$i = 1, ..., N$, can be represented by
+\begin{equation}
+\mitbf{g}_{i} = \mitbf{\gamma}_{i} + \Delta \mitbf{g}_{i} \: ,
+\label{eq:gravity-vector}
+\end{equation}
+where $\mitbf{\gamma}_{i}$ and $\Delta \mitbf{g}_{i}$
+are, respectively, the normal gravity vector and a
+disturbing gravitational attraction produced by the anomalous 
+masses at the point $(x_{i}, y_{i}, z_{i})$.
+
+By definition, the gravity disturbance $\delta g_{i}$,
+at the point $(x_{i}, y_{i}, z_{i})$, is given by 
+\citep{hofmann-wellenhof-moritz2005}:
+\begin{equation}
+\delta g_{i} =  g_{i} - \gamma_{i} \: ,
+\label{eq:gravity-disturbance}
+\end{equation}
+where $g_{i} = \| \mitbf{g}_{i} \|$ and 
+$\gamma_{i} = \| \mitbf{\gamma}_{i} \|$ are,
+respectively, the observed gravity and the normal gravity at the
+point $(x_{i}, y_{i}, z_{i})$.
+Fortunately, the condition 
+$\gamma_{i} \gg \| \Delta \mathbf{g}_{i} \|$ 
+is met at all points located above or on the Earth's surface.
+By combining this condition and the definition of observed gravity vector
+(equation \ref{eq:gravity-vector}), we can approximate the observed gravity
+$g_{i}$ by a first order Taylor's expansion as follows 
+\citep{sanso_sideris2013}:
+\begin{equation}
+g_{i} \approx \gamma_{i} + 
+\hat{\mitbf{\gamma}}_{i}^{\top} \Delta \mitbf{g}_{i} \: ,
+\label{eq:gobs-approx}
+\end{equation}
+where $\hat{\mitbf{\gamma}}_{i} = -\hat{\mitbf{u}}_{i}$ is a unit 
+vector with the same direction as the normal gravity vector 
+$\mitbf{\gamma}_{i}$ at the point $(x_{i}, y_{i}, z_{i})$ and
+$\hat{\mitbf{u}}_{i}$ is the unit vector $\hat{\mitbf{u}}$ 
+(equation \ref{eq:unit-vectors}) evaluated at the point $(h_{i}, \varphi_{i}, \lambda_{i})$, which is obtained by transforming $(x_{i}, y_{i}, z_{i})$
+from the topocentric coordinate system (Figure \ref{fig:fig2}) to the
+geodetic coordinate system (Figure \ref{fig:fig1}).
+
+PAREI AQUI
+
+This approximation, which is known in geodesy \citep[e.g.,][]{sanso_sideris2013},
+is largely used in applied geophysics for representing
+total-field anomalies \citep[e.g.,][]{blakely1996}.
+Notice that, local- or regional-gravity studies, the unit
+vector $\hat{\mitbf{\gamma}}_{i}$ (equation \ref{eq:gobs-approx}) 
+coincides with the $z$ axis
+of the local Cartesian coordinate system defined
+at the beginning of this section. Consequently,
+by using the approximation defined in equation \ref{eq:gobs-approx}, 
+the gravity disturbance (equation \ref{eq:gravity-disturbance}) can be 
+rewritten as follows \begin{equation}
+\delta g^{o}_{i} \approx \hat{\mitbf{z}}^{\top} \Delta \mitbf{g}^{o}_{i} \: ,
+\label{eq:gravity-disturbance-approx}
+\end{equation}
+where $\hat{\mitbf{z}}^{\top} = \left[ \, 0 \; 0 \; 1 \right]$.
+According to equation \ref{eq:gravity-disturbance-approx},
+the gravity disturbance $\delta g^{o}_{i}$ (equation \ref{eq:gravity-disturbance})
+represents the vertical component of the gravitational attraction exerted by
+the gravity sources at the point $(x_{i}, y_{i}, z_{i})$.
+As a consequence, the gravity disturbance produced by
+a homogeneous gravity source can be represented 
+by the following harmonic function
+\begin{equation}
+d^{o}_{i} = k_{g} \, G \, \rho \, \partial_{z} \phi_{i} \: ,
+\label{eq:gz-local}
+\end{equation}
+where $G$ is the Newtonian constant of gravitation
+(in $m^{3} / (kg \, s^{2})$),
+$k_{g} = 10^{5}$ is a constant factor
+transforming from $m/s^{2}$ to milligal (mGal),
+and $\partial_{z} \phi_{i}$ is a harmonic function representing the first 
+derivative, evaluated at the observation point $(x_{i},y_{i},z_{i})$, 
+$i = 1, ..., N$, of the function 
+\begin{equation}
+\phi(x,y,z) = \int\int\limits_{v}\int \frac{1}{\sqrt{(x - x^{\prime})^{2} + 
+(y - y^{\prime})^{2} + (z - z^{\prime})^{2}}} \: dv
+\label{eq:phi}
+\end{equation}
+with respect to the variable $z$. The 
+integral is conducted over the coordinates $x^{\prime}$, $y^{\prime}$ 
+and $z^{\prime}$ within the volume $v$ of the gravity source.
+This equation can be easily generalized for the case
+of multiple gravity sources.
+
+\section{Gravity disturbance versus gravity anomaly}
+
+Almost all interpretation techniques assume, implicitly or
+directly, that the gravity data is harmonic (e.g., 
+upward/downward continuation, 
+data processing with equivalent layer,
+conversions between gravity and magnetic data,
+computation of vertical derivatives via Fourier and Hilbert
+transforms).
+
+Consequently, they implicitly or directly assume that
+the gravity data approximates the gravitational disturbance.
+
+Almost all forward modelling techniques compute
+the vertical component of the gravitational attraction 
+exerted by the geological bodies at the observation points.
+
+Hence, almost all geophysicists implicitly compute 
+the vertical component of the gravitational disturbance.
+
+The gravity anomaly is defined as the difference between 
+the gravity $\| \mitbf{g}_{P} \|$, at a point $P$ on the
+Geoid, and the normal gravity $\| \mitbf{\gamma}_{Q} \|$,
+on the reference elipsoid, where $P$ and $Q$ have the same
+geodetic latitude and longitude.
+
+Consequently, the gravity anomaly is a function of
+the geodetic latitude and longitude only and cannot
+be calculated at arbitrary heights.
+
+Gravity anomalies require the computation of gravity
+within the topographic masses, where the gravitational
+disturbance is not harmonic.
 
@@
 %    \vspace{5.5cm}
-    \includegraphics{figures/geodetic_coordinates.png}
+    \includegraphics{figures/cartesian-geodetic-systems.png}
     \caption{Schematic representation of the geodetic coordinate
@@
     determined by the geometric height $h$, the geodetic latitude 
-    $\varphi$ and longitude $\lambda$. The Earth's origin is represented 
+    $\varphi$ and longitude $\lambda$. 
+    The Earth's center of mass is represented 
     by $O$, $P$ represents a point $(h, \varphi, \lambda)$ and 
@@
     $P^{\prime}$ is represented by the gray triangle in (a) and (b). 
+    The unit vectors 
+    $\hat{\mitbf{u}}_{P}$, $\hat{\mitbf{v}}_{P}$ and 
+    $\hat{\mitbf{w}}_{P}$ define three mutually orthogonal 
+    directions at each point $P$.
     In (b), the point $Q$ represents the projection of $P$ on the 
-    ellipsoid surface.}
+    ellipsoid surface, at the same latitude $\varphi$ and 
+    longitude $\lambda$.}
   \label{fig:fig1}
+\end{figure}
+
+\begin{figure}
+    \includegraphics{figures/local-system.png}
+    \caption{(a) Schematic representation of the gravity vector
+    $\mitbf{g}_{P}$, normal gravity vector $\mitbf{\gamma}_{P}$,
+    disturbing gravitational attraction $\Delta \mitbf{g}_{P}$
+    (equation \ref{eq:gravity-vector}) and the unit vector 
+    $\hat{\mitbf{u}}_{P}$ (equation \ref{eq:unit-vectors}) 
+    at the point $P$ and also the normal gravity vector
+    $\mitbf{\gamma}_{Q}$ at the point $Q$ on the reference ellipsoid.
+    At the point $P$ the normal gravity vector $\mitbf{\gamma}_{P}$ 
+    is opposite to the unit vector $\hat{\mitbf{u}}_{P}$ and the 
+    normal gravity $\mitbf{\gamma}_{Q}$ at the point $Q$ is opposite
+    the another unit vector $\hat{\mitbf{u}}_{Q}$, which is
+    close to $\hat{\mitbf{u}}_{P}$.
+    (b) Schematic representation of a topocentric Cartesian 
+    coordinate system with origin at a point $P$ and $x$, $y$ and $z$ 
+    axes defined by the unit vectors $\hat{\mitbf{v}}_{P}$, 
+    $\hat{\mitbf{w}}_{P}$ and $-\hat{\mitbf{u}}_{P}$ 
+    (Figure \ref{fig:fig1}), respectively.}
+  \label{fig:fig2}
 \end{figure}

### 2018-04-16 manuscript/old_ideas.tex
include new figures and define the unit vectors

 % gjilguid2e.tex
% V2.0 released 1998 December 18
% V2.1 released 2003 October 7 -- Gregor Hutton, updated the web address for the style files.

%\documentclass{gji}
%\documentclass[extra,mreferee]{gji}
\documentclass[mreferee]{gji}
\usepackage{gji_extra}
\usepackage{timet}
%\usepackage{units, amstext, amsmath, amssymb, stfloats, graphics, setspace, dcolumn}
\usepackage{units, amstext, amsmath, amssymb} % There is a problem with some of the above packages
\usepackage{graphicx}
\usepackage{caption}
%\usepackage{subcaption} % There is a problem with this package
\usepackage{subfig}

\title[Differential operators in geodetic coordinates] % short title
  {Differential operators in geodetic coordinates applied to potential-field modelling} % main title
\author[K. A. T. Hallam, V. C. Oliveira Jr.] % first page
  {K. A. T. Hallam$^1$, V. C. Oliveira Jr.$^1$ %
  \\
  $^1$ Department of Geophysics, Observat\'{o}rio Nacional, Rio de Janeiro
  , Brazil
  }
%\date{Received 1998 December 18; in original form 1998 November 22}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
%\volume{200}
\pubyear{2017}

\def\LaTeX{L\kern-.36em\raise.3ex\hbox{{\small A}}\kern-.15em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}
\def\LATeX{L\kern-.36em\raise.3ex\hbox{{\Large A}}\kern-.15em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}
% Authors with AMS fonts and mssymb.tex can comment out the following
% line to get the correct symbol for Geophysical Journal International.
\let\leqslant=\leq

%\newtheorem{theorem}{Theorem}[section]

\begin{document}

\label{firstpage}

\maketitle

\begin{summary}

TEXT TEXT TEXT

\end{summary}

\begin{keywords}
%\LaTeXe\ -- class files: \verb"gji.cls"\ -- sample text -- user guide.
Numerical modelling, Gravity anomalies and Earth structure, Magnetic anomalies: modelling and interpretation
\end{keywords}

\section{Introduction} \label{sec.intro}

Several methods have been developed to
compute the gravitational and magnetic fields produced by
geological structures in Cartesian (CITAR PAPERS CLASSICOS)
and spherical coordinates (CITAR PAPERS CLASSICOS).

The spherical approach is more appropriated than the Cartesian approach
for modelling large-scale structures which require taking the Earth's
curvature into account.

The spherical approach, however, neglects the Earth's ellipticity.

\citet{roussel2015} is the only work presenting a potential-field modelling
that takes the Earth's ellipticity into consideration.
Their method is formulated in ellipsoidal coordinates and
relies on the numerical solution of volume integrals by applying
Gauss-Legendre quadrature.

The gravitational and magnetic fields computed by these methods cannot
be directly compared to geophysical observations because the position of
the last ones are commonly described in geodetic coordinates  
\citep[e.g.,][]{hotine1969, heiskanen-moritz1967, krakiwsky1971, soler1976,
vanicek1987, rapp1991, seeber2003, hofmann-wellenhof-moritz2005, torge2012}.

Consequently, coordinate transformations are required.

Up to now, the geoscientific community lacks of methods for computing
the gravitational and magnetic fields produced by geological bodies
in geodetic coordinates.

We present the differential operators gradient, divergent,
Laplacian and curl in terms of the scale factors and the
normal basis associated to the GCS.

We applied the differential operators presented here to
define the expressions of the gravitational and magnetic fields
produced by spheres.

Such expressions are can be used, for example, for producing regional
magnetic anomaly maps \citep[e.g.,][]{vonfrese-etal1981, mayhew1982, dyment1998},
characterizing the regional gravity field \citep[e.g.,][]{needham1970, balmino1972,
barthelmes1991, barthelmes1991b, lehmann1993, antunes2003, guspi2004, lin2014} or
approximating the gravitational and magnetic fields produced by geological
structures with arbitrary shapes.

%Finally, we also provide a set of routines developed in Python language...

\section{Geodetic curvilinear coordinates}

\subsection{Relationship between Cartesian and Geodetic coordinates}

Consider a terrestrial geocentric system of Cartesian coordinates
having the $z$-axis coincident with the mean rotational axis,
the $x$-axis pointing to the Greenwich meridian and the $y$-axis
directed so as to obtain a right-handed system
(Fig. \ref{fig:cartesian-geodetic-systems}).
Coordinate systems similar to this one may be found in the literature as
Conventional Terrestrial Reference Coordinate System \citep[e.g.,][]{soler1988}, 
International Terrestrial Reference System \citep[e.g.,][]{seeber2003, torge2012} 
or Earth-centered Earth-fixed system \citep[e.g.,][]{bouman2013}, for example.
Here, we opted for simply using the term Geocentric Cartesian System
(GCS).
Let us also consider a terrestrial geocentric system of geodetic coordinates
$h$ (geometric height), $\varphi$ (geodetic latitude), and 
$\lambda$ (longitude) defined with respect to a reference ellipsoid
of revolution having major semi-axis $a$ and minor semi-axis $b$.
For convenience, we call this system Geocentric Geodetic System
(GGS).

The geodetic coordinates $h$, $\varphi$, and $\lambda$ can be transformed
into the Cartesian coordinates $x$, $y$, and $z$ as follows
\citep{heiskanen-moritz1967}:
\begin{equation} \label{eq:geodetic-2-cartesian}
\begin{split}
x &= \left( N + h \right) \, \cos\varphi \, \cos\lambda \\ 
y &= \left( N + h \right) \, \cos\varphi \, \sin\lambda \\
z &= \left[ N \left( 1 - e^{2} \right) + h \right] \, \sin\varphi
\end{split} \: ,
\end{equation}
where $e = \sqrt{a^{2} - b^{2}}/a$ is the first eccentricity and
$N$ is the prime vertical radius of curvature given by
\begin{equation} \label{eq:N-prime-radius}
N = \frac{a}{\left( 1 - e^{2} \sin^{2}\varphi \right)^{\frac{1}{2}}} \: .
\end{equation}
Additionally, let us define the meridian radius of curvature $M$ as
follows:
\begin{equation} \label{eq:M-meridian-radius}
M = \frac{a \left( 1 - e^{2} \right)}{\left( 1 - e^{2} \sin^{2}\varphi \right)^{\frac{3}{2}}} \: .
\end{equation}
It can be easily verified that the prime vertical radius of curvature $N$
(eq. \ref{eq:N-prime-radius}), the meridian radius of curvature $M$
(eq. \ref{eq:M-meridian-radius}) and their first derivatives with respect to
$\varphi$ satisfy the following relationships:
\begin{equation} \label{eq:relation-betwen-N-M}
M = \frac{1 - e^{2}}{1 - e^{2} \sin^{2}\varphi} N
\end{equation}
and
\begin{equation} \label{eq:N-derivative}
\frac{\partial N}{\partial \varphi} = \frac{e^{2} \, \sin \varphi \, \cos \varphi}{1 - e^{2} \sin^{2}\varphi} N \: .
\end{equation}

\subsection{Scale factors and local basis of the GGS}

Having defined the relations between the Cartesian and geodetic coordinates,
we will now apply some well-established concepts of generic curvilinear
coordinates. An in-depth presentation of these concepts may be found at
\citet{kellogg1929}, \citet{hotine1969}, \citet{arfken2001}, or \citet{sttraton2007},
for example. We shall confine ourselves to
the particular case of geodetic curvilinear coordinates.

Let us first make use of the Pythagorean theorem to define the
squared magnitude of the infinitesimal displacement between two
neighbouring points in the GCS system is given by
\begin{equation} \label{eq:ds2-cartesian}
ds^{2} = dx^{2} + dy^{2} + dz^{2} \: ,
\end{equation}
where the differentials $dx$, $dy$, and $dz$ are obtained from
eqs \ref{eq:geodetic-2-cartesian} as follows:
\begin{equation} \label{eq:dalpha}
d\alpha = \frac{\partial \alpha}{\partial h} \, dh +
\frac{\partial \alpha}{\partial \varphi} \, d\varphi +
\frac{\partial \alpha}{\partial \lambda} \, d\lambda
\: , \quad \alpha = x, y, z \: ,
\end{equation}
and the partial derivatives $\frac{\partial \alpha}{\partial \beta}$,
$\alpha = x, y, z$, $\beta = h, \varphi, \lambda$, are given by \citep{soler1976}:
\begin{equation} \label{eq:dalpha-dbeta}
\begin{array}{ccc}
\frac{\partial x}{\partial h} = \cos\varphi \, \cos\lambda &
\frac{\partial x}{\partial \varphi} = -\left( M + h \right) \, \sin\varphi \, \cos\lambda &
\frac{\partial x}{\partial \lambda} = -\left( N + h \right) \, \cos\varphi \, \sin\lambda \\
\frac{\partial y}{\partial h} = \cos\varphi \, \sin\lambda &
\frac{\partial y}{\partial \varphi} = -\left( M + h \right) \, \sin\varphi \, \sin\lambda &
\frac{\partial y}{\partial \lambda} = \left( N + h \right) \, \cos\varphi \, \cos\lambda \\
\frac{\partial z}{\partial h} = \sin\varphi &
\frac{\partial z}{\partial \varphi} = \left( M + h \right) \, \cos\varphi &
\frac{\partial z}{\partial \lambda} = 0 \\
\end{array} \: .
\end{equation}
The partial derivatives with respect to $\varphi$ were deduced from
eqs \ref{eq:geodetic-2-cartesian} by using
eqs \ref{eq:relation-betwen-N-M} and \ref{eq:N-derivative}.

From these differentials $d\alpha$ (eqs \ref{eq:dalpha}), we define the
Jacobian matrix
\begin{equation} \label{eq:Jacobian-matrix}
\mathbf{J} = \begin{bmatrix}
\frac{\partial x}{\partial h} & \frac{\partial x}{\partial \varphi} & \frac{\partial x}{\partial \lambda} \\
\frac{\partial y}{\partial h} & \frac{\partial y}{\partial \varphi} & \frac{\partial y}{\partial \lambda} \\
\frac{\partial z}{\partial h} & \frac{\partial z}{\partial \varphi} & \frac{\partial z}{\partial \lambda}
\end{bmatrix} \: .
\end{equation}

By substituting the differentials $d\alpha$ (eqs \ref{eq:dalpha}) into
eq. \ref{eq:ds2-cartesian}, we obtain a general quadratic form
representing the squared distance $ds^{2}$ given by
\begin{equation} \label{eq:ds2}
ds^{2} = g_{11} \, dh^{2} + g_{22} \, d\varphi^{2} + g_{33} \, d\lambda^{2} + 
2 \left( g_{12} \, dh \, d\varphi + g_{13} \, dh \, d\lambda +
g_{23} \, d\varphi \, d\lambda \right) \: ,
\end{equation}
where
\begin{equation} \label{eq:metric-coeffs-gii}
\begin{split}
g_{11} &= \left( \frac{\partial x}{\partial h} \right)^{2}
+ \left( \frac{\partial y}{\partial h} \right)^{2} +
\left( \frac{\partial z}{\partial h} \right)^{2} \\
g_{22} &= \left( \frac{\partial x}{\partial \varphi} \right)^{2}
+ \left( \frac{\partial y}{\partial \varphi} \right)^{2} +
\left( \frac{\partial z}{\partial \varphi} \right)^{2} \\
g_{33} &= \left( \frac{\partial x}{\partial \lambda} \right)^{2}
+ \left( \frac{\partial y}{\partial \lambda} \right)^{2} +
\left( \frac{\partial z}{\partial \lambda} \right)^{2}
\end{split} \quad ,
\end{equation}
and
\begin{equation} \label{eq:metric-coeffs-gij}
\begin{split}
g_{12} &=
\left( \frac{\partial x}{\partial h} \, \frac{\partial x}{\partial \varphi} \right) +
\left( \frac{\partial y}{\partial h} \, \frac{\partial y}{\partial \varphi} \right) +
\left( \frac{\partial z}{\partial h} \, \frac{\partial z}{\partial \varphi} \right) \\
g_{13} &=
\left( \frac{\partial x}{\partial h} \, \frac{\partial x}{\partial \lambda} \right) +
\left( \frac{\partial y}{\partial h} \, \frac{\partial y}{\partial \lambda} \right) +
\left( \frac{\partial z}{\partial h} \, \frac{\partial z}{\partial \lambda} \right) \\
g_{23} &=
\left( \frac{\partial x}{\partial \varphi} \, \frac{\partial x}{\partial \lambda} \right) +
\left( \frac{\partial y}{\partial \varphi} \, \frac{\partial y}{\partial \lambda} \right) +
\left( \frac{\partial z}{\partial \varphi} \, \frac{\partial z}{\partial \lambda} \right)
\end{split} \quad .
\end{equation}
The coefficients defined by eqs \ref{eq:metric-coeffs-gii} and
\ref{eq:metric-coeffs-gij} are commonly called metrical coefficients.
It can be shown that the coefficients $g_{12}$, $g_{13}$, and $g_{23}$
(eqs \ref{eq:metric-coeffs-gij}) are equal to zero, which implies that
the GGS is an orthogonal system.

The squared infinitesimal distance $ds^{2}$ (eq. \ref{eq:ds2})
can be rewritten as follows:
\begin{equation} \label{eq:ds2-orthogonal}
ds^{2} = \left( h_{1} \, dh \right)^{2} + 
\left( h_{2} \, d\varphi \right)^{2} + \left( h_{3} \, d\lambda \right)^{2} \: ,
\end{equation}
where $h_{i}$, $i = 1, 2, 3$, are called scale factors.
These coefficients are defined as the square root of the metrical coefficients
$g_{ii}$ (eqs \ref{eq:metric-coeffs-gii}) as follows:
\begin{equation} \label{eq:scale-factors-hi}
\begin{split}
h_{1} &= 1 \\
h_{2} &= M + h \\
h_{3} &= \left( N + h \right) \cos\varphi
\end{split} \quad ,
\end{equation}
where $N$ and $M$ are defined by eqs \ref{eq:N-prime-radius} and
\ref{eq:M-meridian-radius}, respectively.
Notice that, in eq. \ref{eq:ds2-orthogonal}, the terms
$h_{1} \, dh$, $h_{2} \, d\varphi$, and $h_{3} \, d\lambda$
represent infinitesimal displacement components along the geodetic
coordinates $h$, $\varphi$, and $\lambda$, respectively.
By using the scale factors $h_{1}$, $h_{2}$, and $h_{3}$ (eqs \ref{eq:scale-factors-hi}),
we may define the metric matrix $\mathbf{H}$ \citep{soler1976} as follows:
\begin{equation} \label{eq:H-metric-matrix}
\mathbf{H} = \begin{bmatrix}
h_{1} & & \\
& h_{2} & \\
& & h_{3}
\end{bmatrix} \: .
\end{equation}

It is also necessary to define a set of mutually orthogonal vectors at
a given fixed point.
To do this, let us first define the \textit{position vector} 
$\mathbf{r}$ with elements defined by the Cartesian coordinates
$x$, $y$ and $z$ (eq. \ref{eq:geodetic-2-cartesian}) as follows:
\begin{equation} \label{eq:position-vector}
\mathbf{r} = \begin{bmatrix}
x(h, \varphi, \lambda) \\
y(h, \varphi, \lambda) \\
z(h, \varphi, \lambda)
\end{bmatrix} \: .
\end{equation}
Then, the set of mutually orthogonal vectors can be
given by
\begin{equation} \label{eq:unitary-vectors}
\begin{split}
\mathbf{e}_{1} &= \frac{\partial \mathbf{r}}{\partial h} \\
\mathbf{e}_{2} &= \frac{\partial \mathbf{r}}{\partial \varphi} \\
\mathbf{e}_{3} &= \frac{\partial \mathbf{r}}{\partial \lambda}
\end{split} \: .
\end{equation}
Such vectors are called \textit{unitary vectors} \citep{sttraton2007}.
Notice that the unitary vectors $\mathbf{e}_{1}$, $\mathbf{e}_{2}$, and $\mathbf{e}_{3}$
are defined, respectively, in the direction of the increasing geodetic coordinates
$h$, $\varphi$, and $\lambda$.
Besides, the magnitude of a unitary vector $\mathbf{e}_{i}$ is equal to
the scale factor $h_{i}$, $i = 1, 2, 3$,
which shows that it is not necessarily a unit vector.
By multiplying the unitary vectors $\mathbf{e}_{i}$ (eq. \ref{eq:unitary-vectors})
by the reciprocal of the scale factors $h_{i}$ (eq. \ref{eq:scale-factors-hi})
and using the partial derivatives defines by eqs \ref{eq:dalpha-dbeta},
we obtain a set of three mutually orthogonal unit vectors given by \citep{soler1976}
\begin{equation} \label{eq:unit-vectors}
\begin{split}
\hat{\mathbf{e}}_{1} &= 
\begin{bmatrix}
\cos\varphi \, \cos\lambda \\
\cos\varphi \, \sin\lambda \\
\sin\varphi
\end{bmatrix} \\
\hat{\mathbf{e}}_{2} &= 
\begin{bmatrix}
-\sin\varphi \, \cos\lambda \\
-\sin\varphi \, \sin\lambda \\
\cos\varphi
\end{bmatrix} \\
\hat{\mathbf{e}}_{3} &= 
\begin{bmatrix}
-\sin\lambda \\
\cos\lambda \\
0
\end{bmatrix}
\end{split} \: ,
\end{equation}
where $N$ and $M$ are defined by eqs \ref{eq:N-prime-radius} and
\ref{eq:M-meridian-radius}, respectively.
These vectors form a local orthonormal basis in the space of Cartesian
coordinates.
By using these vectors (eqs \ref{eq:unit-vectors}), we can define the
matrix equation transforming a vector $\mathbf{v}_{C}$, defined at a
point in the GCS, into a vector $\mathbf{v}_{G}$, defined 
at the same point in the GGS, as follows:
\begin{equation} \label{eq:vC-2-vG}
\mathbf{v}_{G} = \mathbf{R}^{\top} \mathbf{v}_{C} \: ,
\end{equation}
where $\mathbf{R}$ is a rotation matrix \citep{soler1976} given by
\begin{equation} \label{eq:Rotation-matrix}
\mathbf{R} = \begin{bmatrix}
\hat{\mathbf{e}}_{1} & \hat{\mathbf{e}}_{2} & \hat{\mathbf{e}}_{3}
\end{bmatrix} \: .
\end{equation}
It can be easily verified that the Jacobian $\mathbf{J}$ (eq. \ref{eq:Jacobian-matrix}),
metric $\mathbf{H}$ (eq. \ref{eq:H-metric-matrix}), and
rotation $\mathbf{R}$ (eq. \ref{eq:Rotation-matrix}) matrices
satisfy the following relationship \citep{soler1976}:
\begin{equation} \label{eq:J-R-H-relationship}
\mathbf{R} \mathbf{J} = \mathbf{H} \: .
\end{equation}

\subsection{Differential operators}

The \textit{gradient}, \textit{divergent}, \textit{Laplacian}, and
\textit{curl} differential operators in terms of the scale factors
$h_{i}$ (eq. \ref{eq:scale-factors-hi})
and the unit vectors $\hat{\mathbf{e}_{i}}$ (eq \ref{eq:unit-vectors})
as follows \citep[e.g.,][]{kellogg1929, hotine1969, arfken2001, sttraton2007}:

\begin{equation} \label{eq:gradient-geodetic}
\nabla \, \psi
= \frac{\partial \psi}{\partial h} \, \hat{\mathbf{e}}_{1} +
\frac{1}{\left( M + h \right)} \, \frac{\partial \psi}{\partial \varphi} \, \hat{\mathbf{e}}_{2} +
\frac{1}{\left( N + h \right) \cos\varphi} \, \frac{\partial \psi}{\partial \lambda} \, \hat{\mathbf{e}}_{3}
\quad ,
\end{equation}

%\begin{equation} \label{eq:gradient-geodetic}
%\nabla \, \psi
%= \mathbf{R} \mathbf{H}^{-1}
%\begin{bmatrix}
%\frac{\partial \psi}{\partial h} \\
%\frac{\partial \psi}{\partial \varphi} \\
%\frac{\partial \psi}{\partial \lambda}
%\end{bmatrix}
%\quad ,
%\end{equation}

%\begin{equation} \label{eq:divergent-geodetic}
%\begin{split}
%\nabla \cdot \mathbf{V} = \gamma \Bigg \{
%& \frac{\partial}{\partial h} \left[ \left( M + h \right) \left( N + h \right) \cos\varphi \, V_{1} \right] + \\
%& \frac{\partial}{\partial \varphi} \left[ \left( N + h \right) \cos\varphi \, V_{2} \right] + \\
%& \frac{\partial}{\partial \lambda} \left[ \left( M + h \right) \, V_{3} \right]
%\Bigg \} \end{split} \quad ,
%\end{equation}

\begin{equation} \label{eq:divergent-geodetic}
\begin{split}
\nabla \cdot \mathbf{V} =
& \left[ \frac{1}{\left( N + h \right)} + \frac{1}{\left( M + h \right)} \right] \, V_{1} + \frac{\partial V_{1}}{\partial h} \\
+ & \frac{\tan \varphi}{\left( N + h \right)} \, V_{2} + \frac{1}{\left( M + h \right)} \, \frac{\partial V_{2}}{\partial \varphi} +
\frac{1}{\left( N + h \right) \cos\varphi} \, \frac{\partial V_{3}}{\partial \lambda}
\end{split} \quad ,
\end{equation}

%\begin{equation} \label{eq:Laplacian-generic}
%\begin{split}
%\nabla^{2} \psi = \gamma
%\Bigg \{
%& \frac{\partial}{\partial h} \left[ \left( M + h \right) \left( N + h \right) \cos\varphi \frac{\partial \psi}{\partial h} \right] + \\
%& \frac{\partial}{\partial \varphi} \left[ \frac{\left( N + h \right) \cos\varphi}{\left( M + h \right)} \frac{\partial \psi}{\partial \varphi} \right] + \\
%& \frac{\partial}{\partial \lambda} \left[ \frac{\left( M + h \right)}{\left( N + h \right) \cos\varphi} \frac{\partial \psi}{\partial \lambda} \right]
%\Bigg \} \end{split}
%\end{equation}
\begin{equation} \label{eq:Laplacian-geodetic}
\begin{split}
\nabla^{2} \psi = 
& \left[ \frac{1}{\left( N + h \right)} + \frac{1}{\left( M + h \right)} \right] \, \frac{\partial \psi}{\partial h} + 
\frac{\partial^{2} \psi}{\partial h^{2}} \\
- & \left[ \frac{\tan \varphi}{\left( N + h \right) \left( M + h \right)} + \frac{1}{\left( M + h \right)^{3}} \frac{\partial M}{\partial \varphi} \right] \,
\frac{\partial \psi}{\partial \varphi} + \frac{1}{\left( M + h \right)^{2}} \frac{\partial^{2} \psi}{\partial \varphi^{2}} \\
+ & \frac{1}{\left( N + h \right)^{2} \cos^{2} \varphi} \frac{\partial^{2} \psi}{\partial \lambda^{2}}
\end{split}
\end{equation}
and
%\begin{equation} \label{eq:curl-generic}
%\nabla \times \mathbf{V} = \gamma
%\left| \begin{array}{ccc}
%\hat{\mathbf{e}}_{1} & \left( M + h \right) \hat{\mathbf{e}}_{2} & \left( N + h \right) \cos\varphi \, \hat{\mathbf{e}}_{3} \\
%\frac{\partial}{\partial h} & \frac{\partial}{\partial \varphi} & \frac{\partial}{\partial \lambda} \\
%V_{1} & \left( M + h \right) V_{2} & \left( N + h \right) \cos\varphi V_{3}
%\end{array} \right| \: ,
%\end{equation}
\begin{equation} \label{eq:curl-geodetic}
\begin{split}
\nabla \times \mathbf{V} = 
& \left[ - \frac{\tan \varphi}{\left( N + h \right)} V_{3} + \frac{1}{\left( M + h \right)} \frac{\partial V_{3}}{\partial \varphi}
- \frac{1}{\left( N + h \right) \cos \varphi} \frac{\partial V_{2}}{\partial \lambda} \right] \hat{\mathbf{e}}_{1} \\
+ & \left[ - \frac{1}{\left( N + h \right)} V_{3} + \frac{1}{\left( N + h \right) \cos \varphi} \frac{\partial V_{1}}{\partial \lambda}
- \frac{\partial V_{3}}{\partial h} \right] \hat{\mathbf{e}}_{2} \\
+ & \left[ \frac{1}{\left( M + h \right)} V_{2} + \frac{\partial V_{2}}{\partial h}
- \frac{1}{\left( M + h \right) } \frac{\partial V_{1}}{\partial \varphi} \right] \hat{\mathbf{e}}_{3}
\end{split} \quad ,
\end{equation}
where $\psi$ is a scalar field which is invariant to
a rotation of the coordinate system and $\mathbf{V}$
is a vector field with components $V_{1}$, $V_{2}$, and $V_{3}$ along the
unit vectors $\hat{\mathbf{e}}_{1}$, $\hat{\mathbf{e}}_{2}$, and
$\hat{\mathbf{e}}_{3}$ (eqs \ref{eq:unit-vectors}), respectively.

\section{Application to potential-field modelling}

Potential fields are commonly represented
by functions of the form $f(x - x_{0}, y - y_{0}, z - z_{0})$,
which tends to zero as the differences $x - x_{0}$, $y - y_{0}$,
and $z - z_{0}$ increase.

\subsection{Gravitational modelling}

\subsection{Magnetic modelling}

\section{Conclusions}

TEXT TEXT TEX

%
%\subsection{Acknowledgments}
%Acknowledgments after the main text and before the appendices can be
%included with the
%\texttt{acknowledgments} environment, as
%\begin{verbatim}
%\begin{acknowledgments}
%We wish to thank ...
%\end{acknowledgments}
%\end{verbatim}
%There is also a corresponding \texttt{acknowledgment} environment for a
%single acknowledgment.
%
%\begin{acknowledgments}
%A number of colleagues have helped with suggestions for the
%improvement of this material and I would particularly like to thank
%Bob Geller, University of Tokyo for his criticisms and corrections.
%\end{acknowledgments}

\clearpage

%%%% FIGURES

\begin{figure}
    %\figbox{}{}{
    \includegraphics[width=8cm]{figures/cartesian-geodetic-systems.png}%}
    \caption{Schematic representation of the Cartesian and Geodetic systems.}
    \label{fig:cartesian-geodetic-systems}
\end{figure}

\bibliographystyle{gji}

\bibliography{ref}

\appendix
\section{Appendix} \label{appendA}

TEXT TEXT TEXT

\end{document}

### 2018-04-16 manuscript/gjilguid2e.tex

% gjilguid2e.tex
% V2.0 released 1998 December 18
% V2.1 released 2003 October 7 -- Gregor Hutton, updated the web address for the style files.

\documentclass{gji}
\usepackage{timet}

\title[Geophys.\ J.\ Int.: \LaTeXe\ Guide for Authors]
  {Geophysical Journal International:
   \LaTeXe\ style guide for authors}
\author[B.L.N. Kennett]
  {B.L.N. Kennett$^1$\thanks{Pacific Region Office, GJI} \\
  $^1$ Research School of Earth Sciences, Australian National
    University, Canberra ACT \emph{0200}, Australia
  }
\date{Received 1998 December 18; in original form 1998 November 22}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\volume{200}
\pubyear{1998}

%\def\LaTeX{L\kern-.36em\raise.3ex\hbox{{\small A}}\kern-.15em
%    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}
%\def\LATeX{L\kern-.36em\raise.3ex\hbox{{\Large A}}\kern-.15em
%    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}
% Authors with AMS fonts and mssymb.tex can comment out the following
% line to get the correct symbol for Geophysical Journal International.
\let\leqslant=\leq

\newtheorem{theorem}{Theorem}[section]

\begin{document}

\label{firstpage}

\maketitle

\begin{summary}
 This guide is for authors who are preparing papers for
 \textit{Geophysical Journal International} using the
\LaTeXe\ document preparation system and the GJI class file.
\end{summary}

\begin{keywords}
 \LaTeXe\ -- class files: \verb"gji.cls"\ -- sample text -- user guide.
\end{keywords}

\section{Introduction}

In addition to the standard submission of hardcopy from authors,
\textit{Geophysical Journal International} accepts machine-readable forms
of  papers in \LaTeX.
The layout design for \textit{Geophysical Journal International} has been
implemented as a \LaTeXe\ class file derived from the MN style file for
Monthly Notices of the Royal Astronomical Society. The GJI classfile is
based on the \verb"ARTICLE" style as discussed in the \LaTeX\ manual
\cite {la}. Commands which differ from the standard \LaTeX\ interface, or
which are provided in addition to the standard interface, are explained
in this guide. This guide is not a substitute for the \LaTeX\ manual
itself. Authors planning to submit their papers in \LaTeX\ are advised to
use \verb"gji.cls" as  early as possible in the creation of their
files.  This guide is modified from that produced by Woollatt et al (1994)
to describe the features of the MN style.

A very accessible guide to the features of \LaTeXe and the differences
from the earlier version is provided by Kopka \& Daly \shortcite{kd}.  This
reference provides in chapter 9 a summary of \LaTeX\ error messages and also
a full description of standard \LaTeX\ commands in Appendix F.

\subsection{The GJI document classes}

The use of \LaTeX\ document classes allows a simple change of class
(or class option) to transform the appearance of your document. The
GJI class file preserves the standard \LaTeX\ interface such that any
document which can be produced using the standard \LaTeX\ \verb"ARTICLE"
class can also be produced with the GJI class. However, the measure (or
width of text) is narrower than the default for \verb"ARTICLE", therefore
line breaks will change and long equations may need re-setting.

\subsection{General style issues}

For general style issues, authors are referred to the `Instructions for
Authors' on the inside back cover of \textit{Geophysical Journal
International}. Authors
who are interested in the details of style are referred to Butcher
\shortcite {bu} and The Chicago Manual \shortcite {cm}. The language of
the journal is British English and spelling should conform to this.

Use should be made of symbolic references (\verb"\ref") in order to
protect against late changes of order, etc.

\subsection{Submission of \LaTeX\ articles to the journal}

Papers should initially be submitted in the usual way to:
The Executive Secretary, Royal Astronomical Society, {\em or\/}
the EGS Editor,  {\em or\/} the DGG editor,  {\em or\/}
the American Editor,  {\em or\/} the Pacific Region Editor,
as set out in the Instructions for Authors on the inside back cover of
each issue of Geophysical Journal International.
Four hard copies should be supplied including figures, normally using the
\verb"[referee]" option, for papers with a  high mathematical content the
\verb"[mreferee]" option is recommended. In each case a separate page of
figure captions is preferred.
One of the copies should  be single-sided,
while the other two should  be weight-reduced, by being either
single-spaced or double-sided.   Copies of figures should
also be supplied. Authors should  ensure that their figures are suitable
(in terms of lettering size, etc.)  for the reductions they intend; they
should not attempt to include their figures inside a \TeX\ or \LaTeX\
file by using \verb"\special" or one of the style files for figure
handling.

Note that articles, or revised versions thereof, may not currently be
submitted by electronic mail. However when the article is accepted for
publication the \LaTeX\ file can be sent to the publisher by \verb"ftp"
together with appropriate forms of figures.  Instructions will be provided
following acceptance.

\section{Using the GJI class file}

If the file \verb"gji.cls" is not already in the appropriate
system directory for \LaTeX\ files, either arrange for it to be
put there, or copy it to your working directory. The class file
and related material, such as this guide, can be accessed via the
journal web-site  at
http://www.blackwellpublishing.com/journals/gji under {\em Author
Guidelines}.

The GJI document class is implemented as a complete document class, {\em
not\/} a document class option. In   order to use the GJI style, replace
\verb"article" by
\verb"gji" in the
\verb"\documentclass" command at the beginning of your document:
\begin{verbatim}
\documentclass{article}
\end{verbatim}
is replaced by
\begin{verbatim}
\documentclass{gji}
\end{verbatim}
In general, the following standard document class options should {\em
not\/} be used with the GJI style:
\begin{enumerate}
  \item \texttt{10pt}, \texttt{11pt}, \texttt{12pt} -- unavailable;
  \item \texttt{twoside} (no associated style file) --
     \texttt{twoside} is the default;
  \item \texttt{fleqn}, \texttt{leqno}, \texttt{titlepage} --
        should not be used (\verb"fleqn" is already incorporated into
        the GJI style);
  \item \texttt{twocolumn} -- is not necessary as it is the default style.
\end{enumerate}

In \LaTeX2e the use of postscript fonts and the inclusion of non-standard
options is carried out through the \verb"\usepackage" command, rather than
as options as in earlier versions.  Thus the Times font can be used for
text by including
\begin{verbatim}
\usepackage{times}
\end{verbatim}
on the line immediately after the \verb"\documentclass". If necessary,
\texttt{ifthen} and \texttt{bezier} can be included as packages.

The GJI class file has been designed to operate with the standard
version of \verb"lfonts.tex" that is distributed as part of \LaTeX
. If you have access to the source file for this guide,
\verb"gjilguid2e.tex", attempt to typeset it.  If you find font
problems you might investigate whether a non-standard version of
\verb"lfonts.tex" has been installed in your system.

\subsection{Additional document class options}\label{classoptions}

The following additional class options are available with the GJI style:
\begin{description}
  \item \texttt{onecolumn} -- to be used \textit{only} when two-column output
        is unable to accommodate long equations;
  \item \texttt{landscape} -- for producing wide figures and tables which
        need to be included in landscape format (i.e.\ sideways) rather
        than portrait (i.e.\ upright). This option is described below.
  \item \texttt{doublespacing} -- this will double-space your
        article by setting \verb"\baselinestretch" to 2.
  \item \texttt{referee} -- 12/20pt text size, single column,
        designed for submission of papers.
  \item \texttt{mreferee} -- 11/17pt text size, single column
        designed for submission of papers with mathematical content.
  \item \texttt{camera} -- designed for use with computer modern fonts to
        produce a closer representation of GJI style for camera
        ready material.
  \item \texttt{galley} -- no running heads, no attempt to align
        the bottom of columns.
\end{description}

\subsection{Landscape pages}

If a table or illustration is too wide to fit the standard measure, it
must be turned, with its caption, through 90 degrees anticlockwise.
Landscape illustrations and/or tables cannot be produced directly
using the GJI style file because \TeX\ itself cannot turn the
page, and not all device drivers provide such a facility.
The following procedure can be used to produce such pages.
\begin{enumerate}
  \item Use the \verb"table*" or \verb"figure*" environments in your
        document to create the space for your table or figure on the
        appropriate page of your document. Include an empty
        caption in this environment to ensure the correct
        numbering of subsequent tables and figures. For instance, the
        following code prints a page with the running head, a message
        half way down and the figure number towards the bottom. If you
        are including a plate, the running headline is different, and you
        need to key in the three lines which are marked with \verb"% **",
        with an appropriate headline.
\begin{verbatim}
% ** \clearpage
% ** \thispagestyle{plate}
% ** \plate{Opposite p.~812, GJI, \textbf{135}}
\begin{figure*}
  \vbox to220mm{\vfil Landscape figure to
                go here. \vfil}
  \caption{}
  \label{landfig}
\end{figure*}
\end{verbatim}
\item Create a separate document with the corresponding document style
      but also with the \verb"landscape" document style option, and
      include the \verb"\pagestyle" command, as follows:
\begin{verbatim}
\documentclass[landscape]{gji}
\pagestyle{empty}
\end{verbatim}
  \item Include your complete tables and illustrations (or space for
        these) with captions using the \verb"table*" and \verb"figure*"
        environments.
  \item Before each float environment, use the
        \verb"\setcounter" command to ensure the correct numbering of
        the caption. For example,
\begin{verbatim}
\setcounter{table}{0}
\begin{table*}
 \begin{minipage}{115mm}
 \caption{Images of global seismic tomography.}
 \label{tab1}
 \begin{tabular}{@{}llllcll}
   :
 \end{tabular}
 \end{minipage}
\end{table*}
\end{verbatim}
The corresponding example for a figure would be:
\begin{verbatim}
\clearpage
\setcounter{figure}{12}
\begin{figure*}
 \vspace{144mm}
 \caption{Travel times for regional model.}
 \label{fig13}
\end{figure*}
\end{verbatim}
\end{enumerate}

\section{Additional facilities}

In addition to all the standard \LaTeX\ design elements, the GJI style
includes the following features.
\begin{enumerate}
  \item Extended commands for specifying a short version of the title and
        author(s) for the running headlines;
  \item A \verb"summary" environment to produce a suitably indented
        Summary
  \item An \verb"abstract" environment which produces the GJI style of
        Summary
  \item A \verb"keywords" environment and a \verb"\nokeywords" command;
  \item Use of the \verb"description" environment for unnumbered lists.
  \item A starred version of the \verb"\caption" command to produce
        captions for continued figures or tables.
 \end{enumerate}
 In general, once you have used the additional \verb"gji.cls" facilities
in your document, do not process it with a standard \LaTeX\ style file.

\subsection{Titles and author's name}

In the GJI style, the title of the article and the author's name (or
authors' names) are used both at the beginning of the article for the
main title and throughout the article as running headlines at the top
of every page. The title is used on odd-numbered pages (rectos) and the
author's name appears on even-numbered pages (versos). Although the
main heading can run to several lines of text, the running headline
must be a single line ($\leqslant 45$ characters). Moreover, the main
heading can also incorporate new line commands (e.g. \verb"\\") but
these are not acceptable in a running headline. To enable you to
specify an alternative short title and an alternative short author's
name, the standard \verb"\title" and \verb"\author" commands have been
extended to take an optional argument to be used as the running
headline. The running headlines for this guide were produced using the
following code:
\begin{verbatim}
\title[Geophys.\ J.\ Int.:
       \LaTeXe\ Guide for Authors]
  {Geophysical Journal International:
   \LaTeXe\ style guide for authors}
\end{verbatim}
and
\begin{verbatim}
\author[B.L.N. Kennett]
   {B.L.N. Kennett$^1$
  \thanks{Pacific Region Office, GJI} \\
  $^{1}$Research School of Earth Sciences,
    Australian National University,
    Canberra ACT \emph{0200}, Australia
  }
\end{verbatim}
The \verb"\thanks" note produces a footnote to the title or author.

\subsection{Key words and Summary}

At the beginning of your article, the title should be generated in the
usual way using the \verb"\maketitle" command. Immediately following
the title you should include a Summary followed by a list of key
words. The summary should be enclosed within an \verb"summary"
environment, followed immediately by the key words enclosed in a
\verb"keywords" environment. For example, the titles for this guide
were produced by the following source:
\begin{verbatim}
\maketitle
\begin{summary}
 This guide is for authors who are preparing
 papers for \textit{Geophysical Journal
 International} using the \LaTeXe\ document
 preparation system and the GJI style file.
\end{summary}
\begin{keywords}
 \LaTeXe\ -- class files: \verb"gji.cls"\ --
 sample text -- user guide.
\end{keywords}

\section{Introduction}
  :
\end{verbatim}
The heading `\textbf{Key words}' is included automatically and the key
words are followed by vertical space. If, for any reason, there are no
key words, you should insert the \verb"\nokeywords" command immediately
after the end of the \verb"summary" or \verb"abstract" environment. This
ensures that the   vertical space after the abstract and/or title is
correct and that any
\verb"thanks" acknowledgments are correctly included at the bottom of
the first column. For example,
\begin{verbatim}
\maketitle
\begin{abstract}
  :
\end{abstract}
\nokeywords

\section{Introduction}
  :
\end{verbatim}

Note that the \verb"summary" and \verb"abstract" environments have the same
effect for the documentclass \verb"gji.cls"

\subsection{Lists}

The GJI style provides numbered lists using the \verb"enumerate"
environment and unnumbered lists using the \verb"description"
environment with an empty label. Bulleted lists are not part of the GJI
style and the \verb"itemize" environment should not be used.

The enumerated list numbers each list item with roman numerals:
\begin{enumerate}
  \item first item
  \item second item
  \item third item
\end{enumerate}
Alternative numbering styles can be achieved by inserting a
redefinition of the number labelling command after the
\verb"\begin{enumerate}". For example, the list
\begin{enumerate}
\renewcommand{\theenumi}{(\arabic{enumi})}
  \item first item
  \item second item
  \item etc\ldots
\end{enumerate}
was produced by:
\begin{verbatim}
\begin{enumerate}
 \renewcommand{\theenumi}{(\arabic{enumi})}
  \item first item
       :
\end{enumerate}
\end{verbatim}
Unnumbered lists are provided using the \verb"description" environment.
For example,
\begin{description}
  \item First unnumbered item which has no label and is indented from
        the left margin.
  \item Second unnumbered item.
  \item Third unnumbered item.
\end{description}
was produced by,
\begin{verbatim}
\begin{description}
 \item First unnumbered item...
 \item Second unnumbered item.
 \item Third unnumbered item.
\end{description}
\end{verbatim}

\subsection{Captions for continued figures and tables}

The \verb"\caption*" command may be used to produce a caption with the
same number as the previous caption (for the corresponding type of
float). For instance, if a very large table does not fit on one page,
it must be split into two floats; the second float should use the
\verb"caption*" command with a suitable caption:
\begin{verbatim}
\begin{table}
 \caption*{-- \textit{continued}}
  \begin{tabular}{@{}lccll}
  :
  \end{tabular}
\end{table}
\end{verbatim}

 \begin{figure}
     \vspace{5.5cm}
     \caption{An example figure in which space has been
              left for the artwork.}
     \label{sample-figure}
  \end{figure}

\section[]{Some guidelines for using\\* standard facilities}

The following notes may help you achieve the best effects with the GJI
style file.

\subsection{Sections}

\LaTeX\ provides five levels of section headings and they are all
defined in the GJI style file:
\begin{description}
  \item \verb"\section"
  \item \verb"\subsection"
  \item \verb"\subsubsection"
  \item \verb"\paragraph"
  \item \verb"\subparagraph"
\end{description}
Section numbers are given for section, subsection, subsubsection
and paragraph headings.  Section headings are automatically converted to
upper case; if you need any other style, see the example in section~\ref{headings}.

If you find your section/subsection (etc.)\ headings are wrapping round,
you must use the \verb"\\*" to end individual lines and include the
optional argument \verb"[]" in the section command. This ensures that
the turnover is flushleft.

\subsection{Illustrations (or figures)}

\begin{figure*}
 \vspace{5.5cm}
     \caption{An example figure spanning two-columns
             in which space has been left for the artwork.}
     \label{twocol-figure}
\end{figure*}

The GJI style will cope with positioning of your illustrations and
you should not use the positional qualifiers on the
\verb"figure" environment which would override these decisions. See
`Instructions for Authors' in {\em Geophysical Journal International\/}
for submission of
artwork. Figure captions should be below the figure itself, therefore
the \verb"\caption" command should appear after the figure or space
left for an illustration. For example, Fig.~\ref{sample-figure} is
produced using the following commands:
\begin{verbatim}
\begin{figure}
 \vspace{5.5cm}
 \caption{An example figure in which space has
          been left for the artwork.}
 \label{sample-figure}
\end{figure}
\end{verbatim}

Where a figure needs to span two-columns the \verb"figure*" environment
should be used as in  Fig.~\ref{twocol-figure} using the following commands
\begin{verbatim}
\begin{figure*}
 \vspace{5.5cm}
   \caption{An example figure spanning two-columns
     in which space has been left for the artwork.}
   \label{twocol-figure}
\end{figure*}
\end{verbatim}

\subsection{Tables}

The GJI style will cope with positioning of your tables and you
should not use the positional qualifiers on the
\verb"table" environment which would override these decisions. Table
captions should be at the top, therefore the \verb"\caption" command
should appear before the body of the table.

The \verb"tabular" environment can be used to produce tables with
single horizontal rules, which are allowed, if desired, at the head and
foot only. This environment has been modified for the GJI style in the
following ways:
\begin{enumerate}
  \item additional vertical space is inserted on either side of a rule;
  \item vertical lines are not produced.
\end{enumerate}
Commands to redefine quantities such as \verb"\arraystretch" should be
omitted. For example, Table~\ref{symbols} is produced using the
following commands.
\begin{table}
 \caption{Seismic velocities at major discontinuities.}
 \label{symbols}
 \begin{tabular}{@{}lcccccc}
  Class & depth & radius
        & $\alpha _{-}$ & $\alpha _{+}$
        & $\beta _{-}$ & $\beta _{+}$ \\
  ICB & 5154 & 1217 & 11.091 & 10.258
        & 3.438 &  0. \\
  CMB & 2889 & 3482 & 8.009 & 13.691
        & 0. & 7.301 \\
 \end{tabular}

 \medskip
 The ICB represents the boundary between the inner and outer cores and
the CMB the boundary between the core and the mantle.  Velocities with
subscript $-$ are evaluated just below the discontinuity and
those with subscript $+$ are evaluated just above the discontinuity.
\end{table}
\begin{verbatim}
\begin{table}
 \caption{Seismic velocities at major
          discontinuities.}
 \label{symbols}
 \begin{tabular}{@{}lcccccc}
  Class & depth & radius
        & $\alpha _{-}$ & $\alpha _{+}$
        & $\beta _{-}$ & $\beta _{+}$ \\
  ICB & 5154 & 1217 & 11.091 & 10.258
        & 3.438 &  0. \\
  CMB & 2889 & 3482 & 8.009 & 13.691
        & 0. & 7.301 \\
 \end{tabular}

 \medskip
 The ICB represents the boundary ...
... evaluated just above the discontinuity.

\end{table}
\end{verbatim}

If you have a table that is to extend over two columns, you need to use
\verb"table*" in a minipage environment, i.e., you can say
\begin{verbatim}
\begin{table*}
\begin{minipage}{80mm}
 \caption{Caption which will wrap round to the
          width of the minipage environment.}
 \begin{tabular}{%
      :
 \end{tabular}
\end{minipage}
\end{table*}
\end{verbatim}
The width of the minipage should more or less be the width of your table,
so you can only guess on a value on the first pass. The value will have to
be adjusted when your article is finally typeset, so don't worry
about making it the exact size.

\subsection{Running headlines}

As described above, the title of the article and the author's name (or
authors' names) are used as running headlines at the top of every page.
The headline on right pages can list up to three names; for more than
three use et~al. The \verb"\pagestyle" and \verb"\thispagestyle"
commands should {\em not\/} be used. Similarly, the commands
\verb"\markright" and \verb"\markboth" should not be necessary.

\subsection{Typesetting mathematics}

\subsubsection{Displayed mathematics}

The GJI style will set displayed mathematics flush with the left margin,
provided that you use the \LaTeX\ standard of open and closed square brackets
as delimiters. The equation
\[
 \sum_{i=1}^p \lambda_i =
{\mathrm{trace}}(\mathbf{S})
\]
was typeset in the GJI style using the commands
\begin{verbatim}
\[
 \sum_{i=1}^p \lambda_i =
{\mathrm{trace}}(\mathbf{S})
\]
\end{verbatim}
This correct positioning should be compared with that for
the following centred equation,
$$ \alpha_{j+1} > \bar{\alpha}+ks_{\alpha} $$
which was (wrongly) typeset using double dollars as follows:
\begin{verbatim}
$$ \alpha_{j+1} > \bar{\alpha}+ks_{\alpha} $$
\end{verbatim}
Note that \verb"\mathrm" will produce a roman character
in math mode.

For numbered equations use the \verb"equation" and \verb"eqnarray"
environments which will give the correct positioning.
If equation numbering by section is required the command
\verb"\eqsecnum" should appear after \verb"begin{document}"
at the head of the file.

\subsubsection{Bold math italic}\label{boldmathitalic}

The class file provides a font \verb"\mitbf" defined as:
\begin{verbatim}
\newcommand{\mitbf}[1]{
  \hbox{\mathversion{bold}$#1$}}
\end{verbatim}
Which can be used as follows, to typset the equation
\begin{equation}
  d(\mitbf{{s_{t_u}}}) = \langle [RM(\mitbf{{x_y}}
  + \mitbf{{s_t}}) - RM(\mitbf{{x_y}})]^2 \rangle
\end{equation}
the input should be
\begin{verbatim}
\begin{equation}
  d(\mitbf{{s_{t_u}}}) = \langle [RM(\mitbf{{x_y}}
  + \mitbf{{s_t}}) - RM(\mitbf{{x_y}})]^2 \rangle
\end{equation}
\end{verbatim}

If you are using version 1 of the New Font Selection Scheme, you may
have some messages in your log file that read something like ``Warning:
Font/shape `cmm/b/it' in size~\hbox{$< \!\! 9 \!\! >$} not available
on input line 649. Warning: Using external font `cmmi9' instead on input
line 649.'' If you have such messages, your system will have substituted
math italic characters where you wanted bold math italic ones: you
are advised to upgrade to version 2.

\subsubsection{Bold Greek}\label{boldgreek}

To get bold Greek you use the same
method as for bold math italic. Thus you can input
\begin{verbatim}
\[ \mitbf{{\alpha_{\mu}}} =
\mitbf{\Theta} \alpha. \]
\end{verbatim}
to typeset the equation
\[ \mitbf{{\alpha_{\mu}}} = \mitbf{\Theta} \alpha . \]

\subsection{Points to note in formatting text}\label{formtext}

A number of text characters require special attention
so that \LaTeX\ can properly format a file.

The following characters must be preceded by a
backslash or \LaTeX\ will interpret them as commands:
\begin{quote}
~~~~~~~~~\$~~~\&~~~\%~~~\#~~~\_~~~\{~~~and~~~\}
\end{quote}
must be typed
\begin{center}
\begin{quote}
~~~~~~\verb"\$"~~~\verb"\&"~~~\verb"\%"~~~\verb"\#"
~~~\verb"\_"~~~\verb"\{"~~~and~~~\verb"\}".
\end{quote}
\end{center}

\LaTeX\ interprets all double quotes as closing quotes.
Therefore quotation marks must be typed as pairs of
opening and closing single quotes, for example,
\texttt{ ``quoted text.''}

Note that \LaTeX\ will not recognize greater than or
less than symbols unless they are typed within math
commands (\verb"$>$" or \verb"$<$").

\subsubsection{Special symbols}

The macros for the special symbols in Tables~\ref{mathmode}
and~\ref{anymode}
have been taken from the Springer Verlag `Astronomy and Astrophysics'
design, with their permission. They are directly compatible and use the
same macro names.
These symbols will work in all text sizes, but are only guaranteed to work
in text and displaystyles. Some of the symbols will not get any smaller
when they are used in sub- or superscripts, and will therefore be
displayed at the wrong size. Don't worry about this as the typesetter
will be able to sort this out.
%
\begin{table*}
\begin{minipage}{106mm}
\caption{Special symbols which can only be used in math mode.}
\label{mathmode}
\begin{tabular}{@{}llllll}
Input & Explanation & Output & Input & Explanation & Output\\
\hline
\verb"\la"     & less or approx       & $\la$     &
  \verb"\ga"     & greater or approx    & $\ga$\\[2pt]
\verb"\getsto" & gets over to         & $\getsto$ &
  \verb"\cor"    & corresponds to       & $\cor$\\[2pt]
\verb"\lid"    & less or equal        & $\lid$    &
  \verb"\gid"    & greater or equal     & $\gid$\\[2pt]
\verb"\sol"    & similar over less    & $\sol$    &
  \verb"\sog"    & similar over greater & $\sog$\\[2pt]
\verb"\lse"    & less over simeq      & $\lse$    &
  \verb"\gse"    & greater over simeq   & $\gse$\\[2pt]
\verb"\grole"  & greater over less    & $\grole$  &
  \verb"\leogr"  & less over greater    & $\leogr$\\[2pt]
\verb"\loa"    & less over approx     & $\loa$    &
  \verb"\goa"    & greater over approx  & $\goa$\\
\hline
\end{tabular}
\end{minipage}
\end{table*}
%
\begin{table*}
\begin{minipage}{115mm}
\caption{Special symbols which don't have to be
used in math mode.}
\label{anymode}
\begin{tabular}{@{}llllll}
Input & Explanation & Output & Input & Explanation & Output\\
\hline
\verb"\sun"      & sun symbol            & $\sun$     &
  \verb"\earth"     & earth symbol         & $\earth$   \\[2pt]
\verb"\degr"     & degree                &$\degr$     &
  \verb"\micron"   & \micron               & \micron    \\[2pt]
\verb"\diameter" & diameter              & \diameter  &
  \verb"\sq"       & square                & \squareforqed\\[2pt]
\verb"\fd"       & fraction of day       & \fd        &
  \verb"\fh"       & fraction of hour      & \fh\\[2pt]
\verb"\fm"       & fraction of minute    & \fm        &
  \verb"\fs"       & fraction of second    & \fs\\[2pt]
\verb"\fdg"      & fraction of degree    & \fdg       &
  \verb"\fp"       & fraction of period    & \fp\\[2pt]
\verb"\farcs"    & fraction of arcsecond & \farcs     &
  \verb"\farcm"    & fraction of arcmin    & \farcm\\[2pt]
\verb"\arcsec"   & arcsecond             & \arcsec    &
  \verb"\arcmin"   & arcminute             & \arcmin\\

\hline
\end{tabular}
\end{minipage}
\end{table*}

The command \verb"\chemical" is provided to set chemical species with
an even level for subscripts (not produced in standard mathematics mode).
Thus \verb"\chemical{Fe_{2}^{2+}Cr_{2}O_{4}}" will produce
\chemical{Fe_{2}^{2+}Cr_{2}O_{4}}.

\subsection{Bibliography}

Two methods are provided for managing citations and
references.   The first approach uses the
\verb"\begin{thebibliography}{}"
and \verb"\end{thebibliography}{}" commands.

The second approach uses a simplified scheme using
\verb"\begin{references}" and \verb"\end{references}" commands.

References to published literature should be quoted in text by author
and date; e.g. Draine (1978) or (Begelman, Blandford \& Rees 1984).
Where more than one reference is cited having the same author(s) and date,
the letters a,b,c, \ldots\ should follow the date; e.g.\ Smith (1988a),
Smith (1988b), etc.
The first time you introduce a three-author paper, you should list all
three authors at the first citation, and thereafter, use et al.

\subsubsection{Biblography References in the text}

References in the text are given by author and date, and, whichever
method is used to produce the bibliography, the references in the text
are done in the same way. Each bibliographical entry has a key, which
is assigned by the author and used to refer to that entry in the text.
There is one form of citation -- \verb"\cite{key}" -- to produce the
author and date, and another form -- \verb"\shortcite{key}" -- which
produces the date only. Thus, Rutherford \& Hawker \shortcite{rh} is
produced by
\begin{verbatim}
Rutherford \& Hawker \shortcite{rh}
\end{verbatim}
while \cite{hi} is produced by
\begin{verbatim}
\cite{hi}
\end{verbatim}

\subsubsection{The bibliography}

The following listing shows some references prepared in the style of
the journal; the code produces the references at the end of this guide.
The following rules apply for the ordering of your references:
\begin{enumerate}
  \item if an author has written several papers, some with other authors,
        the rule is that the single-author papers precede the two-author
        papers, which, in turn, precede the multi-author papers;
  \item within the two-author paper citations, the order is determined
        by the second author's surname, regardless of date;
  \item within the multi-author paper citiations, the order is
        chronological, regardless of author's surnames.
\end{enumerate}
%
\begin{verbatim}
\begin{thebibliography}{}
  \bibitem[\protect\citename{Butcher }1992]{bu}
    Butcher J. 1992. \textit{Copy-editing: The
    Cambridge Handbook}, 3rd edn, Cambridge
    Univ. Press, Cambridge.
  \bibitem[\protect\citename{The Chicago Manual }%
    1982]{cm} \textit{The Chicago Manual of Style},
    Univ. Chicago Press, Chicago, 1982.
  \bibitem[\protect\citename{Chao }1985]{ch}
    Chao, B. F., 1985. Normal mode study of the
    Earth's rigid body motions,
    \textit{Geophys. Res. Lett.}, \textbf{12}, 526-529.
  \bibitem[\protect\citename{Hinderer }1986]{hi}
    Hinderer, J., 1986. Resonance effects of the
    earth's fluid core in earth rotation,
    in \textit{Solved and Unsolved Problems},
    pp. 277-296, ed. Cazenave A., Reidel,
    Dordrecht.
  \bibitem[\protect\citename{Lamport }1986]{la}
    Lamport L., 1986,  \LaTeX: \textit{A Document
    Preparation System}, Addison--Wesley, New York
  \bibitem[\protect\citename{Lindberg }1986]{li}
    Lindberg, C., 1986.  Multiple taper harmonic
    analysis of terrestrial free oscillations,
    \textit{PhD thesis}, University of California.
  \bibitem[\protect\citename{Maupin }1992]{ma}
    Maupin, V., 1992. Modelling of laterally
    trapped surface waves with application to
    Rayleigh waves in the Hawaiian swell,
    \textit{Geophys. J. Int.}, \textbf{110}, 553-570.
  \bibitem[\protect\citename{Rutherford
    \& Hawker }1981]{rh} Rutherford, S. R.
    \& Hawker, K. E.,  1981, Consistent coupled
    mode theory of sound propagation for a
    class of non-separable problems,
    \textit{J. acoust. Soc. Am.}, \textbf{71},
    554-564
\end{thebibliography}
\end{verbatim}
Each entry takes the form
\begin{verbatim}
\bibitem[\protect\citename{Author(s), }%
  Date]{tag} Bibliography entry
\end{verbatim}
where \verb"Author(s)" should be the author names as they are cited in
the text, \verb"Date" is the date to be cited in the text, and
\verb"tag" is the tag that is to be used as an argument for the
\verb"\cite{}" and \verb"\shortcite{}" commands. \verb"Bibliography entry"
should be the material that is to appear in the bibliography,
suitably formatted.

\subsubsection{Simplified References and Citations}

The second approach to referencing is taken with permission from
the American Geophysical Union Latex macros

The reference section is started using a
\verb"\begin{references}" command which will
automatically produce a correctly formatted
``Reference'' head.  Each reference is then
preceded by a \verb"\reference"
command.  It is the author's
responsibility to place bibliographic reference
information in the proper order with correct
punctuation.  After the last reference in your
reference section, type an \verb"\end{references}"
command.

Authors may enter properly formatted citations directly
in the manuscript text and enclose those citations in
\verb"\markcite{}" commands.  This approach
marks all citations in your manuscript, but there
is no interaction between the \verb"\markcite"
commands and the reference section.

To create in-text citations, enclose each citation
within a \verb"\markcite" command.
There are two ways to include in-text citations,
depending on the way you phrase your sentence.
You may either include an entire reference within
brackets \markcite{(Merritt et al., 1996)} or you
may mention the author as part of your sentence and
include only the year in brackets, as in \markcite{Ono (1996)}.

As an example
\begin{verbatim}
\begin{references}
\reference
Azimi, Sh.A., Kalinin, A.Y., Kalinin, V.B.,
\& Pivovarov, B.L., 1968.
Impulse and transient characteristics of media
with linear and quadratic absorption laws,
\textit{Izv. Earth Phys.} (English Transl.),
\textbf{2}, 88--93.
\reference
Dahlen, F.A., \& Smith, M.L., 1975.
The influence of rotation on the free
oscillations of the Earth,
\textit{Phil. Trans. R. Soc. London Ser. A},
\textbf{279}, 143--167.
\reference
Durek, J.J., Ritzwoller, M.H.,
\& Woodhouse, J.H., 1993.
Constraining upper mantle anelasticity
using surface wave amplitude anomalies,
\gji, \textbf{114}, 249--272.
\end{references}
\end{verbatim}
produces the reference list
\begin{references}
 \reference
Azimi, Sh.A., Kalinin, A.Y., Kalinin, V.B.,
\& Pivovarov, B.L., 1968.
Impulse and transient characteristics of media
with linear and  quadratic absorption laws,
\textit{Izv. Earth Phys.} (English Transl.),
\textbf{2}, 88--93.
\reference
Dahlen, F.A., \& Smith, M.L., 1975.
The influence of rotation on the free oscillations of the Earth,
\textit{Phil. Trans. R. Soc. London Ser. A}, \textbf{279}, 143--167.
\reference
Durek, J.J., Ritzwoller, M.H., \& Woodhouse, J.H., 1993.
Constraining upper mantle anelasticity
using surface wave amplitude anomalies,
\gji \textbf{114}, 249--272.
\end{references}

\subsubsection{Common Journals}

The following abbreviations are provided for commonly cited
journals and can be used directly in the bibliography.

In the following table the abbreviation and the form of the
associated entry are presented
\newline
\begin{tabular}{ll}
\verb"\areps" & \areps \\
\verb"\bssa"  & \bssa \\
\verb"\eos"   & \eos  \\
\verb"\eps"   & \eps \\
\verb"\epsl"  & \epsl \\
\verb"\gca"   & \gca \\
\verb"\geo"   & \geo \\
\verb"\geop"  & \geop \\
\verb"\gji"   & \gji \\
\verb"\gjras" & \gjras \\
\verb"\grl"   & \grl \\
\verb"\gsab"  & \gsab \\
\verb"\gs"    & \gs \\
\verb"\jgr"   & \jgr \\
\verb"\jseis" & \jseis \\
\verb"\mnras" & \mnras \\
\verb"\pag"   & \pag \\
\verb"\pepi"  & \pepi \\
\verb"\rg"    & \rg \\
\verb"\tecto" & \tecto \\
\end{tabular}
%
% The following two tables have been moved back in the text to
% improve page layout
%
\begin{table*}
\begin{minipage}{130mm}
\caption{Authors' notes.}
\label{authors}
\begin{tabular}{@{}ll}
\verb"\title[optional short title]{long title}"
                    & short title used in running head\\
\verb"\author[optional short author(s)]{long author(s)}"
                    & short author(s) used in running head\\
\verb"\begin{abstract}...\end{abstract}"& for summary on
titlepage\\
\verb"\begin{summary}...\end{summary}"& for abstract on
titlepage\\
\verb"\begin{keywords}...\end{keywords}"& for keywords on titlepage\\
\verb"\nokeywords"  & if there are no keywords on titlepage\\
\verb"\begin{figure*}...\end{figure*}" & for a double spanning figure in two-column mode\\
\verb"\begin{table*}...\end{table*}" & for a double spanning table in
                                       two-column mode\\
\verb"\caption*"    & for continuation figure captions\\
\verb"\resetfigno" & resets figures numbers after an appendix\\
\verb"[referee]" & documentclass option for 12/20pt, single col,
                   for manuscript submission\\
\verb"[mreferee]" & documentclass option for 11/17pt, single col,
                   for submission of papers with extensive mathematics\\
\end{tabular}
\end{minipage}
\end{table*}
%
\begin{table*}
\begin{minipage}{130mm}
\caption{Editors' notes.}
\label{editors}
\begin{tabular}{@{}lp{270pt}}
\verb"\pagerange{000--000}"& for catchline, note use of en-rule\\
\verb"\pagerange{L00--L00}"& for letters option, used in catchline\\
\verb"\volume{000}" & volume number, for catchline\\
\verb"\pubyear{0000}" & publication year, for catchline\\
\verb"\microfiche{GJI000/0}" & for articles accompanied by microfiche\\
\verb"\journal" & replace the whole catchline at one go\\
\verb"[doublespacing]" & documentclass option for doublespacing\\
\verb"[galley]" & documentclass option for running to galley\\
\verb"[landscape]" & documentclass option for landscape illustrations\\
\verb"[fasttrack]" & documentclass option, for rapid short communications
                   (adds F to folios)\\
\verb"[onecolumn]" & documentclass option for one-column \\
\verb"\bsp"& typesets the final phrase `This paper has been produced
 using the Blackwell Publishing GJI \LaTeX2e\ class file.'\\
\end{tabular}
\end{minipage}
\end{table*}

\subsection{Appendices}

The appendices in this guide were generated by typing:
\begin{verbatim}
\appendix
\section{For authors}
     :
\section{For editors}
\end{verbatim}
You only need to type \verb"\appendix" once. Thereafter, every
\verb"\section" command will generate a new appendix which will be
numbered A, B, etc.

If figure captions are to provided after an appendix the figure number can
be reset to avoid extraneous labelling using the command
\verb"\resetfigno".

\section[]{Example of section heading with\\*
  {\mdseries \textsc{S}\lowercase{\textsc{mall}}
  \textsc{C}\lowercase{\textsc{aps}}},
  \lowercase{lowercase},
  \textit{ italic}, and bold\\* Greek such as
  $\mitbf{{\mu^\kappa}}$}\label{headings}

This can be built up using text commands and the \verb"mitbf"
command introduced above

\begin{verbatim}
\section[]{Example of section heading with\\*
  {\mdseries \textsc{S}\lowercase{\textsc{mall}}
  \textsc{C}\lowercase{\textsc{aps}}},
  \lowercase{lowercase},
  \textit{ italic}, and bold\\* Greek such as
  $\mitbf{{\mu^\kappa}}$}\label{headings}
\end{verbatim}

\subsection{Acknowledgments}
Acknowledgments after the main text and before the appendices can be
included with the
\texttt{acknowledgments} environment, as
\begin{verbatim}
\begin{acknowledgments}
We wish to thank ...
\end{acknowledgments}
\end{verbatim}
There is also a corresponding \texttt{acknowledgment} environment for a
single acknowledgment.

\begin{acknowledgments}
A number of colleagues have helped with suggestions for the
improvement of this material and I would particularly like to thank
Bob Geller, University of Tokyo for his criticisms and corrections.
\end{acknowledgments}

\begin{thebibliography}{}
  \bibitem[\protect\citename{Butcher }1992]{bu}
    Butcher J. 1992. \textit{Copy-editing: The Cambridge
    Handbook}, 3rd edn, Cambridge Univ. Press, Cambridge.
  \bibitem[\protect\citename{The Chicago Manual }%
    1982]{cm} \textit{The Chicago Manual of Style}, Univ.
    Chicago Press, Chicago, 1982.
  \bibitem[\protect\citename{Chao }1985]{ch}
    Chao, B. F., 1985. Normal mode study of the Earth's rigid
     body motions, \textit{Geophys. Res. Lett.}, \textbf{12}, 526-529.
  \bibitem[\protect\citename{Hinderer }1986]{hi}
    Hinderer, J., 1986. Resonance effects of the earth's fluid
    core in earth rotation, in \textit{Solved and Unsolved
    Problems}, pp. 277-296, ed. Cazenave A., Reidel,
    Dordrecht.
  \bibitem[\protect\citename{Kopka \& Daly}1995]{kd}
    Kopka H. \& Daly P.W., 1995, \textit{A guide to} \LaTeX2e,
    Addison--Wesley, New York
  \bibitem[\protect\citename{Lamport }1986]{la}
    Lamport L., 1986,  \LaTeX: \textit{A Document
    Preparation System}, Addison--Wesley, New York
  \bibitem[\protect\citename{Lindberg }1986]{bl}
    Lindberg, C., 1986.  Multiple taper harmonic analysis of
    terrestrial free oscillations, \textit{PhD thesis},
    University of California.
  \bibitem[\protect\citename{Maupin }1992]{ma}
    Maupin, V., 1992. Modelling of laterally trapped surface
    waves with application to Rayleigh waves in the Hawaiian
    swell, \textit{Geophys. J. Int.}. \textbf{110}, 553-570.
  \bibitem[\protect\citename{Rutherford \& Hawker }%
    1981]{rh}  Rutherford, S. R. \& Hawker, K. E., 1981,
    Consistent coupled mode theory of sound propagation for a
    class of non-separable problems,
   \textit{J. acoust. Soc. Am.}, \textbf{71}, 554-564
\end{thebibliography}

\appendix
\section{For authors}

Table~\ref{authors} is a list of design macros which are unique to GJI. The
list displays each macro's name and description.

\section{For editors}

The additional features shown in Table~\ref{editors} may be used for
production purposes.

\bsp % ``This paper has been produced using the Blackwell
     %   Publishing GJI \LaTeXe\ class file.''

\label{lastpage}

\end{document}

### 2018-04-16 manuscript/letter.tex

@@
 \textit{gravity vector} and its intensity is called simply 
-\textit{gravity} \citep{hofmann-wellenhof-moritz2005}.
+\textit{gravity} \citep{heiskanen-moritz1967,
+hofmann-wellenhof-moritz2005}.
 In the case of gravimetry on moving platforms (e.g., airplanes,
@@
 is called meridian plane (gray plane in Figure \ref{fig:fig1}).
-
 At a given point $(h, \varphi, \lambda)$, there are three 
@@
 \end{equation}
-Notice that these unit vectors are mutually orthogonal.
-The normal gravity vector $\mitbf{\gamma}_{P}$ at a point $P$
+Notice that these unit vectors are mutually orthogonal and
+that the normal gravity vector $\mitbf{\gamma}_{P}$ at a point $P$
 is opposite to the unit vector $\hat{\mitbf{u}}_{P}$ at $P$.
@@
 In applied geophysics, these density differences are generally 
-called ``anomalous masses" (e.g., \citeauthor{hammer1945}, 
+called \textit{anomalous masses} (e.g., \citeauthor{hammer1945}, 
 \citeyear{hammer1945}; \citeauthor{lafehr1965}, \citeyear{lafehr1965}),
-``density anomalies" (e.g., \citeauthor{forsberg1984}, \citeyear{forsberg1984})
-or ``gravity sources" (e.g., \citeauthor{blakely1996}, 
+\textit{density anomalies} (e.g., \citeauthor{forsberg1984}, \citeyear{forsberg1984})
+or \textit{gravity sources} (e.g., \citeauthor{blakely1996}, 
 \citeyear{blakely1996}). Here, we opted for using the last term.
@@
 The difference between the observed gravity and the
-normal gravity, at the same point, is called gravity disturbance
-\citep{hofmann-wellenhof-moritz2005}.
+normal gravity, at the same point, is called \textit{gravity disturbance}
+\citep{heiskanen-moritz1967, hofmann-wellenhof-moritz2005}.
 Notice that the gravity disturbance is not equivalent to the
 magnitude of the difference between the gravity vector
-and the normal gravity vector at the same point.
+and the normal gravity vector at the same point 
+\citep{barthelmes2013, sanso_sideris2013}.
 As properly pointed out by \citet{hackney-featherstone2003},
@@
 
+For each point $(x_{i}, y_{i}, z_{i})$ in the topocentric
+Cartesian coordinate system (Figure \ref{fig:fig2}b), there is
+a corresponding point $(h_{i}, \varphi_{i}, \lambda_{i})$
+in the geodetic coordinate system (Figure \ref{fig:fig1}).
+Here, we opted for omitting the equations to convert a point
+$(x_{i}, y_{i}, z_{i})$ into a point $(h_{i}, \varphi_{i}, \lambda_{i})$
+and vice versa. These equations, however, may be easily
+found in the literature \citep[e.g.,][]{heiskanen-moritz1967, soler1976, torge2012, bouman_etal2013}.
+
 By definition, the gravity disturbance $\delta g_{i}$,
 at the point $(x_{i}, y_{i}, z_{i})$, is given by 
-\citep{hofmann-wellenhof-moritz2005}:
+\citep{heiskanen-moritz1967, hofmann-wellenhof-moritz2005}:
 \begin{equation}
@@
 \end{equation}
-where $\hat{\mitbf{\gamma}}_{i} = -\hat{\mitbf{u}}_{i}$ is a unit 
+where $^{\top}$ denotes transposition,
+$\hat{\mitbf{\gamma}}_{i} = -\hat{\mitbf{u}}_{i}$ is a unit 
 vector with the same direction as the normal gravity vector 
-$\mitbf{\gamma}_{i}$ at the point $(x_{i}, y_{i}, z_{i})$ and
+$\mitbf{\gamma}_{i}$ at the point $(x_{i}, y_{i}, z_{i})$, in the
+topocentric Cartesian coordinate system (Figure \ref{fig:fig2}b), and
 $\hat{\mitbf{u}}_{i}$ is the unit vector $\hat{\mitbf{u}}$ 
-(equation \ref{eq:unit-vectors}) evaluated at the point $(h_{i}, \varphi_{i}, \lambda_{i})$, which is obtained by transforming $(x_{i}, y_{i}, z_{i})$
-from the topocentric coordinate system (Figure \ref{fig:fig2}) to the
+(equation \ref{eq:unit-vectors}) evaluated at the corresponding 
+point $(h_{i}, \varphi_{i}, \lambda_{i})$ in the 
 geodetic coordinate system (Figure \ref{fig:fig1}).
-
-PAREI AQUI
 
@@
 total-field anomalies \citep[e.g.,][]{blakely1996}.
-Notice that, local- or regional-gravity studies, the unit
+Notice that, in local- or regional-gravity studies, the unit
 vector $\hat{\mitbf{\gamma}}_{i}$ (equation \ref{eq:gobs-approx}) 
-coincides with the $z$ axis
-of the local Cartesian coordinate system defined
-at the beginning of this section. Consequently,
-by using the approximation defined in equation \ref{eq:gobs-approx}, 
-the gravity disturbance (equation \ref{eq:gravity-disturbance}) can be 
-rewritten as follows \begin{equation}
-\delta g^{o}_{i} \approx \hat{\mitbf{z}}^{\top} \Delta \mitbf{g}^{o}_{i} \: ,
+may be considered constant throughout the study area and 
+parallel to the $z$ axis of the topocentric Cartesian coordinate 
+system (Figure \ref{fig:fig2}b).
+Consequently, by using the approximation defined in equation 
+\ref{eq:gobs-approx}, the gravity disturbance 
+(equation \ref{eq:gravity-disturbance}) can be rewritten as 
+follows 
+\begin{equation}
+\delta g_{i} \approx \hat{\mitbf{\gamma}}_{P}^{\top} \Delta \mitbf{g}_{i} \: ,
 \label{eq:gravity-disturbance-approx}
 \end{equation}
-where $\hat{\mitbf{z}}^{\top} = \left[ \, 0 \; 0 \; 1 \right]$.
-According to equation \ref{eq:gravity-disturbance-approx},
-the gravity disturbance $\delta g^{o}_{i}$ (equation \ref{eq:gravity-disturbance})
-represents the vertical component of the gravitational attraction exerted by
+where $\hat{\mitbf{\gamma}}_{P}$ represents the unit vector with
+the same direction as the normal gravity vector $\mitbf{\gamma}_{P}$
+at the origin $P$ of the topocentric Cartesian coordinate system
+(Figure \ref{fig:fig2}).
+This equation shows that the gravity disturbance $\delta g_{i}$ (equation \ref{eq:gravity-disturbance}) is different from the magnitude
+of the disturbing gravitational attraction $\Delta \mitbf{g}_{i}$
+produced by the gravity sources. Rather, it represents the 
+component of the $\Delta \mitbf{g}_{i}$ on the direction of the normal
+gravity vector. In the topocentric Cartesian coordinate system,
+the gravity disturbance $\delta g_{i}$ (equation \ref{eq:gravity-disturbance-approx})
+can be defined as the vertical component of the gravitational attraction exerted by
 the gravity sources at the point $(x_{i}, y_{i}, z_{i})$.
 As a consequence, the gravity disturbance produced by
-a homogeneous gravity source can be represented 
-by the following harmonic function
+a homogeneous gravity source with density contrast $\Delta\rho$
+(in $kg / m^{3}$) can be represented by the following harmonic function:
 \begin{equation}
-d^{o}_{i} = k_{g} \, G \, \rho \, \partial_{z} \phi_{i} \: ,
+d_{i} = k_{g} \, G \, \Delta\rho \, \partial_{z} \phi_{i} \: ,
 \label{eq:gz-local}
@@
 
-\section{Gravity disturbance versus gravity anomaly}
+Practically all the literature about gravity modeling
+use the quantity $d_{i}$ (equation \ref{eq:gz-local})
+to represent the gravity anomaly produced by 
+gravity sources \citep[e.g.,][]{blakely1996}. 
+Consequently, almost all the geophysicists use the gravity
+anomaly as an approximation of the gravity disturbance
+produced by the gravity sources. Notice that $d_{i}$ 
+(equation \ref{eq:gz-local}) does not depend on the
+geoidal surface. Rather, it depends on the relative position 
+of the observation points $(x_{i}, y_{i}, z_{i})$ with respect to
+the gravity sources in a topocentric Cartesian coordinate
+system (Figure \ref{fig:fig2}b).
 
-Almost all interpretation techniques assume, implicitly or
+\section{Conclusions}
+
+We debate the conceptual differences between the gravity
+disturbance and the gravity anomaly.
+Our reasoning suggest that the gravity disturbance is the
+more appropriated quantity for representing the gravity
+effect produced by the gravity sources.
+In summary, we point out that:
+
+\begin{enumerate}
+%\renewcommand{\theenumi}{(\arabic{enumi})}
+\renewcommand{\theenumii}{\alph{enumii}}
+
+\item Almost all interpretation techniques assume, implicitly or
 directly, that the gravity data is harmonic (e.g., 
@@
 computation of vertical derivatives via Fourier and Hilbert
-transforms).
-
-Consequently, they implicitly or directly assume that
+transforms). As a consequence, they implicitly or directly assume that
 the gravity data approximates the gravitational disturbance.
 
-Almost all forward modelling techniques compute
+\item Almost all forward modeling techniques compute
 the vertical component of the gravitational attraction 
 exerted by the geological bodies at the observation points.
+Notice that the gravity anomaly requires the computation of gravity
+on the Geoid, which is generally within the topographic masses.
+Hence, almost all geophysicists implicitly compute 
+the gravity disturbance.
 
-Hence, almost all geophysicists implicitly compute 
-the vertical component of the gravitational disturbance.
-
-The gravity anomaly is defined as the difference between 
-the gravity $\| \mitbf{g}_{P} \|$, at a point $P$ on the
-Geoid, and the normal gravity $\| \mitbf{\gamma}_{Q} \|$,
-on the reference elipsoid, where $P$ and $Q$ have the same
-geodetic latitude and longitude.
-
+\item The gravity anomaly is defined as the difference between 
+the gravity on the Geoid and the normal gravity on the reference 
+ellipsoid, both at the same geodetic latitude and longitude.
 Consequently, the gravity anomaly is a function of
 the geodetic latitude and longitude only and cannot
-be calculated at arbitrary heights.
+be calculated at arbitrary heights. On the other hand,
+the gravity disturbance can be computed at arbitrary points
+outside the sources.
 
-Gravity anomalies require the computation of gravity
-within the topographic masses, where the gravitational
-disturbance is not harmonic.
+\end{enumerate}
+
+The principal theoretical implication of this study is that
+is that, although the gravity anomaly may be used as a good approximation of
+the gravity effect produced by the sources for most practical applications,
+the more appropriated quantity for gravity modeling is the
+gravity disturbance. We stress that, more important than the 
+the terminology changes, the geophysicist must bear in mind the 
+conceptual assumptions used in gravity modeling.
 

### 2018-04-17 manuscript/letter.tex
coordinate systems

@@
 
-\title[Gravity disturbance or gravity anomaly?]
-      {Gravity disturbance or gravity anomaly?}
+\title[Theoretical aspects in gravity modeling]
+      {Theoretical aspects in gravity modeling}
 
@@
 on a body at rest on the Earth's surface is called 
-\textit{gravity vector} and its intensity is called simply 
+\textit{gravity vector} and its intensity is called 
 \textit{gravity} \citep{heiskanen-moritz1967,
@@
 density distribution.
-%In applied geophysics, gravity surveys are usually conducted 
-%over local or regional areas on the Earth's surface for the purpose of 
-%characterizing geological structures located within the
-%crust and upper mantle.
-%
-%Consequently, geophysicists are only interested in the
-%particular gravitational component of gravity which 
-%is produced by these target geological structures.
+
 The first step of the procedure for isolating this 
@@
 distributions related to geological structures in subsurface 
-are the main goals in applied geophysics 
-\citep{blakely1996}.
+are the main goals in applied geophysics \citep{blakely1996}.
+The computation of the gravitational effect produced by
+the geological structures is commonly known in geophysics as 
+\textit{gravity modeling}.
+This term is different from \textit{gravity field modeling}, which
+is commonly used in geodesy to denote the characterization of the 
+gravity field in a local, regional or global scale.
 
@@
 Our reasoning suggests that the gravity 
-disturbance is more appropriated than gravity anomalies for 
-approximating the gravitational effect produced by the Earth's 
-internal density distribution.
+disturbance is more appropriated than the gravity anomaly 
+for geophysical purposes.
 
@@
 Traditionally, the Earth's gravity field is approximated 
-by the gravity field produced by a geocentric and rigid ellipsoid 
-of revolution, which has the minor axis $b$ 
+by the gravity field produced by a reference ellipsoid
+(or level ellipsoid), which is rigid and geocentric.
+This reference ellipsoid has the minor axis $b$ 
 coincident with the mean rotating axis of the Earth $Z$, the 
@@
 \textit{normal gravity vector} and its intensity is called 
-simply \textit{normal gravity}.
+\textit{normal gravity}.
 In geodesy, any model used to represent the normal gravity field
@@
 
-The normal Earth gives rise to the \textit{geodetic coordinate system}
-\citep{heiskanen-moritz1967, soler1976, torge2012, bouman_etal2013}.
-In this coordinate system, the position of a point $P$
+\section{Terrestrial Reference Systems used in gravity modeling}
+
+For geophysical purposes, there are three important Terrestrial Reference Systems used in gravity modeling. 
+They rotate with the Earth and are used for describing
+positions and movements of objects on and close to the Earth’s surface
+\citep{torge2012}.
+
+The first one is a geocentric system of Cartesian coordinates having 
+the $Z$-axis coincident with the mean Earth's rotational axis,
+the $X$-axis pointing to the Greenwich meridian and the $Y$-axis
+directed so as to obtain a right-handed system (Figure \ref{fig:fig1}).
+This reference system can be found in the literature with different names: Mean Terrestrial System \citep[e.g.,][]{soler1976}, Earth-fixed geocentric Cartesian system \citep[e.g.,][]{torge2012} 
+or Earth-centered Earth-fixed system \citep[e.g.,][]{bouman_etal2013}, for example. Here, we opted for simply using the term Geocentric Cartesian System (GCS).
+
+Other important reference system is a geocentric system of 
+geodetic coordinates, which is defined by the reference ellipsoid 
+used in the Normal Earth model \citep{heiskanen-moritz1967, soler1976, 
+torge2012, bouman_etal2013}. 
+In this coordinate system, the position of a point 
 is defined by the \textit{geometric height} $h$, 
-\textit{geodetic latitude}
-$\varphi$ and \textit{longitude} $\lambda$ (Figure \ref{fig:fig1}).
-Geodetic coordinates $(h, \varphi, \lambda)$ can be easily 
-converted into geocentric Cartesian coordinates $(X, Y, Z)$
-(Figure \ref{fig:fig1}).
-The plane containing the point $P$, the axis $Z$ and
-the origin $O$ of this \textit{geocentric Cartesian coordinate system}
-is called meridian plane (gray plane in Figure \ref{fig:fig1}).
+\textit{geodetic latitude} $\varphi$ and \textit{longitude} $\lambda$ (Figure \ref{fig:fig1}).
+For convenience, we call this system Geocentric Geodetic System
+(GGS). 
 At a given point $(h, \varphi, \lambda)$, there are three 
-unit vectors (Figure \ref{fig:fig1}) given by \citep{soler1976}:
+mutually-orthogonal unit vectors (Figure \ref{fig:fig1}) given by \citep{soler1976}:
 \begin{equation}
@@
 \end{bmatrix}
-\end{split} \: ,
+\end{split} \: .
 \label{eq:unit-vectors}
 \end{equation}
-Notice that these unit vectors are mutually orthogonal and
-that the normal gravity vector $\mitbf{\gamma}_{P}$ at a point $P$
-is opposite to the unit vector $\hat{\mitbf{u}}_{P}$ at $P$.
+The required equations to convert coordinates $(h, \varphi, 
+\lambda)$ referred to the GGS into coordinates $(X, Y, Z)$ referred to 
+the GCS (Figure \ref{fig:fig1}) and vice versa can be easily found in 
+the literature \citep[e.g.,][]{heiskanen-moritz1967, torge2012, 
+bouman_etal2013}.
 
-\section{Gravity disturbance}
+PAREI AQUI
+
+In a local or regional study, geophysicists commonly use a topocentric 
+Cartesian coordinate system (TCS) with origin at a point $P$ on or 
+close to the Earth's surface and axes $x$, $y$ and $z$ (Figure 
+\ref{fig:fig2}). In the TCS, the axes $x$ and $y$ are parallel to the unit vectors $\hat{\mitbf{v}}_{P}$ $\hat{\mitbf{w}}_{P}$, respectively. 
+The $z$-axis, however, is opposite to the unit vector $\hat{\mitbf{u}}_{P}$ (equation \ref{eq:unit-vectors}) and points downward.
+
+\section{Basic concepts}
 
@@
 
-In a local- or regional-gravity study, 
-geophysicists commonly use a topocentric Cartesian coordinate system
-with origin at a point $P$ and axes $x$, $y$ and $z$ defined
-by the unit vectors $\hat{\mitbf{v}}_{P}$,
-$\hat{\mitbf{w}}_{P}$ and $-\hat{\mitbf{u}}_{P}$ 
-(equation \ref{eq:unit-vectors}), respectively
-(Figures \ref{fig:fig1} and \ref{fig:fig2}).
 In this coordinate system, the observed gravity vector
@@
 in the geodetic coordinate system (Figure \ref{fig:fig1}).
-Here, we opted for omitting the equations to convert a point
-$(x_{i}, y_{i}, z_{i})$ into a point $(h_{i}, \varphi_{i}, \lambda_{i})$
-and vice versa. These equations, however, may be easily
-found in the literature \citep[e.g.,][]{heiskanen-moritz1967, soler1976, torge2012, bouman_etal2013}.
 
@@
     \includegraphics{figures/cartesian-geodetic-systems.png}
-    \caption{Schematic representation of the geodetic coordinate
-    system defined by an oblate ellipsoid with semi-minor axis $b$, 
-    coincident with the mean Earth's rotation axis, and a semi-major
+    \caption{Schematic representation of the Geocentric Cartesian System (GCS) and the Geocentric Geodetic System (GGS). The GCS has the $Z$-axis coincident with the mean Earth's rotational axis,
+    the $X$-axis pointing to the Greenwich meridian and the $Y$-axis
+    directed so as to obtain a right-handed system. The GGS is      defined by an oblate ellipsoid with semi-minor axis $b$, 
+    coincident with the $Z$-axis of GCS, and a semi-major
     axis $a$. In this coordinate system, the position of a point is
@@
     The Earth's center of mass is represented 
-    by $O$, $P$ represents a point $(h, \varphi, \lambda)$ and 
-    $P^{\prime}$ represents the projection of $P$ on the plane $XY$ 
+    by $O$, $P$ represents a point $(h_{P}, \varphi_{P}, \lambda_{P})$ and $P^{\prime}$ its projection onto the plane $XY$ 
     (Equatorial plane). The plane containing $O$, $P$ and 
-    $P^{\prime}$ is represented by the gray triangle in (a) and (b). 
-    The unit vectors 
-    $\hat{\mitbf{u}}_{P}$, $\hat{\mitbf{v}}_{P}$ and 
+    $P^{\prime}$ is represented in gray in (a) and (b). 
+    The unit vectors $\hat{\mitbf{u}}_{P}$, $\hat{\mitbf{v}}_{P}$ and 
     $\hat{\mitbf{w}}_{P}$ define three mutually orthogonal 
-    directions at each point $P$.
-    In (b), the point $Q$ represents the projection of $P$ on the 
-    ellipsoid surface, at the same latitude $\varphi$ and 
-    longitude $\lambda$.}
+    directions at $P$ (equation \ref{eq:unit-vectors}).
+    In (b), $Q$ represents a point $(h_{Q}, \varphi_{Q}, \lambda_{Q})$,
+    which is the projection of $P$ onto the reference ellipsoid, at
+    the same latitude and longitude ($\varphi_{Q} = \varphi_{P}$ and $\lambda_{Q} = \lambda_{P}$).}
   \label{fig:fig1}
@@
     disturbing gravitational attraction $\Delta \mitbf{g}_{P}$
-    (equation \ref{eq:gravity-vector}) and the unit vector 
-    $\hat{\mitbf{u}}_{P}$ (equation \ref{eq:unit-vectors}) 
-    at the point $P$ and also the normal gravity vector
-    $\mitbf{\gamma}_{Q}$ at the point $Q$ on the reference ellipsoid.
-    At the point $P$ the normal gravity vector $\mitbf{\gamma}_{P}$ 
-    is opposite to the unit vector $\hat{\mitbf{u}}_{P}$ and the 
-    normal gravity $\mitbf{\gamma}_{Q}$ at the point $Q$ is opposite
-    the another unit vector $\hat{\mitbf{u}}_{Q}$, which is
-    close to $\hat{\mitbf{u}}_{P}$.
+    (equation \ref{eq:gravity-vector}), unit vector 
+    $\hat{\mitbf{u}}_{P}$ (equation \ref{eq:unit-vectors}) and
+    the normal gravity vector $\mitbf{\gamma}_{Q}$ at the point $Q$ on the reference ellipsoid.
+    At $P$, the normal gravity vector $\mitbf{\gamma}_{P}$ 
+    is opposite to the unit vector $\hat{\mitbf{u}}_{P}$ and, at $Q$, the normal gravity $\mitbf{\gamma}_{Q}$ is opposite
+    to the unit vector $\hat{\mitbf{u}}_{Q}$. These unit vectors are
+    close to each other.
     (b) Schematic representation of a topocentric Cartesian 
-    coordinate system with origin at a point $P$ and $x$, $y$ and $z$ 
-    axes defined by the unit vectors $\hat{\mitbf{v}}_{P}$, 
-    $\hat{\mitbf{w}}_{P}$ and $-\hat{\mitbf{u}}_{P}$ 
-    (Figure \ref{fig:fig1}), respectively.}
+    coordinate system with origin at a point $P$. The axes $x$
+    and $y$ are parallel to the unit vectors $\hat{\mitbf{v}}_{P}$
+    and $\hat{\mitbf{w}}_{P}$ (equation \ref{eq:unit-vectors} and
+    Figure \ref{fig:fig1}), respectively. The $z$ axis is opposite
+    to the unit vector $\hat{\mitbf{u}}_{P}$ (equation 
+    \ref{eq:unit-vectors} and Figure \ref{fig:fig1}) and points downward.}
   \label{fig:fig2}

### 2018-04-18 manuscript/letter.tex
eq relating GGC and TCS coordinates and fig 3

@@
 the $X$-axis pointing to the Greenwich meridian and the $Y$-axis
-directed so as to obtain a right-handed system (Figure \ref{fig:fig1}).
+directed so as to obtain a right-handed system (Figure \ref{fig:GCS-GGS}).
 This reference system can be found in the literature with different names: Mean Terrestrial System \citep[e.g.,][]{soler1976}, Earth-fixed geocentric Cartesian system \citep[e.g.,][]{torge2012} 
@@
 is defined by the \textit{geometric height} $h$, 
-\textit{geodetic latitude} $\varphi$ and \textit{longitude} $\lambda$ (Figure \ref{fig:fig1}).
+\textit{geodetic latitude} $\varphi$ and \textit{longitude} $\lambda$ (Figure \ref{fig:GCS-GGS}).
 For convenience, we call this system Geocentric Geodetic System
@@
 At a given point $(h, \varphi, \lambda)$, there are three 
-mutually-orthogonal unit vectors (Figure \ref{fig:fig1}) given by \citep{soler1976}:
+mutually-orthogonal unit vectors (Figure \ref{fig:GCS-GGS}) given by \citep{soler1976}:
 \begin{equation}
@@
 \lambda)$ referred to the GGS into coordinates $(X, Y, Z)$ referred to 
-the GCS (Figure \ref{fig:fig1}) and vice versa can be easily found in 
+the GCS (Figure \ref{fig:GCS-GGS}) and vice versa can be easily found in 
 the literature \citep[e.g.,][]{heiskanen-moritz1967, torge2012, 
 bouman_etal2013}.
-
-PAREI AQUI
 
@@
 close to the Earth's surface and axes $x$, $y$ and $z$ (Figure 
-\ref{fig:fig2}). In the TCS, the axes $x$ and $y$ are parallel to the unit vectors $\hat{\mitbf{v}}_{P}$ $\hat{\mitbf{w}}_{P}$, respectively. 
-The $z$-axis, however, is opposite to the unit vector $\hat{\mitbf{u}}_{P}$ (equation \ref{eq:unit-vectors}) and points downward.
+\ref{fig:TCS}). In the TCS, the axes $x$ and $y$ are parallel to 
+the unit vectors $\hat{\mitbf{v}}_{P}$ $\hat{\mitbf{w}}_{P}$, respectively,
+whereas the $z$-axis is opposite to the unit vector 
+$\hat{\mitbf{u}}_{P}$ (equation \ref{eq:unit-vectors}) and points downward.
+
+Consider a TCS (Figure \ref{fig:TCS}) with origin at a point $P$ with
+coordinates $(X_{P}, Y_{P}, Z_{P})$ referred to the GCS (Figure \ref{fig:fig1}).
+In this case, the relationship between the coordinates $(X, Y, Z)$ of a 
+point in the GCS and the coordinates $(x, y, z)$ of the same point in the 
+TCS is given by:
+\begin{equation}
+\begin{bmatrix}
+x \\
+y \\
+z 
+\end{bmatrix} =
+\mitbf{R}^{\top} \begin{bmatrix}
+X - X_{P} \\
+Y - Y_{P} \\
+Z - Z_{P}
+\end{bmatrix} \: ,
+\label{eq:GCS2TCS}
+\end{equation}
+where $\mitbf{R}$ is an orthogonal matrix whose columns
+are defined by the unit vectors $\hat{\mitbf{u}}$, $\hat{\mitbf{v}}$ and
+$\hat{\mitbf{w}}$ (equation \ref{eq:unit-vectors}) as follows:
+\begin{equation}
+\mitbf{R} = 
+\begin{bmatrix}
+\hat{\mitbf{v}} & \hat{\mitbf{w}} & -\hat{\mitbf{u}}
+\end{bmatrix} \: .
+\label{eq:R}
+\end{equation}
 
@@
 For each point $(x_{i}, y_{i}, z_{i})$ in the topocentric
-Cartesian coordinate system (Figure \ref{fig:fig2}b), there is
+Cartesian coordinate system (Figure \ref{fig:TCS}b), there is
 a corresponding point $(h_{i}, \varphi_{i}, \lambda_{i})$
-in the geodetic coordinate system (Figure \ref{fig:fig1}).
+in the geodetic coordinate system (Figure \ref{fig:GCS-GGS}).
 
@@
 $\mitbf{\gamma}_{i}$ at the point $(x_{i}, y_{i}, z_{i})$, in the
-topocentric Cartesian coordinate system (Figure \ref{fig:fig2}b), and
+topocentric Cartesian coordinate system (Figure \ref{fig:TCS}b), and
 $\hat{\mitbf{u}}_{i}$ is the unit vector $\hat{\mitbf{u}}$ 
@@
 point $(h_{i}, \varphi_{i}, \lambda_{i})$ in the 
-geodetic coordinate system (Figure \ref{fig:fig1}).
+geodetic coordinate system (Figure \ref{fig:GCS-GGS}).
 
@@
 parallel to the $z$ axis of the topocentric Cartesian coordinate 
-system (Figure \ref{fig:fig2}b).
+system (Figure \ref{fig:TCS}b).
 Consequently, by using the approximation defined in equation 
@@
 at the origin $P$ of the topocentric Cartesian coordinate system
-(Figure \ref{fig:fig2}).
+(Figure \ref{fig:TCS}).
 This equation shows that the gravity disturbance $\delta g_{i}$ (equation \ref{eq:gravity-disturbance}) is different from the magnitude
@@
 the gravity sources in a topocentric Cartesian coordinate
-system (Figure \ref{fig:fig2}b).
+system (Figure \ref{fig:TCS}b).
 
@@
     the same latitude and longitude ($\varphi_{Q} = \varphi_{P}$ and $\lambda_{Q} = \lambda_{P}$).}
-  \label{fig:fig1}
+  \label{fig:GCS-GGS}
+\end{figure}
+
+\begin{figure}
+    \includegraphics{figures/surfaces.png}
+    \caption{text text text text}
+  \label{fig:surfaces}
 \end{figure}
@@
     and $\hat{\mitbf{w}}_{P}$ (equation \ref{eq:unit-vectors} and
-    Figure \ref{fig:fig1}), respectively. The $z$ axis is opposite
+    Figure \ref{fig:GCS-GGS}), respectively. The $z$ axis is opposite
     to the unit vector $\hat{\mitbf{u}}_{P}$ (equation 
-    \ref{eq:unit-vectors} and Figure \ref{fig:fig1}) and points downward.}
-  \label{fig:fig2}
+    \ref{eq:unit-vectors} and Figure \ref{fig:GCS-GGS}) and points downward.}
+  \label{fig:TCS}
 \end{figure}

### 2018-04-20 manuscript/letter.tex
start sec anomalous quantities in gravity modeling

@@
 Consider a TCS (Figure \ref{fig:TCS}) with origin at a point $P$ with
-coordinates $(X_{P}, Y_{P}, Z_{P})$ referred to the GCS (Figure \ref{fig:fig1}).
+coordinates $(X_{P}, Y_{P}, Z_{P})$ referred to the GCS (Figure \ref{fig:GCS-GGS}).
 In this case, the relationship between the coordinates $(X, Y, Z)$ of a 
@@
 
-\section{Basic concepts}
+\section{Anomalous quantities in gravity modeling}
 
-It is worth noting that, by definition, 
-the centrifugal component of the normal gravity field is
-equal to the centrifugal component of the Earth's gravity
-field if they are evaluated at the same point.
-Then, the differences between the gravity vector
-(corrected from non-gravitational effects) 
-and the normal gravity vector, at the same point, represents a purely 
-gravitational and consequently harmonic disturbing field.
-This gravitational disturbance 
-is caused by contrasts between the actual internal 
-density distribution of the Earth and the internal density 
+Let $\mitbf{g}_{P}$ and $\mitbf{\gamma}_{P}$ be, respectively, the 
+gravity vector (corrected from non-gravitational effects due to vehicle 
+motion and time variations such as Earth tides, instrumental drift and 
+barometric pressure changes, for example) and the normal gravity vector 
+at a point $P$. The difference between these two vectors, at the same 
+point $P$, defines a quantity called \textit{gravity disturbance 
+vector}, which is given by:
+\begin{equation}
+\delta \mitbf{g}_{P} = 
+\mitbf{g}_{P} - \mitbf{\gamma}_{P} \: .
+\label{eq:gravity-disturbance-vector}
+\end{equation}
+By definition, the centrifugal component of the normal gravity vector 
+is equal to the centrifugal component of the gravity vector if they 
+are evaluated at the same point.
+Then, the gravity disturbance vector $\delta \mitbf{g}_{P}$ represents 
+a purely gravitational (and consequently harmonic) quantity, which is 
+caused by contrasts between the actual internal 
+density distribution of the Earth and the (unknown) internal density 
 distribution of the normal Earth.
@@
 \citeyear{hammer1945}; \citeauthor{lafehr1965}, \citeyear{lafehr1965}),
-\textit{density anomalies} (e.g., \citeauthor{forsberg1984}, \citeyear{forsberg1984})
-or \textit{gravity sources} (e.g., \citeauthor{blakely1996}, 
-\citeyear{blakely1996}). Here, we opted for using the last term.
+\textit{density anomalies} (e.g., \citeauthor{forsberg1984}, 
+\citeyear{forsberg1984}) or \textit{gravity sources} (e.g., 
+\citeauthor{blakely1996}, \citeyear{blakely1996}). Here, we opted for 
+using the last term.
+Figure \ref{fig:surfaces} illustrates the gravity vector 
+$\mitbf{g}_{P}$, normal gravity vector $\mitbf{\gamma}_{P}$ and 
+the gravity disturbance vector $\delta \mitbf{g}_{P}$ at a point $P$ 
+located on the Earth's surface.
 
-The difference between the observed gravity and the
-normal gravity, at the same point, is called \textit{gravity disturbance}
-\citep{heiskanen-moritz1967, hofmann-wellenhof-moritz2005}.
-Notice that the gravity disturbance is not equivalent to the
-magnitude of the difference between the gravity vector
-and the normal gravity vector at the same point 
-\citep{barthelmes2013, sanso_sideris2013}.
+The difference between the gravity $g_{P}$ and the normal gravity 
+$\gamma_{P}$, at the same point $P$, is called \textit{gravity 
+disturbance} \citep{heiskanen-moritz1967, hofmann-wellenhof-moritz2005} 
+and can be represented as follows:
+\begin{equation}
+\delta g_{P} = g_{P} - \gamma_{P} \: .
+\label{eq:gravity-disturbance}
+\end{equation}
+Notice that the gravity disturbance $\delta g_{P}$ is not equivalent 
+to the magnitude of the gravity disturbance vector 
+$\delta \mitbf{g}_{P}$ \citep{barthelmes2013, sanso_sideris2013}.
 As properly pointed out by \citet{hackney-featherstone2003},
@@
 but appears to be less well known in geophysics.
+
+%Consider a line (dashed line in Figure \ref{fig:surfaces}) passing 
+%through a point $Q$ on the surface of the reference ellipsoid and 
+%a point $P$ on the Earth's surface
+
+PAREI AQUI - precisa descrever as demais quantidade na fig 3
 
@@
     axis $a$. In this coordinate system, the position of a point is
-    determined by the geometric height $h$, the geodetic latitude 
+    determined by the geometric height $h$, geodetic latitude 
     $\varphi$ and longitude $\lambda$. 
@@
 \begin{figure}
-    \includegraphics{figures/surfaces.png}
-    \caption{text text text text}
-  \label{fig:surfaces}
-\end{figure}
-
-\begin{figure}
     \includegraphics{figures/local-system.png}
@@
     $\mitbf{g}_{P}$, normal gravity vector $\mitbf{\gamma}_{P}$,
-    disturbing gravitational attraction $\Delta \mitbf{g}_{P}$
-    (equation \ref{eq:gravity-vector}), unit vector 
-    $\hat{\mitbf{u}}_{P}$ (equation \ref{eq:unit-vectors}) and
-    the normal gravity vector $\mitbf{\gamma}_{Q}$ at the point $Q$ on the reference ellipsoid.
-    At $P$, the normal gravity vector $\mitbf{\gamma}_{P}$ 
-    is opposite to the unit vector $\hat{\mitbf{u}}_{P}$ and, at $Q$, the normal gravity $\mitbf{\gamma}_{Q}$ is opposite
-    to the unit vector $\hat{\mitbf{u}}_{Q}$. These unit vectors are
-    close to each other.
+    gravity disturbance vector $\delta \mitbf{g}_{P}$
+    (equation \ref{eq:gravity-vector}) and unit vector 
+    $\hat{\mitbf{u}}_{P}$ (equation \ref{eq:unit-vectors}) at a point 
+    $P$.
     (b) Schematic representation of a topocentric Cartesian 
@@
     and $\hat{\mitbf{w}}_{P}$ (equation \ref{eq:unit-vectors} and
-    Figure \ref{fig:GCS-GGS}), respectively. The $z$ axis is opposite
-    to the unit vector $\hat{\mitbf{u}}_{P}$ (equation 
-    \ref{eq:unit-vectors} and Figure \ref{fig:GCS-GGS}) and points downward.}
+    Figure \ref{fig:GCS-GGS}), respectively. On the other hand, the $z$ axis is opposite to the unit vector $\hat{\mitbf{u}}_{P}$ (equation 
+    \ref{eq:unit-vectors} and Figure \ref{fig:GCS-GGS}) and points downward. The gray plane shown in (b) is the same gray plane
+    shown in Figure \ref{fig:GCS-GGS}.}
   \label{fig:TCS}
+\end{figure}
+
+\begin{figure}
+    \includegraphics{figures/surfaces.png}
+    \caption{Schematic representation of the gravity vector
+        $\mitbf{g}_{P}$, normal gravity vector $\mitbf{\gamma}_{P}$,
+        gravity disturbance vector $\delta \mitbf{g}_{P}$
+        (equation \ref{eq:gravity-vector}) and unit vector 
+        $\hat{\mitbf{u}}_{P}$ (equation \ref{eq:unit-vectors}) at a point $P$ on the surface of the Earth; gravity vector 
+        $\mitbf{g}_{Q^{\prime}}$ at a point $Q^{\prime}$ on the Geoid;
+        normal gravity vector $\mitbf{\gamma}_{Q}$ at a point $Q$ on
+        the reference ellipsoid. The dashed line is normal to the 
+        surface of the reference ellipsoid at $Q$.}
+  \label{fig:surfaces}
 \end{figure}

### 2018-04-22 manuscript/letter.tex
finish the anomalous quantities section

@@
 barometric pressure changes, for example) and the normal gravity vector 
-at a point $P$. The difference between these two vectors, at the same 
+at a point $P$. 
+In this case, the gravity vector $\mitbf{g}_{P}$ represents the 
+gradient of a scalar potential called \textit{gravity potential}, 
+which is the sum of a scalar gravitational potential and a 
+scalar centrifugal potential.
+Similarly, the normal gravity vector $\mitbf{\gamma}_{P}$ represents the 
+gradient of a scalar potential called \textit{normal potential}, 
+which is also the sum of a scalar gravitational potential and a 
+scalar centrifugal potential.
+By definition, the centrifugal part of the normal potential is equal 
+to that of the gravity potential.
+Then, the both gravity vector and normal gravity vectors are represented 
+by the sum of two vectors: the gradient of a gravitational potential and
+the gradient of a centrifugal potential. Notice that, as the centrifugal 
+part of the gravity potential and normal potential are the same, the
+centrifugal part of the gravity vector and normal gravity vector are
+also equal to each other.
+
+The difference between $\mitbf{g}_{P}$ and $\mitbf{\gamma}_{P}$, at the same 
 point $P$, defines a quantity called \textit{gravity disturbance 
@@
 \end{equation}
-By definition, the centrifugal component of the normal gravity vector 
-is equal to the centrifugal component of the gravity vector if they 
-are evaluated at the same point.
-Then, the gravity disturbance vector $\delta \mitbf{g}_{P}$ represents 
+As the centrifugal part of the gravity vector and the normal gravity vector 
+is equal to the centrifugal component of the gravity vector, 
+the gravity disturbance vector $\delta \mitbf{g}_{P}$ represents 
 a purely gravitational (and consequently harmonic) quantity, which is 
@@
 
-The difference between the gravity $g_{P}$ and the normal gravity 
-$\gamma_{P}$, at the same point $P$, is called \textit{gravity 
-disturbance} \citep{heiskanen-moritz1967, hofmann-wellenhof-moritz2005} 
+The difference between gravity $g_{P}$ (magnitude of the gravity 
+vector $\mitbf{g}_{P}$) and normal gravity $\gamma_{P}$ 
+(magnitude of the normal gravity vector $\mitbf{\gamma}_{P}$), at the 
+same point $P$, is called \textit{gravity disturbance} 
+\citep{heiskanen-moritz1967, hofmann-wellenhof-moritz2005} 
 and can be represented as follows:
@@
 
-PAREI AQUI - precisa descrever as demais quantidade na fig 3
+The gravity anomaly is the commonly used quantity in applied 
+geophysics. It is defined as the difference between the 
+gravity at a point $Q^{\prime}$ on the geoid and the normal gravity 
+at a point $Q$ on the ellipsoid surface,
+both at the same geodetic latitude and longitude (Figure \ref{fig:surfaces}).
+Notice that, by definition, the gravity anomaly depends on 
+longitude and latitude only and is not a function of height 
+\citep{barthelmes2013}. As a consequence, it is not possible to compute,
+for example, the upward of a gravity anomaly.
 
-The gravity anomaly is the commonly used quantity in applied 
-geophysics. It is defined as the difference
-between the gravity on the geoid and the normal gravity on the ellipsoid,
-both at the same geodetic latitude and longitude.
-Notice that, by definition, the gravity anomaly depends on 
-longitude and latitude only and is not a function in space 
-\citep{barthelmes2013}.
-Different gravity anomalies can be calculated, depending on the
-corrections applied to them \citep{blakely1996, hofmann-wellenhof-moritz2005}.
+Different gravity anomalies can be calculated (e.g., Free-air and 
+Bouguer anomalies), depending on the corrections applied to 
+them \citep{blakely1996, hofmann-wellenhof-moritz2005}.
 These corrections are usually called gravity reductions.
-For example, the Free-air anomaly is an approximation of the
+It can be shown that the Free-air anomaly is an approximation of the
 gravity disturbance whereas the Bouguer anomaly
 is an approximation of the terrain corrected gravity disturbance.
-The last one is commonly used by geophysicists as the
+The Bouguer anomaly is commonly used by geophysicists as the
 gravitational effect produced by the gravity sources.
@@
 component of the $\Delta \mitbf{g}_{i}$ on the direction of the normal
-gravity vector. In the topocentric Cartesian coordinate system,
+gravity vector \citep{sanso_sideris2013}. 
+In the topocentric Cartesian coordinate system,
 the gravity disturbance $\delta g_{i}$ (equation \ref{eq:gravity-disturbance-approx})

### 2018-04-22 manuscript/letter.tex

@@
 $\hat{\mitbf{u}}_{P}$ (equation \ref{eq:unit-vectors}) and points downward.
-
-Consider a TCS (Figure \ref{fig:TCS}) with origin at a point $P$ with
+Consider, for example, a TCS (Figure \ref{fig:TCS}) with origin at a point $P$ with
 coordinates $(X_{P}, Y_{P}, Z_{P})$ referred to the GCS (Figure \ref{fig:GCS-GGS}).
@@
 geophysics. It is defined as the difference between the 
-gravity at a point $Q^{\prime}$ on the geoid and the normal gravity 
+gravity at a point $Q^{\prime}$ on the geoid (a particular 
+equipotential surface of the gravity potential) and the normal gravity 
 at a point $Q$ on the ellipsoid surface,
@@
 \citep{barthelmes2013}. As a consequence, it is not possible to compute,
-for example, the upward of a gravity anomaly.
+for example, the upward continuation of a gravity anomaly.
+However, many authors in the literature have computed the upward 
+continuation of gravity anomalies. All these authors have implicitly 
+considered that the gravity anomaly is an approximation of 
+the gravity disturbance, which in turn can be represented by a 
+harmonic function that can be continued upward.
 

### 2018-04-22 manuscript/letter.tex

@@
 
-\title[Theoretical aspects in gravity modeling]
-      {Theoretical aspects in gravity modeling}
+\title[Theoretical aspects in gravity modelling]
+      {Theoretical aspects in gravity modelling}
 
-\author[Oliveira Jr, Barbosa and Uieda]
-{Vanderlei C. Oliveira Jr$^1$\thanks{Sei la, Nacional, Brazil}, Val\'{e}ria C. F. Barbosa$^1$ and Leonardo Uieda$^2$ \\
+\author[Oliveira Jr et al.]
+{Vanderlei C. Oliveira Jr$^1$\thanks{Email: vanderlei@on.br}, 
+Kristoffer A. T. Hallam$^{1}$, Val\'{e}ria C. F. Barbosa$^1$
+and Leonardo Uieda$^2$ \\
 $^1$ Department of Geophysics, Observat\'{o}rio Nacional, Rio de Janeiro, Brazil \\
@@
 \begin{summary}
- Gravity anomalies have long been used by geophysicists for the 
- purpose of determining density distributions in subsurface.
+ The gravity anomaly has long been used by geophysicists for 
+ representing the gravitational effect produced by gravity
+ sources.This quantity, however, can be used in this way as a good
+ approximation. Rigorously, the gravitational effect produced by the sources
+ is represented by the gravity disturbance, which is a well-known 
+ quantity in geodesy.
+ In this paper, we discuss the fundamental concepts supporting 
+ the fact that gravity disturbance is the more appropriated quantity
+ to be used for estimating density variations related to geological
+ bodies in subsurface. We stress that almost all authors in the 
+ literature have implicitly used the gravity anomaly as an approximation 
+ of the gravity disturbance, which can be represented by a harmonic 
+ function and, consequently, satisfies the premise behind most of 
+ the interpretation techniques used in geophysics (e.g., 
+ upward/downward continuation, 
+ data processing with equivalent layer,
+ conversions between gravity and magnetic data,
+ computation of vertical derivatives via Fourier and Hilbert
+ transforms).
+ Although the gravity anomaly can be considered a good approximation of the 
+ gravity disturbance for most practical applications, 
+ it is important to bear in mind not only the terminology 
+ changes, but also the conceptual assumptions.
 
- However, gravity anomalies
-
- In this paper, we discuss the fundamental concepts
-
- URGENTE: \citep{marussi1974}, \citep{torge2012}, section 4.2.1
 \end{summary}
@@
 \begin{keywords}
- potential fields -- gravity disturbance -- gravity anomaly -- gravity modeling.
+ potential fields -- gravity disturbance -- gravity anomaly -- gravity modelling.
 \end{keywords}
@@
 internal density distribution whereas geodesists use
-gravity to estimate the geoid \citep{li2001}.
+gravity to estimate the Geoid \citep{li2001}.
 Hence, geophysicists are usually interested 
@@
 the geological structures is commonly known in geophysics as 
-\textit{gravity modeling}.
-This term is different from \textit{gravity field modeling}, which
+\textit{gravity modelling}.
+This term is different from \textit{gravity field modelling}, which
 is commonly used in geodesy to denote the characterization of the 
@@
 
-\section{Terrestrial Reference Systems used in gravity modeling}
+\section{Terrestrial Reference Systems used in gravity modelling}
 
-For geophysical purposes, there are three important Terrestrial Reference Systems used in gravity modeling. 
+For geophysical purposes, there are three important Terrestrial Reference Systems used in gravity modelling. 
 They rotate with the Earth and are used for describing
@@
 the $X$-axis pointing to the Greenwich meridian and the $Y$-axis
-directed so as to obtain a right-handed system (Figure \ref{fig:GCS-GGS}).
+directed so as to obtain a right-handed system (Fig. \ref{fig:GCS-GGS}).
 This reference system can be found in the literature with different names: Mean Terrestrial System \citep[e.g.,][]{soler1976}, Earth-fixed geocentric Cartesian system \citep[e.g.,][]{torge2012} 
@@
 is defined by the \textit{geometric height} $h$, 
-\textit{geodetic latitude} $\varphi$ and \textit{longitude} $\lambda$ (Figure \ref{fig:GCS-GGS}).
+\textit{geodetic latitude} $\varphi$ and \textit{longitude} $\lambda$ (Fig. \ref{fig:GCS-GGS}).
 For convenience, we call this system Geocentric Geodetic System
@@
 At a given point $(h, \varphi, \lambda)$, there are three 
-mutually-orthogonal unit vectors (Figure \ref{fig:GCS-GGS}) given by \citep{soler1976}:
+mutually-orthogonal unit vectors (Fig. \ref{fig:GCS-GGS}) given by \citep{soler1976}:
 \begin{equation}
-\begin{split}
-\hat{\mitbf{u}} &= 
+\hat{\mitbf{u}} = 
 \begin{bmatrix}
@@
 \sin\varphi
-\end{bmatrix} \\
-\hat{\mitbf{v}} &= 
+\end{bmatrix}
+\hat{\mitbf{v}} = 
 \begin{bmatrix}
@@
 \cos\varphi
-\end{bmatrix} \\
-\hat{\mitbf{w}} &= 
+\end{bmatrix}
+\hat{\mitbf{w}} = 
 \begin{bmatrix}
@@
 0
-\end{bmatrix}
-\end{split} \: .
+\end{bmatrix} .
 \label{eq:unit-vectors}
@@
 \lambda)$ referred to the GGS into coordinates $(X, Y, Z)$ referred to 
-the GCS (Figure \ref{fig:GCS-GGS}) and vice versa can be easily found in 
+the GCS (Fig. \ref{fig:GCS-GGS}) and vice versa can be easily found in 
 the literature \citep[e.g.,][]{heiskanen-moritz1967, torge2012, 
@@
 Cartesian coordinate system (TCS) with origin at a point $P$ on or 
-close to the Earth's surface and axes $x$, $y$ and $z$ (Figure 
+close to the Earth's surface and axes $x$, $y$ and $z$ (Fig. 
 \ref{fig:TCS}). In the TCS, the axes $x$ and $y$ are parallel to 
@@
 whereas the $z$-axis is opposite to the unit vector 
-$\hat{\mitbf{u}}_{P}$ (equation \ref{eq:unit-vectors}) and points downward.
-Consider, for example, a TCS (Figure \ref{fig:TCS}) with origin at a point $P$ with
-coordinates $(X_{P}, Y_{P}, Z_{P})$ referred to the GCS (Figure \ref{fig:GCS-GGS}).
+$\hat{\mitbf{u}}_{P}$ (eq. \ref{eq:unit-vectors}) and points downward.
+Consider, for example, a TCS (Fig. \ref{fig:TCS}) with origin at a point $P$ with
+coordinates $(X_{P}, Y_{P}, Z_{P})$ referred to the GCS (Fig. \ref{fig:GCS-GGS}).
 In this case, the relationship between the coordinates $(X, Y, Z)$ of a 
@@
 are defined by the unit vectors $\hat{\mitbf{u}}$, $\hat{\mitbf{v}}$ and
-$\hat{\mitbf{w}}$ (equation \ref{eq:unit-vectors}) as follows:
+$\hat{\mitbf{w}}$ (eq. \ref{eq:unit-vectors}), evaluated at the
+point $P$, as follows:
 \begin{equation}
@@
 
-\section{Anomalous quantities in gravity modeling}
+\section{Anomalous quantities in gravity modelling}
 
@@
 using the last term.
-Figure \ref{fig:surfaces} illustrates the gravity vector 
+Fig. \ref{fig:surfaces} illustrates the gravity vector 
 $\mitbf{g}_{P}$, normal gravity vector $\mitbf{\gamma}_{P}$ and 
@@
 \begin{equation}
-\delta g_{P} = g_{P} - \gamma_{P} \: .
+\delta g_{P} = g_{P} - \gamma_{P} \: ,
 \label{eq:gravity-disturbance}
 \end{equation}
+where $g_{P} = \| \mitbf{g}_{P} \|$ and 
+$\gamma_{P} = \| \mitbf{\gamma}_{P} \|$ are,
+respectively, the gravity and the normal gravity at the
+point $P$.
 Notice that the gravity disturbance $\delta g_{P}$ is not equivalent 
@@
 
-%Consider a line (dashed line in Figure \ref{fig:surfaces}) passing 
-%through a point $Q$ on the surface of the reference ellipsoid and 
-%a point $P$ on the Earth's surface
-
 The gravity anomaly is the commonly used quantity in applied 
 geophysics. It is defined as the difference between the 
-gravity at a point $Q^{\prime}$ on the geoid (a particular 
+gravity at a point $Q^{\prime}$ on the Geoid (a particular 
 equipotential surface of the gravity potential) and the normal gravity 
 at a point $Q$ on the ellipsoid surface,
-both at the same geodetic latitude and longitude (Figure \ref{fig:surfaces}).
+both at the same geodetic latitude and longitude (Fig. \ref{fig:surfaces}).
 Notice that, by definition, the gravity anomaly depends on 
@@
 
-\section{Mathematical description of the gravity disturbance in a local coordinate system}
+\section{Mathematical description of the gravity disturbance at the TCS}
 
-In this coordinate system, the observed gravity vector
-$\mitbf{g}_{i}$, at a point $(x_{i}, y_{i}, z_{i})$, 
-$i = 1, ..., N$, can be represented by
+From eq. \ref{eq:gravity-disturbance-vector}, the gravity vector $\mitbf{g}_{i}$, at a point 
+$(x_{i}, y_{i}, z_{i})$ in the TCS (Fig. \ref{fig:TCS}b), can be represented by:
 \begin{equation}
-\mitbf{g}_{i} = \mitbf{\gamma}_{i} + \Delta \mitbf{g}_{i} \: ,
-\label{eq:gravity-vector}
+\mitbf{g}_{i} = \mitbf{\gamma}_{i} + \delta \mitbf{g}_{i} \: ,
+\label{eq:gravity-vector-TCS}
 \end{equation}
-where $\mitbf{\gamma}_{i}$ and $\Delta \mitbf{g}_{i}$
-are, respectively, the normal gravity vector and a
-disturbing gravitational attraction produced by the anomalous 
+where $\mitbf{\gamma}_{i}$ and $\delta \mitbf{g}_{i}$
+are, respectively, the normal gravity vector and the 
+gravity disturbance vector produced by the anomalous 
 masses at the point $(x_{i}, y_{i}, z_{i})$.
 
-For each point $(x_{i}, y_{i}, z_{i})$ in the topocentric
-Cartesian coordinate system (Figure \ref{fig:TCS}b), there is
-a corresponding point $(h_{i}, \varphi_{i}, \lambda_{i})$
-in the geodetic coordinate system (Figure \ref{fig:GCS-GGS}).
-
-By definition, the gravity disturbance $\delta g_{i}$,
-at the point $(x_{i}, y_{i}, z_{i})$, is given by 
-\citep{heiskanen-moritz1967, hofmann-wellenhof-moritz2005}:
-\begin{equation}
-\delta g_{i} =  g_{i} - \gamma_{i} \: ,
-\label{eq:gravity-disturbance}
-\end{equation}
-where $g_{i} = \| \mitbf{g}_{i} \|$ and 
-$\gamma_{i} = \| \mitbf{\gamma}_{i} \|$ are,
-respectively, the observed gravity and the normal gravity at the
-point $(x_{i}, y_{i}, z_{i})$.
-Fortunately, the condition 
-$\gamma_{i} \gg \| \Delta \mathbf{g}_{i} \|$ 
-is met at all points located above or on the Earth's surface.
-By combining this condition and the definition of observed gravity vector
-(equation \ref{eq:gravity-vector}), we can approximate the observed gravity
-$g_{i}$ by a first order Taylor's expansion as follows 
-\citep{sanso_sideris2013}:
+The gravity $g_{i}$ can be approximated by a first order Taylor's 
+expansion as follows \citep{sanso_sideris2013}:
 \begin{equation}
 g_{i} \approx \gamma_{i} + 
-\hat{\mitbf{\gamma}}_{i}^{\top} \Delta \mitbf{g}_{i} \: ,
+\hat{\mitbf{\gamma}}_{i}^{\top} \delta \mitbf{g}_{i} \: ,
 \label{eq:gobs-approx}
@@
 $\mitbf{\gamma}_{i}$ at the point $(x_{i}, y_{i}, z_{i})$, in the
-topocentric Cartesian coordinate system (Figure \ref{fig:TCS}b), and
+topocentric Cartesian coordinate system (Fig. \ref{fig:TCS}b), and
 $\hat{\mitbf{u}}_{i}$ is the unit vector $\hat{\mitbf{u}}$ 
-(equation \ref{eq:unit-vectors}) evaluated at the corresponding 
+(eq. \ref{eq:unit-vectors}) evaluated at the corresponding 
 point $(h_{i}, \varphi_{i}, \lambda_{i})$ in the 
-geodetic coordinate system (Figure \ref{fig:GCS-GGS}).
-
+GGS (Fig. \ref{fig:GCS-GGS}).
+This approximation can be made because, fortunately, 
+the condition $\gamma_{i} \gg \| \delta \mathbf{g}_{i} \|$ 
+is met at all points located above or on the Earth's surface.
 This approximation, which is known in geodesy \citep[e.g.,][]{sanso_sideris2013},
@@
 Notice that, in local- or regional-gravity studies, the unit
-vector $\hat{\mitbf{\gamma}}_{i}$ (equation \ref{eq:gobs-approx}) 
+vector $\hat{\mitbf{\gamma}}_{i}$ (eq. \ref{eq:gobs-approx}) 
 may be considered constant throughout the study area and 
-parallel to the $z$ axis of the topocentric Cartesian coordinate 
-system (Figure \ref{fig:TCS}b).
-Consequently, by using the approximation defined in equation 
+parallel to the $z$ axis of the TCS (Fig. \ref{fig:TCS}b).
+By using the approximation defined in eq. 
 \ref{eq:gobs-approx}, the gravity disturbance 
-(equation \ref{eq:gravity-disturbance}) can be rewritten as 
+(eq. \ref{eq:gravity-disturbance}) can be rewritten as 
 follows 
 \begin{equation}
-\delta g_{i} \approx \hat{\mitbf{\gamma}}_{P}^{\top} \Delta \mitbf{g}_{i} \: ,
+\delta g_{i} \approx \hat{\mitbf{\gamma}}_{P}^{\top} \delta \mitbf{g}_{i} \: ,
 \label{eq:gravity-disturbance-approx}
@@
 the same direction as the normal gravity vector $\mitbf{\gamma}_{P}$
-at the origin $P$ of the topocentric Cartesian coordinate system
-(Figure \ref{fig:TCS}).
-This equation shows that the gravity disturbance $\delta g_{i}$ (equation \ref{eq:gravity-disturbance}) is different from the magnitude
-of the disturbing gravitational attraction $\Delta \mitbf{g}_{i}$
+at the origin $P$ of the TCS (Fig. \ref{fig:TCS}b).
+This equation shows that the gravity disturbance $\delta g_{i}$ (eq. \ref{eq:gravity-disturbance}) is different from the magnitude
+of the gravity disturbance vector $\delta \mitbf{g}_{i}$
 produced by the gravity sources. Rather, it represents the 
-component of the $\Delta \mitbf{g}_{i}$ on the direction of the normal
+component of $\delta \mitbf{g}_{i}$ on the direction of the normal
 gravity vector \citep{sanso_sideris2013}.
-In the topocentric Cartesian coordinate system,
-the gravity disturbance $\delta g_{i}$ (equation \ref{eq:gravity-disturbance-approx})
-can be defined as the vertical component of the gravitational attraction exerted by
+
+In the TCS (Fig. \ref{fig:TCS}b),
+the gravity disturbance $\delta g_{i}$ (eq. \ref{eq:gravity-disturbance-approx})
+can be defined as the $z$-component of the gravitational attraction exerted by
 the gravity sources at the point $(x_{i}, y_{i}, z_{i})$.
 As a consequence, the gravity disturbance produced by
-a homogeneous gravity source with density contrast $\Delta\rho$
-(in $kg / m^{3}$) can be represented by the following harmonic function:
+a gravity source can be represented by the following harmonic function:
 \begin{equation}
-d_{i} = k_{g} \, G \, \Delta\rho \, \partial_{z} \phi_{i} \: ,
+d_{i} = \partial_{z} \phi_{i} \: ,
 \label{eq:gz-local}
 \end{equation}
-where $G$ is the Newtonian constant of gravitation
-(in $m^{3} / (kg \, s^{2})$),
-$k_{g} = 10^{5}$ is a constant factor
-transforming from $m/s^{2}$ to milligal (mGal),
-and $\partial_{z} \phi_{i}$ is a harmonic function representing the first 
-derivative, evaluated at the observation point $(x_{i},y_{i},z_{i})$, 
-$i = 1, ..., N$, of the function 
+where $\partial_{z} \phi_{i}$ is the first 
+derivative, evaluated at the point $(x_{i},y_{i},z_{i})$, of the function 
 \begin{equation}
-\phi(x,y,z) = \int\int\limits_{v}\int \frac{1}{\sqrt{(x - x^{\prime})^{2} + 
-(y - y^{\prime})^{2} + (z - z^{\prime})^{2}}} \: dv
+\phi(x,y,z) = \int\int\limits_{v}\int 
+\frac{c_{g} \, G \, \Delta\rho(x^{\prime}, y^{\prime}, z^{\prime}) \: dv}
+{\sqrt{(x - x^{\prime})^{2} + 
+(y - y^{\prime})^{2} + (z - z^{\prime})^{2}}} \: ,
 \label{eq:phi}
 \end{equation}
-with respect to the variable $z$. The 
-integral is conducted over the coordinates $x^{\prime}$, $y^{\prime}$ 
-and $z^{\prime}$ within the volume $v$ of the gravity source.
+with respect to the variable $z$. Is this integral, 
+$G$ is the Newtonian constant of gravitation (in $m^{3} kg^{-1} s^{-2}$),
+$c_{g} = 10^{5}$ is a constant factor transforming from $m s^{-2}$ 
+to milligal (mGal), $\Delta\rho(x^{\prime}, y^{\prime}, z^{\prime})$
+is the density contrast (in $kg / m^{3}$) at a point 
+$(x^{\prime}, y^{\prime}, z^{\prime})$ within 
+the volume $v$ of the gravity source and the integration is conducted 
+over $x^{\prime}$, $y^{\prime}$ and $z^{\prime}$.
 This equation can be easily generalized for the case
@@
 
-Practically all the literature about gravity modeling
-use the quantity $d_{i}$ (equation \ref{eq:gz-local})
+Practically all the literature about gravity modelling
+use the quantity $d_{i}$ (eq. \ref{eq:gz-local})
 to represent the gravity anomaly produced by 
 gravity sources \citep[e.g.,][]{blakely1996}. 
-Consequently, almost all the geophysicists use the gravity
+Consequently, almost all geophysicists use the gravity
 anomaly as an approximation of the gravity disturbance
 produced by the gravity sources. Notice that $d_{i}$ 
-(equation \ref{eq:gz-local}) does not depend on the
-geoidal surface. Rather, it depends on the relative position 
+(eq. \ref{eq:gz-local}) does not depend on the
+height with respect to the Geoid (orthometric height). 
+Rather, it depends on the relative position 
 of the observation points $(x_{i}, y_{i}, z_{i})$ with respect to
-the gravity sources in a topocentric Cartesian coordinate
-system (Figure \ref{fig:TCS}b).
+the gravity sources in the TCS (Fig. \ref{fig:TCS}b).
+It is important to notice that, contrary to what is written in 
+some basic texts, gravimeters do not measure the quantity
+$d_{i}$ (eq. \ref{eq:gz-local}).
 
@@
 disturbance and the gravity anomaly.
-Our reasoning suggest that the gravity disturbance is the
+Our reasoning suggests that the gravity disturbance is the
 more appropriated quantity for representing the gravity
@@
 transforms). As a consequence, they implicitly or directly assume that
-the gravity data approximates the gravitational disturbance.
+the gravity data approximate the gravitational disturbance.
 
-\item Almost all forward modeling techniques compute
+\item Almost all forward modelling techniques compute
 the vertical component of the gravitational attraction 
-exerted by the geological bodies at the observation points.
+exerted by the gravity sources at the observation points.
 Notice that the gravity anomaly requires the computation of gravity
 on the Geoid, which is generally within the topographic masses.
-Hence, almost all geophysicists implicitly compute 
+Hence, almost all forward modelling techniques compute 
 the gravity disturbance.
@@
 the gravity disturbance can be computed at arbitrary points
-outside the sources.
+outside the sources. Then, computing the upward 
+continuation of gravity anomalies implicitly assumes 
+that they approximate the gravity disturbance.
 
@@
 
-The principal theoretical implication of this study is that
-is that, although the gravity anomaly may be used as a good approximation of
-the gravity effect produced by the sources for most practical applications,
-the more appropriated quantity for gravity modeling is the
+The principal theoretical implication of this study is that,
+although the gravity anomaly may be used as a good approximation of
+the gravitational effect produced by the sources for most practical applications,
+the more appropriated quantity for gravity modelling is the
 gravity disturbance. We stress that, more important than the 
-the terminology changes, the geophysicist must bear in mind the 
-conceptual assumptions used in gravity modeling.
+the terminology changes, the authors must bear in mind the 
+conceptual assumptions used in gravity modelling.
 
@@
     $\hat{\mitbf{w}}_{P}$ define three mutually orthogonal 
-    directions at $P$ (equation \ref{eq:unit-vectors}).
+    directions at $P$ (eq. \ref{eq:unit-vectors}).
     In (b), $Q$ represents a point $(h_{Q}, \varphi_{Q}, \lambda_{Q})$,
@@
     gravity disturbance vector $\delta \mitbf{g}_{P}$
-    (equation \ref{eq:gravity-vector}) and unit vector 
-    $\hat{\mitbf{u}}_{P}$ (equation \ref{eq:unit-vectors}) at a point 
+    (eq. \ref{eq:gravity-disturbance-vector}) and unit vector 
+    $\hat{\mitbf{u}}_{P}$ (eq. \ref{eq:unit-vectors}) at a point 
     $P$.
@@
     and $y$ are parallel to the unit vectors $\hat{\mitbf{v}}_{P}$
-    and $\hat{\mitbf{w}}_{P}$ (equation \ref{eq:unit-vectors} and
-    Figure \ref{fig:GCS-GGS}), respectively. On the other hand, the $z$ axis is opposite to the unit vector $\hat{\mitbf{u}}_{P}$ (equation 
-    \ref{eq:unit-vectors} and Figure \ref{fig:GCS-GGS}) and points downward. The gray plane shown in (b) is the same gray plane
-    shown in Figure \ref{fig:GCS-GGS}.}
+    and $\hat{\mitbf{w}}_{P}$ (eq. \ref{eq:unit-vectors} and
+    Fig. \ref{fig:GCS-GGS}), respectively. On the other hand, the $z$ axis is opposite to the unit vector $\hat{\mitbf{u}}_{P}$ (eq. 
+    \ref{eq:unit-vectors} and Fig. \ref{fig:GCS-GGS}) and points downward. The gray plane shown in (b) is the same gray plane
+    shown in Fig. \ref{fig:GCS-GGS}.}
   \label{fig:TCS}
@@
         gravity disturbance vector $\delta \mitbf{g}_{P}$
-        (equation \ref{eq:gravity-vector}) and unit vector 
-        $\hat{\mitbf{u}}_{P}$ (equation \ref{eq:unit-vectors}) at a point $P$ on the surface of the Earth; gravity vector 
+        (eq. \ref{eq:gravity-disturbance-vector}) and unit vector 
+        $\hat{\mitbf{u}}_{P}$ (eq. \ref{eq:unit-vectors}) at a point $P$ on the surface of the Earth; gravity vector 
         $\mitbf{g}_{Q^{\prime}}$ at a point $Q^{\prime}$ on the Geoid;
         normal gravity vector $\mitbf{\gamma}_{Q}$ at a point $Q$ on
-        the reference ellipsoid. The dashed line is normal to the 
+        the reference ellipsoid. The dashed line passing through
+        $Q$, $Q^{\prime}$ and $P$ is normal to the 
         surface of the reference ellipsoid at $Q$.}

### 2018-04-26 manuscript/letter.tex
Including freeair and disturbance comparison

@@
 
-Other important reference system is a geocentric system of 
+Another important reference system is a geocentric system of 
 geodetic coordinates, which is defined by the reference ellipsoid 
@@
 motion and time variations such as Earth tides, instrumental drift and 
-barometric pressure changes, for example) and the normal gravity vector 
+barometric pressure changes, for instance) and the normal gravity vector 
 at a point $P$. 
@@
 \end{equation}
-As the centrifugal part of the gravity vector and the normal gravity vector 
+As the centrifugal part of the normal gravity vector 
 is equal to the centrifugal component of the gravity vector, 
@@
 The gravity anomaly is the commonly used quantity in applied 
-geophysics. It is defined as the difference between the 
-gravity at a point $Q^{\prime}$ on the Geoid (a particular 
+geophysics. At point $P$, it is defined as the difference between the 
+gravity vector at a point $Q^{\prime}$ on the Geoid (a particular 
 equipotential surface of the gravity potential) and the normal gravity 
-at a point $Q$ on the ellipsoid surface,
-both at the same geodetic latitude and longitude (Fig. \ref{fig:surfaces}).
+vector at a point $Q$ on the ellipsoid surface, both at the same
+geodetic latitude and longitude (Fig. \ref{fig:surfaces}) as in
+\begin{equation}
+\Delta \mitbf{g}_{P} = \mitbf{g}_{Q^{\prime} } - \mitbf{\gamma}_{Q} .
+\label{eq:gravity-anomaly-vector}
+\end{equation}
+Similarly as the gravity disturbance expression (eq.
+\ref{eq:gravity-disturbance}), the gravity anomaly is given by
+\begin{equation}
+\Delta g_{P} = g_{Q^{\prime} } - \gamma_{Q} .
+\label{eq:gravity-anomaly}
+\end{equation}
 Notice that, by definition, the gravity anomaly depends on 
@@
 for example, the upward continuation of a gravity anomaly.
-However, many authors in the literature have computed the upward 
-continuation of gravity anomalies. All these authors have implicitly 
-considered that the gravity anomaly is an approximation of 
+However, many authors in the literature have. All these authors have
+implicitly considered that the gravity anomaly is an approximation of 
 the gravity disturbance, which in turn can be represented by a 
-harmonic function that can be continued upward.
+harmonic function that is allowed to be upward continued.
 
@@
 
+\section{Comparing Free-air anomaly with gravity disturbance}
+
+By definition, for a small height $h$, the normal gravity at point
+$P(h, \varphi, \lambda)$ can be expanded into a Taylor series 
+in terms of $h$ \citep{heiskanen-moritz1967,
+hofmann-wellenhof-moritz2005} as in
+\begin{equation}
+\gamma_{P} \approx \gamma_{Q} + \frac{\partial \gamma}{\partial h} h ,
+\label{eq:gamma-expansion}
+\end{equation}
+where higher orders of $h$ were neglected. $h$ represents the
+distance between points $P$ and $Q$ shown in (Fig. \ref{fig:heights}).
+Therefore, substituting \ref{eq:gamma-expansion}'s result into
+eq. \ref{eq:gravity-disturbance}, the gravity disturbance becomes
+\begin{equation}
+\delta g_{P} \approx g_{P} - (\gamma_{Q} + \frac{\partial \gamma}{\partial h} h) .
+\label{eq:disturb-approx}
+\end{equation}
+According to \citet{heiskanen-moritz1967, blakely1996}, the relation
+between gravity on the geoid and on the reference ellipsoid is
+a similar Taylor expansion given by
+\begin{equation}
+g_{Q^{\prime}} = g_{P} - \frac{\partial \gamma}{\partial h} H
+\label{eq:geod2phys_surf}
+\end{equation}
+where $H$ is defined as the orthometric height (Fig. \ref{fig:heights})
+measured along the curved plumb line, starting from the Geoid
+\citep{heiskanen-moritz1967, hofmann-wellenhof-moritz2005}.
+Thus, the Free-air anomaly $(\Delta g^{a})$, which corrects for the
+decrease in gravity due to the elevation of the observation point $P$, 
+can be written as
+\begin{equation}
+\Delta g^{a}_{P} \approx g_{P} - (\gamma_{Q} + \frac{\partial \gamma}{\partial h} H)
+\label{eq:free-air}
+\end{equation}
+By inspection, one easily concludes that the difference is the first
+order derivative term
+\begin{equation}
+\delta g_{P} - \Delta g^{a}_{P} \approx - \frac{\partial \gamma}{\partial h} (h-H) = - \frac{\partial \gamma}{\partial h} N ,
+\label{eq:disturb-FA}
+\end{equation}
+where $N$ is the geoidal undulation (Fig. \ref{fig:heights}) and 
+$\frac{\partial \gamma}{\partial h} = -0.3086$.  Hence, using a
+mean value of $\tilde{N} = 120$m, the absolute value of the
+difference in eq. \ref{eq:disturb-FA} is computed
+\begin{equation}
+\parallel \delta g_{P} - \Delta g^{a}_{P} \parallel \leq \parallel - \frac{\partial \gamma}{\partial h} \tilde{N} \parallel = 37.032
+\label{eq:diff-constant}
+\end{equation}
+
 \section{Conclusions}
@@
 
+\begin{figure}
+    \includegraphics{figures/different-heights.png}
+    \caption{Schematic representation of the geometric height
+        $h$, orthometric height $H$ and geoidal undulation $N$.
+        $H$ is curved because it follows the plumbline.}
+  \label{fig:heights}
+\end{figure}
+
 \label{lastpage}

### 2018-04-27 manuscript/letter.tex
relationship between disturbance and free-air

@@
  to be used for estimating density variations related to geological
- bodies in subsurface. We stress that almost all authors in the 
- literature have implicitly used the gravity anomaly as an approximation 
- of the gravity disturbance, which can be represented by a harmonic 
- function and, consequently, satisfies the premise behind most of 
+ bodies in subsurface. We show that the maximum absolute difference 
+ between the gravity disturbance and the Free-air anomaly in Brazil 
+ reaches a maximum absolute value of $\approx 9.3$ mGal. We also stress 
+ that almost all authors in the literature have implicitly used the 
+ gravity anomaly as an approximation of the gravity disturbance, 
+ which can be represented by a harmonic function and, consequently, 
+ satisfies the premise behind most of 
  the interpretation techniques used in geophysics (e.g., 
- upward/downward continuation, 
- data processing with equivalent layer,
- conversions between gravity and magnetic data,
- computation of vertical derivatives via Fourier and Hilbert
- transforms).
+ upward/downward continuation, data processing with equivalent layer,
+ conversions between gravity and magnetic data, computation of vertical 
+ derivatives via Fourier and Hilbert transforms).
  Although the gravity anomaly can be considered a good approximation of the 
@@
 
-\section{Anomalous quantities in gravity modelling}
+\section{Relationship between gravity disturbance and gravity anomaly}
 
@@
 \begin{equation}
-\Delta \mitbf{g}_{P} = \mitbf{g}_{Q^{\prime} } - \mitbf{\gamma}_{Q} .
+\Delta \mitbf{g}_{P} = \mitbf{g}_{Q^{\prime}} - \mitbf{\gamma}_{Q} .
 \label{eq:gravity-anomaly-vector}
 \end{equation}
-Similarly as the gravity disturbance expression (eq.
+Similarly to the gravity disturbance expression (eq.
 \ref{eq:gravity-disturbance}), the gravity anomaly is given by
 \begin{equation}
-\Delta g_{P} = g_{Q^{\prime} } - \gamma_{Q} .
+\Delta g_{P} = g_{Q^{\prime}} - \gamma_{Q} ,
 \label{eq:gravity-anomaly}
 \end{equation}
+where $g_{Q^{\prime} } = \| \mitbf{g}_{Q^{\prime}} \|$ is
+the magnitude of the gravity vector $\mitbf{g}_{Q}$ at $Q^{\prime}$, 
+on the Geoid, and $\gamma_{Q} = \| \mitbf{\gamma}_{Q} \|$ is the 
+magnitude of the normal gravity vector $\mitbf{\gamma}_{Q}$ at $Q$, 
+on the reference ellipsoid (Fig. \ref{fig:surfaces}).
 Notice that, by definition, the gravity anomaly depends on 
@@
 for example, the upward continuation of a gravity anomaly.
-However, many authors in the literature have. All these authors have
+However, many authors in the literature have computed the upward 
+continuation of gravity anomalies. All these authors have
 implicitly considered that the gravity anomaly is an approximation of 
@@
 Bouguer anomalies), depending on the corrections applied to 
-them \citep{blakely1996, hofmann-wellenhof-moritz2005}.
-These corrections are usually called gravity reductions.
-It can be shown that the Free-air anomaly is an approximation of the
-gravity disturbance whereas the Bouguer anomaly
+them. These corrections are usually called gravity reductions. 
+The Free-air anomaly, for example, is defined as follows 
+\citep{blakely1996, hofmann-wellenhof-moritz2005}:
+\begin{equation}
+\Delta g_{P}^{a} 
+= g_{P} - \left( \gamma_{Q} + \frac{\partial \gamma}{\partial h} \, H \right)
+\: ,
+\label{eq:free-air-anomaly}
+\end{equation}
+where $\frac{\partial \gamma}{\partial h} \approx -0.3086$ mGal/m is the
+derivative of the normal gravity with respect to the geometric height $h$
+and $H$ is the orthometric height (Fig. \ref{fig:surfaces}). 
+Notice that the term between parenthesis represents an 
+approximation of the normal gravity at the point $Q^{\prime}$ on the
+Geoid. By using a similar approach, we can approximate the normal gravity
+$\gamma_{P}$, at a point $P$ on the Earth's surface, so that the 
+gravity disturbance $\delta g_{P}$ (eq. \ref{eq:gravity-disturbance}) 
+can be rewritten as follows:
+\begin{equation}
+\delta g_{P} \approx g_{P} - 
+\left( \gamma_{Q} + \frac{\partial \gamma}{\partial h} h \right) \: .
+\label{eq:gravity-disturbance-approx}
+\end{equation}
+By inspection, one easily concludes that the absolute difference 
+between the approximated gravity disturbance $\delta g_{P}$ 
+(eq. \ref{eq:gravity-disturbance-approx}), at a point $P$ on or close to
+the Earth's surface, and the Free-air anomaly
+$\Delta g_{P}^{a}$ (eq. \ref{eq:free-air-anomaly}), is given by:
+\begin{equation}
+\vert \delta g_{P} - \Delta g^{a}_{P} \vert \approx 
+\vert \frac{\partial \gamma}{\partial h} N \vert \: ,
+\label{eq:difference-disturbance-free-air}
+\end{equation}
+where the geoidal undulation $N \approx h - H$ (Fig. \ref{fig:surfaces}) assumes 
+that the Geoid and the surface of the reference ellipsoid 
+are approximately parallel at $P$ and the surrounding area.
+We know empirically that the geoidal undulation $N$ in the world 
+is $\approx \pm 1$ on the oceans and reaches a 
+maximum absolute value of $\approx 120$ m \citep[e.g.,][]{torge2012, 
+sanso_sideris2013}. In Brazil, for example, the geoidal undulation 
+reaches $\approx \pm 30$ m \citep{ibge_mapgeo2015}. Consequently, the
+maximum absolute differences between gravity disturbance and
+Free-air anomaly, according to eq. \ref{eq:difference-disturbance-free-air},
+is $\approx 9.258$ mGal.
+
+The approximations defined by eqs \ref{eq:gravity-disturbance-approx}
+and \ref{eq:difference-disturbance-free-air} are commonly used
+in geodesy to define the \textit{fundamental equation of physical
+geodesy} \citep{hofmann-wellenhof-moritz2005}.
+A similar approach can be used to show that the Bouguer anomaly
 is an approximation of the terrain corrected gravity disturbance.
@@
 \delta g_{i} \approx \hat{\mitbf{\gamma}}_{P}^{\top} \delta \mitbf{g}_{i} \: ,
-\label{eq:gravity-disturbance-approx}
+\label{eq:gravity-disturbance-approx-TCS}
 \end{equation}
@@
 component of $\delta \mitbf{g}_{i}$ on the direction of the normal
-gravity vector \citep{sanso_sideris2013}.
+gravity vector \citep{hofmann-wellenhof-moritz2005, sanso_sideris2013}.
 
 In the TCS (Fig. \ref{fig:TCS}b),
-the gravity disturbance $\delta g_{i}$ (eq. \ref{eq:gravity-disturbance-approx})
+the gravity disturbance $\delta g_{i}$ (eq. \ref{eq:gravity-disturbance-approx-TCS})
 can be defined as the $z$-component of the gravitational attraction exerted by
@@
 
-\section{Comparing Free-air anomaly with gravity disturbance}
-
-By definition, for a small height $h$, the normal gravity at point
-$P(h, \varphi, \lambda)$ can be expanded into a Taylor series 
-in terms of $h$ \citep{heiskanen-moritz1967,
-hofmann-wellenhof-moritz2005} as in
-\begin{equation}
-\gamma_{P} \approx \gamma_{Q} + \frac{\partial \gamma}{\partial h} h ,
-\label{eq:gamma-expansion}
-\end{equation}
-where higher orders of $h$ were neglected. $h$ represents the
-distance between points $P$ and $Q$ shown in (Fig. \ref{fig:heights}).
-Therefore, substituting \ref{eq:gamma-expansion}'s result into
-eq. \ref{eq:gravity-disturbance}, the gravity disturbance becomes
-\begin{equation}
-\delta g_{P} \approx g_{P} - (\gamma_{Q} + \frac{\partial \gamma}{\partial h} h) .
-\label{eq:disturb-approx}
-\end{equation}
-According to \citet{heiskanen-moritz1967, blakely1996}, the relation
-between gravity on the geoid and on the reference ellipsoid is
-a similar Taylor expansion given by
-\begin{equation}
-g_{Q^{\prime}} = g_{P} - \frac{\partial \gamma}{\partial h} H
-\label{eq:geod2phys_surf}
-\end{equation}
-where $H$ is defined as the orthometric height (Fig. \ref{fig:heights})
-measured along the curved plumb line, starting from the Geoid
-\citep{heiskanen-moritz1967, hofmann-wellenhof-moritz2005}.
-Thus, the Free-air anomaly $(\Delta g^{a})$, which corrects for the
-decrease in gravity due to the elevation of the observation point $P$, 
-can be written as
-\begin{equation}
-\Delta g^{a}_{P} \approx g_{P} - (\gamma_{Q} + \frac{\partial \gamma}{\partial h} H)
-\label{eq:free-air}
-\end{equation}
-By inspection, one easily concludes that the difference is the first
-order derivative term
-\begin{equation}
-\delta g_{P} - \Delta g^{a}_{P} \approx - \frac{\partial \gamma}{\partial h} (h-H) = - \frac{\partial \gamma}{\partial h} N ,
-\label{eq:disturb-FA}
-\end{equation}
-where $N$ is the geoidal undulation (Fig. \ref{fig:heights}) and 
-$\frac{\partial \gamma}{\partial h} = -0.3086$.  Hence, using a
-mean value of $\tilde{N} = 120$m, the absolute value of the
-difference in eq. \ref{eq:disturb-FA} is computed
-\begin{equation}
-\parallel \delta g_{P} - \Delta g^{a}_{P} \parallel \leq \parallel - \frac{\partial \gamma}{\partial h} \tilde{N} \parallel = 37.032
-\label{eq:diff-constant}
-\end{equation}
-
 \section{Conclusions}
@@
 
+\item In Brazil, the maximum difference between the gravity 
+disturbance (on or close to the Earth's surface) and the Free-air
+anomaly reaches $\approx 9.3$ mGal.
+
 \end{enumerate}
@@
 gravity disturbance. We stress that, more important than the 
-the terminology changes, the authors must bear in mind the 
+terminology changes, the researchers must bear in mind the 
 conceptual assumptions used in gravity modelling.
@@
         gravity disturbance vector $\delta \mitbf{g}_{P}$
-        (eq. \ref{eq:gravity-disturbance-vector}) and unit vector 
-        $\hat{\mitbf{u}}_{P}$ (eq. \ref{eq:unit-vectors}) at a point $P$ on the surface of the Earth; gravity vector 
-        $\mitbf{g}_{Q^{\prime}}$ at a point $Q^{\prime}$ on the Geoid;
-        normal gravity vector $\mitbf{\gamma}_{Q}$ at a point $Q$ on
-        the reference ellipsoid. The dashed line passing through
+        (eq. \ref{eq:gravity-disturbance-vector}), unit vector 
+        $\hat{\mitbf{u}}_{P}$ (eq. \ref{eq:unit-vectors}) at a point $P$ 
+        on the surface of the Earth, gravity vector $\mitbf{g}_{Q^{\prime}}$ 
+        at a point $Q^{\prime}$ on the Geoid, normal gravity vector 
+        $\mitbf{\gamma}_{Q}$ at a point $Q$ on the reference ellipsoid,
+        geometric height $h$, orthometric height $H$ and geoidal undulation 
+        $N$ \citep{heiskanen-moritz1967}.
+        The dashed line passing through
         $Q$, $Q^{\prime}$ and $P$ is normal to the 
-        surface of the reference ellipsoid at $Q$.}
+        surface of the reference ellipsoid at $Q$. This figure
+        shows a commonly used approximation in which the ellipsoid 
+        surface and the Geoid are represented as parallel surfaces, 
+        so that $h \approx H + N$.}
   \label{fig:surfaces}
-\end{figure}
-
-\begin{figure}
-    \includegraphics{figures/different-heights.png}
-    \caption{Schematic representation of the geometric height
-        $h$, orthometric height $H$ and geoidal undulation $N$.
-        $H$ is curved because it follows the plumbline.}
-  \label{fig:heights}
 \end{figure}

### 2018-04-27 manuscript/letter.tex
Updating article

@@
 
-The gravity anomaly is the commonly used quantity in applied 
-geophysics. At point $P$, it is defined as the difference between the 
-gravity vector at a point $Q^{\prime}$ on the Geoid (a particular 
-equipotential surface of the gravity potential) and the normal gravity 
-vector at a point $Q$ on the ellipsoid surface, both at the same
-geodetic latitude and longitude (Fig. \ref{fig:surfaces}) as in
+Another anomalous quantity is the \textit{gravity anomaly vector}.
+It is defined, at point $P$, as the difference between the gravity
+vector at a point $Q^{\prime}$ on the Geoid (a particular
+equipotential surface of the gravity potential) and the normal
+gravity vector at a point $Q$ on the ellipsoid surface, both at the
+same geodetic latitude and longitude (Fig. \ref{fig:surfaces}) as in
 \begin{equation}
@@
 transforms). As a consequence, they implicitly or directly assume that
-the gravity data approximate the gravitational disturbance.
+the gravity data approximate the gravity disturbance.
 
@@
 on the Geoid, which is generally within the topographic masses.
-Hence, almost all forward modelling techniques compute 
-the gravity disturbance.
+Hence, almost all these forward modelling techniques in fact
+compute the gravity disturbance.
 

### 2018-05-06 manuscript/letter.tex
small change conclusion, deletion unused files
Val's general review

@@
 
-\title[Theoretical aspects in gravity modelling]
-      {Theoretical aspects in gravity modelling}
+\title[Gravity anomaly or gravity disturbance?]
+      {Should geophysicists use gravity anomaly or gravity disturbance?}
 
-\author[Oliveira Jr et al.]
-{Vanderlei C. Oliveira Jr$^1$\thanks{Email: vanderlei@on.br}, 
-Kristoffer A. T. Hallam$^{1}$, Val\'{e}ria C. F. Barbosa$^1$
-and Leonardo Uieda$^2$ \\
+\author[Hallam et al.]
+{Kristoffer A. T. Hallam$^{1}$\thanks{Emails: kristoffer.hallam@gmail.com,
+vanderlei@on.br}, Vanderlei C. Oliveira Jr$^1$,
+Val\'{e}ria C. F. Barbosa$^1$ \linebreak and Leonardo Uieda$^2$ \\
 $^1$ Department of Geophysics, Observat\'{o}rio Nacional, Rio de Janeiro, Brazil \\
@@
 
-The first step of the procedure for isolating this 
+For isolating this 
 gravitational component consists in removing the non-gravitational 
@@
 is commonly used in geodesy to denote the characterization of the 
-gravity field in a local, regional or global scale.
+gravity field in a local, regional or global scales.
 
 Based on well-established concepts of the literature,
-we present a discussion aiming at bringing some
-light to the following question: in geophysical applications,
+we present a discussion aiming at critically examining
+the following question: in geophysical applications,
 should we use the gravity disturbance or gravity anomaly?
@@
 by the gravity field produced by a reference ellipsoid
-(or level ellipsoid), which is rigid and geocentric.
+(or level ellipsoid), which is a rigid and geocentric model.
 This reference ellipsoid has the minor axis $b$ 
@@
 
-Similarly to the gravity vector and gravity, 
-the resultant of the virtual 
+The resultant of the virtual 
 gravitational and centrifugal forces exerted by the normal
@@
 \end{equation}
-where $\mitbf{R}$ is an orthogonal matrix whose columns
-are defined by the unit vectors $\hat{\mitbf{u}}$, $\hat{\mitbf{v}}$ and
-$\hat{\mitbf{w}}$ (eq. \ref{eq:unit-vectors}), evaluated at the
-point $P$, as follows:
-\begin{equation}
-\mitbf{R} = 
-\begin{bmatrix}
-\hat{\mitbf{v}} & \hat{\mitbf{w}} & -\hat{\mitbf{u}}
-\end{bmatrix} \: .
-\label{eq:R}
-\end{equation}
+where $\mitbf{R}$ is a $3 \times 3$ orthogonal matrix whose 
+first, second and third columns are defined, respectively, 
+by the unit vectors $\hat{\mitbf{v}}$, $\hat{\mitbf{w}}$ 
+and $-\hat{\mitbf{u}}$ (eq. \ref{eq:unit-vectors}), 
+evaluated at the point $P$.
 
@@
 to that of the gravity potential.
-Then, the both gravity vector and normal gravity vectors are represented 
-by the sum of two vectors: the gradient of a gravitational potential and
-the gradient of a centrifugal potential. Notice that, as the centrifugal 
-part of the gravity potential and normal potential are the same, the
-centrifugal part of the gravity vector and normal gravity vector are
-also equal to each other.
 
@@
 \end{equation}
-As the centrifugal part of the normal gravity vector 
+Because the centrifugal part of the normal gravity vector 
 is equal to the centrifugal component of the gravity vector, 
@@
 caused by contrasts between the actual internal 
-density distribution of the Earth and the (unknown) internal density 
+density distribution of the Earth and the internal density 
 distribution of the normal Earth.
+It is important to stress that both density distributions are unknown.
 In applied geophysics, these density differences are generally 
@@
 
-The difference between gravity $g_{P}$ (magnitude of the gravity 
-vector $\mitbf{g}_{P}$) and normal gravity $\gamma_{P}$ 
-(magnitude of the normal gravity vector $\mitbf{\gamma}_{P}$), at the 
+The difference between the magnitudes of gravity vector
+$g_{P} = \| \mitbf{g}_{P} \|$ and the normal gravity vector
+$\gamma_{P} = \| \mitbf{\gamma}_{P} \|$, at the 
 same point $P$, is called \textit{gravity disturbance} 
@@
 \begin{equation}
-\delta g_{P} = g_{P} - \gamma_{P} \: ,
+\delta g_{P} = g_{P} - \gamma_{P} \: .
 \label{eq:gravity-disturbance}
 \end{equation}
-where $g_{P} = \| \mitbf{g}_{P} \|$ and 
-$\gamma_{P} = \| \mitbf{\gamma}_{P} \|$ are,
-respectively, the gravity and the normal gravity at the
-point $P$.
 Notice that the gravity disturbance $\delta g_{P}$ is not equivalent 
@@
 $\delta \mitbf{g}_{P}$ \citep{barthelmes2013, sanso_sideris2013}.
-As properly pointed out by \citet{hackney-featherstone2003},
-the gravity disturbance is a very-well established quantity in geodesy,
-but appears to be less well known in geophysics.
-
 Another anomalous quantity is the \textit{gravity anomaly vector}.
@@
 where $g_{Q^{\prime} } = \| \mitbf{g}_{Q^{\prime}} \|$ is
-the magnitude of the gravity vector $\mitbf{g}_{Q}$ at $Q^{\prime}$, 
-on the Geoid, and $\gamma_{Q} = \| \mitbf{\gamma}_{Q} \|$ is the 
-magnitude of the normal gravity vector $\mitbf{\gamma}_{Q}$ at $Q$, 
-on the reference ellipsoid (Fig. \ref{fig:surfaces}).
+the magnitude of the gravity vector $\mitbf{g}_{Q}$ on the Geoid,
+at $Q^{\prime}$, and $\gamma_{Q} = \| \mitbf{\gamma}_{Q} \|$ is the 
+magnitude of the normal gravity vector $\mitbf{\gamma}_{Q}$ 
+on the reference ellipsoid, at the point $Q$ (Fig. \ref{fig:surfaces}).
 Notice that, by definition, the gravity anomaly depends on 
@@
 
-Different gravity anomalies can be calculated (e.g., Free-air and 
-Bouguer anomalies), depending on the corrections applied to 
-them. These corrections are usually called gravity reductions. 
-The Free-air anomaly, for example, is defined as follows 
-\citep{blakely1996, hofmann-wellenhof-moritz2005}:
+Different gravity anomalies can be calculated, 
+depending on the corrections applied to them. 
+These corrections are usually called \textit{gravity reductions}. 
+The Free-air anomaly, for example, may be defined as 
+follows \citep{blakely1996, hofmann-wellenhof-moritz2005}:
 \begin{equation}
-\Delta g_{P}^{a} 
-= g_{P} - \left( \gamma_{Q} + \frac{\partial \gamma}{\partial h} \, H \right)
-\: ,
+\Delta g_{P}^{F} 
+= g_{P} - \left( \gamma_{Q} + \frac{\partial \gamma}{\partial h} \, H_{P} \right) \: ,
 \label{eq:free-air-anomaly}
@@
 derivative of the normal gravity with respect to the geometric height $h$
-and $H$ is the orthometric height (Fig. \ref{fig:surfaces}). 
+and $H_{P}$ is the orthometric height $H$ (Fig. \ref{fig:surfaces}) at the
+point $P$.
 Notice that the term between parenthesis represents an 
@@
 the Earth's surface, and the Free-air anomaly
-$\Delta g_{P}^{a}$ (eq. \ref{eq:free-air-anomaly}), is given by:
+$\Delta g_{P}^{F}$ (eq. \ref{eq:free-air-anomaly}), is given by:
 \begin{equation}
-\vert \delta g_{P} - \Delta g^{a}_{P} \vert \approx 
+\vert \delta g_{P} - \Delta g^{F}_{P} \vert \approx 
 \vert \frac{\partial \gamma}{\partial h} N \vert \: ,
@@
 \end{equation}
-where the geoidal undulation $N \approx h - H$ (Fig. \ref{fig:surfaces}) assumes 
-that the Geoid and the surface of the reference ellipsoid 
-are approximately parallel at $P$ and the surrounding area.
+where $N \approx h - H$ is the geoidal undulation (Fig. \ref{fig:surfaces}).
+This approximation assumes that the Geoid and the surface of 
+the reference ellipsoid are approximately parallel at $P$ 
+and the surrounding area.
 We know empirically that the geoidal undulation $N$ in the world 
-is $\approx \pm 1$ on the oceans and reaches a 
+is $\approx \pm 1$ m on the oceans and reaches a 
 maximum absolute value of $\approx 120$ m \citep[e.g.,][]{torge2012, 
@@
 is $\approx 9.258$ mGal.
-
 The approximations defined by eqs \ref{eq:gravity-disturbance-approx}
@@
 geodesy} \citep{hofmann-wellenhof-moritz2005}.
-A similar approach can be used to show that the Bouguer anomaly
-is an approximation of the terrain corrected gravity disturbance.
-The Bouguer anomaly is commonly used by geophysicists as the
+
+The simple Bouguer anomaly is commonly used by geophysicists as the
 gravitational effect produced by the gravity sources.
+It is defined, over the continents, as follows 
+\citep{blakely1996, hofmann-wellenhof-moritz2005}:
+\begin{equation}
+\Delta g_{P}^{B} 
+= \Delta g_{P}^{F} - 2 \pi G \rho_{t} \, H_{P} \: ,
+\label{eq:simple-bouguer-anomaly}
+\end{equation}
+where $\Delta g_{P}^{F}$ is the Free-air anomaly (eq. \ref{eq:free-air-anomaly})
+and the last term on the right side is the simple Bouguer correction.
+This term approximates the gravitational attraction that all topographic 
+masses above the Geoid exerts at the point $P$ by the gravitational 
+attraction of a homogeneous, infinitely extended slab of 
+constant density $\rho_{t}$ and thickness equal to $H_{P}$, which is 
+the orthometric height $H$ (Fig. \ref{fig:surfaces}) at the point $P$.
+By computing the simple Bouguer correction at a set of points
+of a survey around $P$, we approximately remove, 
+from the gravity measured at the points of the survey,
+the gravitational effect of a homogeneous model representing 
+the topographic masses above the Geoid.
+It is evident from eqs \ref{eq:free-air-anomaly}, \ref{eq:gravity-disturbance-approx}
+and \ref{eq:difference-disturbance-free-air} that the simple Bouguer
+anomaly $\Delta g_{P}^{B}$ (eq. \ref{eq:simple-bouguer-anomaly}) represents
+the gravity disturbance $\delta g_{P}$ (eq. \ref{eq:gravity-disturbance})
+minus the gravitational effect produced by this homogeneous topographic model.
 Although this approximation is valid for most practical applications,
@@
 From eq. \ref{eq:gravity-disturbance-vector}, the gravity vector $\mitbf{g}_{i}$, at a point 
-$(x_{i}, y_{i}, z_{i})$ in the TCS (Fig. \ref{fig:TCS}b), can be represented by:
+$(x_{i}, y_{i}, z_{i})$ in the TCS (Fig. \ref{fig:TCS}a), can be represented by:
 \begin{equation}
@@
 $\mitbf{\gamma}_{i}$ at the point $(x_{i}, y_{i}, z_{i})$, in the
-topocentric Cartesian coordinate system (Fig. \ref{fig:TCS}b), and
+topocentric Cartesian coordinate system (Fig. \ref{fig:TCS}a), and
 $\hat{\mitbf{u}}_{i}$ is the unit vector $\hat{\mitbf{u}}$ 
@@
 is met at all points located above or on the Earth's surface.
-This approximation, which is known in geodesy \citep[e.g.,][]{sanso_sideris2013},
-is largely used in applied geophysics for representing
-total-field anomalies \citep[e.g.,][]{blakely1996}.
+This approximation is known in geodesy \citep[e.g.,][]{sanso_sideris2013}
+and a similar approximation is largely used in applied geophysics 
+for representing total-field anomalies \citep[e.g.,][]{blakely1996}.
 Notice that, in local- or regional-gravity studies, the unit
@@
 may be considered constant throughout the study area and 
-parallel to the $z$ axis of the TCS (Fig. \ref{fig:TCS}b).
+parallel to the $z$ axis of the TCS (Fig. \ref{fig:TCS}a).
 By using the approximation defined in eq. 
@@
 the same direction as the normal gravity vector $\mitbf{\gamma}_{P}$
-at the origin $P$ of the TCS (Fig. \ref{fig:TCS}b).
+at the origin $P$ of the TCS (Fig. \ref{fig:TCS}a).
 This equation shows that the gravity disturbance $\delta g_{i}$ (eq. \ref{eq:gravity-disturbance}) is different from the magnitude
@@
 
-In the TCS (Fig. \ref{fig:TCS}b),
+In the TCS (Fig. \ref{fig:TCS}a),
 the gravity disturbance $\delta g_{i}$ (eq. \ref{eq:gravity-disturbance-approx-TCS})
@@
 \phi(x,y,z) = \int\int\limits_{v}\int 
-\frac{c_{g} \, G \, \Delta\rho(x^{\prime}, y^{\prime}, z^{\prime}) \: dv}
+\frac{c_{g} \, G \, \Delta\rho(x^{\prime}, y^{\prime}, z^{\prime}) \: dv^{\prime}}
 {\sqrt{(x - x^{\prime})^{2} + 
@@
 with respect to the variable $z$. Is this integral, 
-$G$ is the Newtonian constant of gravitation (in $m^{3} kg^{-1} s^{-2}$),
-$c_{g} = 10^{5}$ is a constant factor transforming from $m s^{-2}$ 
+$G$ is the Newtonian constant of gravitation (in $m^{3} \, kg^{-1} \, s^{-2}$),
+$c_{g} = 10^{5}$ is a constant factor transforming from $m \, s^{-2}$ 
 to milligal (mGal), $\Delta\rho(x^{\prime}, y^{\prime}, z^{\prime})$
-is the density contrast (in $kg / m^{3}$) at a point 
+is the density contrast (in $kg \, m^{-3}$) at a point 
 $(x^{\prime}, y^{\prime}, z^{\prime})$ within 
@@
 of multiple gravity sources.
-
-Practically all the literature about gravity modelling
+Practically, all articles accomplishing gravity modelling
 use the quantity $d_{i}$ (eq. \ref{eq:gz-local})
@@
 of the observation points $(x_{i}, y_{i}, z_{i})$ with respect to
-the gravity sources in the TCS (Fig. \ref{fig:TCS}b).
-It is important to notice that, contrary to what is written in 
-some basic texts, gravimeters do not measure the quantity
-$d_{i}$ (eq. \ref{eq:gz-local}).
+the gravity sources in the TCS (Fig. \ref{fig:TCS}a).
+Finally, it is important to notice that, contrary to what 
+is written in some basic texts, gravimeters do not measure 
+the quantity $d_{i}$ (eq. \ref{eq:gz-local}).
 
@@
 on the Geoid, which is generally within the topographic masses.
-Hence, almost all these forward modelling techniques in fact
-compute the gravity disturbance.
+Hence, almost all these techniques in fact compute the gravity 
+disturbance.
 
@@
     \includegraphics{figures/local-system.png}
-    \caption{(a) Schematic representation of the gravity vector
+    \caption{(a) Schematic representation of a topocentric Cartesian 
+    coordinate system (TCS) with origin at a point $P$. The axes $x$
+    and $y$ are parallel to the unit vectors $\hat{\mitbf{v}}_{P}$
+    and $\hat{\mitbf{w}}_{P}$ (eq. \ref{eq:unit-vectors} and
+    Fig. \ref{fig:GCS-GGS}), respectively. On the other hand, the $z$ axis is opposite to the unit vector $\hat{\mitbf{u}}_{P}$ (eq. 
+    \ref{eq:unit-vectors} and Fig. \ref{fig:GCS-GGS}) and points downward. The gray plane is the same shown in Fig. \ref{fig:GCS-GGS}.
+    (b) Schematic representation of the gravity vector
     $\mitbf{g}_{P}$, normal gravity vector $\mitbf{\gamma}_{P}$,
@@
     $\hat{\mitbf{u}}_{P}$ (eq. \ref{eq:unit-vectors}) at a point 
-    $P$.
-    (b) Schematic representation of a topocentric Cartesian 
-    coordinate system with origin at a point $P$. The axes $x$
-    and $y$ are parallel to the unit vectors $\hat{\mitbf{v}}_{P}$
-    and $\hat{\mitbf{w}}_{P}$ (eq. \ref{eq:unit-vectors} and
-    Fig. \ref{fig:GCS-GGS}), respectively. On the other hand, the $z$ axis is opposite to the unit vector $\hat{\mitbf{u}}_{P}$ (eq. 
-    \ref{eq:unit-vectors} and Fig. \ref{fig:GCS-GGS}) and points downward. The gray plane shown in (b) is the same gray plane
-    shown in Fig. \ref{fig:GCS-GGS}.}
+    $P$.}
   \label{fig:TCS}

### 2018-05-06 manuscript/letter.tex
review the summary and other small stuff

@@
 \begin{summary}
- The gravity anomaly has long been used by geophysicists for 
- representing the gravitational effect produced by gravity
- sources.This quantity, however, can be used in this way as a good
- approximation. Rigorously, the gravitational effect produced by the sources
- is represented by the gravity disturbance, which is a well-known 
- quantity in geodesy.
- In this paper, we discuss the fundamental concepts supporting 
- the fact that gravity disturbance is the more appropriated quantity
- to be used for estimating density variations related to geological
- bodies in subsurface. We show that the maximum absolute difference 
- between the gravity disturbance and the Free-air anomaly in Brazil 
- reaches a maximum absolute value of $\approx 9.3$ mGal. We also stress 
- that almost all authors in the literature have implicitly used the 
- gravity anomaly as an approximation of the gravity disturbance, 
- which can be represented by a harmonic function and, consequently, 
- satisfies the premise behind most of 
- the interpretation techniques used in geophysics (e.g., 
- upward/downward continuation, data processing with equivalent layer,
- conversions between gravity and magnetic data, computation of vertical 
- derivatives via Fourier and Hilbert transforms).
- Although the gravity anomaly can be considered a good approximation of the 
- gravity disturbance for most practical applications, 
- it is important to bear in mind not only the terminology 
- changes, but also the conceptual assumptions.
+ In applied geophysics, the gravity anomaly has long been used for representing the
+ gravitational effect produced by gravity sources, which are defined as the density
+ contrasts between the internal density distribution of the Earth and the normal Earth.
+ The gravity anomaly is defined by the difference between the gravity (produced by the
+ Earth) on the Geoid and the normal gravity (produced by the normal Earth)
+ on reference ellipsoid, both at the same geodetic latitude and longitude.
+ Because these quantities are not defined at the same point, 
+ the gravity anomaly does not represent the gravitational effect due to the gravity
+ sources and, consequently, cannot be considered a harmonic function.
+ On the contrary, the gravity disturbance is defined by the difference between
+ the gravity and normal gravity at the same point. Consequently, the centrifugal 
+ effects can be neglected, the gravity disturbance can be considered a harmonic 
+ function representing the gravitational effect due to the gravity sources and then 
+ it satisfies the premise behind most of the processing techniques of potential-field
+ data (e.g., upward/downward continuations). Then, rigorously, the gravity disturbance
+ is  more appropriated to be used for estimating density variations in subsurface.
+ The use the gravity anomaly for most of the geophysicists carries the
+ implicit meaning that it can be considered a good approximation of the gravity
+ disturbance. It is important to bear in mind, however, that in Brazil, for example, 
+ the maximum absolute difference between the gravity disturbance (on or close 
+ to the Earth’s surface) and the Free-air anomaly reaches 9.3 mGal.
 
@@
 as possible \citep{vanicek1987}.
-
 It is worth noting that, although the normal Earth has the
@@
 \begin{equation}
-d_{i} = \partial_{z} \phi_{i} \: ,
+d_{i} = \int\int\limits_{v}\int 
+\frac{c_{g} \, G \, \Delta\rho(x^{\prime}, y^{\prime}, z^{\prime}) 
+\, (z_{i} - z^{\prime}) \: dv^{\prime}}
+{\sqrt{(x_{i} - x^{\prime})^{2} + 
+(y_{i} - y^{\prime})^{2} + (z_{i} - z^{\prime})^{2}}} \: ,
 \label{eq:gz-local}
 \end{equation}
-where $\partial_{z} \phi_{i}$ is the first 
-derivative, evaluated at the point $(x_{i},y_{i},z_{i})$, of the function 
-\begin{equation}
-\phi(x,y,z) = \int\int\limits_{v}\int 
-\frac{c_{g} \, G \, \Delta\rho(x^{\prime}, y^{\prime}, z^{\prime}) \: dv^{\prime}}
-{\sqrt{(x - x^{\prime})^{2} + 
-(y - y^{\prime})^{2} + (z - z^{\prime})^{2}}} \: ,
-\label{eq:phi}
-\end{equation}
-with respect to the variable $z$. Is this integral, 
-$G$ is the Newtonian constant of gravitation (in $m^{3} \, kg^{-1} \, s^{-2}$),
+where $G$ is the Newtonian constant of gravitation (in $m^{3} \, kg^{-1} \, s^{-2}$),
 $c_{g} = 10^{5}$ is a constant factor transforming from $m \, s^{-2}$ 

### 2018-05-10 manuscript/letter.tex
Review of the introduction with slight changes
I reworded some parts of the introduction to make it read more smoothly.
Removed some repetition and added more idioms.

--- 2018-04-27 manuscript/letter.tex
@@
 $^1$ Department of Geophysics, Observat\'{o}rio Nacional, Rio de Janeiro, Brazil \\
-$^2$ Department of Geology and Geophysics, University of Hawaii, Manoa, USA
+$^2$ Department of Geology and Geophysics, SOEST, University of Hawai'i at M\={a}noa, Honolulu, USA
 }
@@
  representing the gravitational effect produced by gravity
- sources.This quantity, however, can be used in this way as a good
+ sources. This quantity, however, can be used in this way as a good
  approximation. Rigorously, the gravitational effect produced by the sources
@@
 
-The resultant of gravitational force and centrifugal force acting 
-on a body at rest on the Earth's surface is called 
-\textit{gravity vector} and its intensity is called 
-\textit{gravity} \citep{heiskanen-moritz1967,
-hofmann-wellenhof-moritz2005}.
-In the case of gravimetry on moving platforms (e.g., airplanes,
+The \textit{gravity vector} is the sum of the gravitational and centrifugal
+accelerations felt by a body at rest on the Earth's surface.
+Its intensity is what is geoscientists call \textit{gravity}
+\citep{heiskanen-moritz1967, hofmann-wellenhof-moritz2005}.
+When measured on moving platforms (e.g., airplanes,
 helicopters, marine vessels), there are additional
@@
 \citep{glennie-etal2000,nabighian-etal2005-grav,baumann-etal2012}.
-Geophysicists use gravity for estimating the Earth's 
-internal density distribution whereas geodesists use
-gravity to estimate the Geoid \citep{li2001}.
-Hence, geophysicists are usually interested 
-in the gravitational component of the observed gravity, 
-which is produced by the Earth's internal 
-density distribution.
+Geodesists use gravity measurements to estimate the Geoid \citep{li2001}
+whereas geophysicists use gravity to estimate the Earth's
+internal density distribution.
+Hence, geophysicists are usually only interested
+in the gravitational component of the observed gravity
+because it is the part that reflects the Earth's internal density distribution.
 
-The first step of the procedure for isolating this 
-gravitational component consists in removing the non-gravitational 
-effects due to the vehicle motion and also the time variations 
-such as Earth tides, instrumental drift and barometric 
-pressure changes, for example.
+The first step toward isolating the gravitational component
+is to remove the effects of vehicle motion, Earth tides, instrumental drift,
+and barometric pressure changes, among others.
 If these effects are properly removed, the resultant
-gravity data can be considered as the sum of a 
-centrifugal component due to the Earth's rotation and
-a gravitational component produced by the whole Earth's
-internal density distribution.
+observations can be considered to be solely the sum of the
+centrifugal acceleration due to the Earth's rotation and
+the gravitational acceleration produced by
+the internal density distribution of the whole Earth.
 The isolation of this particular gravitational component
@@
 distributions related to geological structures in subsurface
-are the main goals in applied geophysics \citep{blakely1996}.
+are the main goals of applied gravimetry \citep{blakely1996}.
 The computation of the gravitational effect produced by
@@
 This term is different from \textit{gravity field modelling}, which
-is commonly used in geodesy to denote the characterization of the 
-gravity field in a local, regional or global scale.
+is commonly used in geodesy to denote the mathematical characterization of the
+gravity field in local, regional, or global scales (e.g., spherical harmonic
+analysis).
 
-Based on well-established concepts of the literature,
-we present a discussion aiming at bringing some
-light to the following question: in geophysical applications,
-should we use the gravity disturbance or gravity anomaly?
-It seems that this theoretical issue has been 
-debated within the scientific community from a 
-more geodetic than geophysical point of view
+We present a discussion with the aim of shedding some light on the question:
+\textit{should we use the gravity disturbance or gravity anomaly in geophysical
+applications}?
+The concepts discussed here are well-established in the literature.
+However, the debate around this theoretical issue has been
+carried about from a more geodetic than a geophysical point of view
 \citep{lafehr1991,chapin1996,li2001,fairhead2003,
 hackney-featherstone2003,hinze2005}.
-Our reasoning suggests that the gravity 
-disturbance is more appropriated than the gravity anomaly 
-for geophysical purposes.
+Our reasoning suggests that the gravity disturbance is more appropriated for
+geophysical purposes than the gravity anomaly.
 
@@
 on the Geoid, which is generally within the topographic masses.
-Hence, almost all these forward modelling techniques in fact
-compute the gravity disturbance.
+Hence, almost all these techniques in fact compute the gravity
+disturbance.
 

### 2018-05-10 manuscript/letter.tex
Some more changes to the introduction
Some more rewording of a few phases.

@@
 helicopters, marine vessels), there are additional
-non-gravitational accelerations due to the vehicle motion,
-such as Coriolis acceleration and high-frequency vibrations
+non-gravitational accelerations due to the motion of the vehicle,
+such as the Coriolis acceleration and high-frequency vibrations
 \citep{glennie-etal2000,nabighian-etal2005-grav,baumann-etal2012}.
@@
 in the gravitational component of the observed gravity
-because it is the part that reflects the Earth's internal density distribution.
+because that is what that reflects the Earth's internal density distribution.
 
@@
 and barometric pressure changes, among others.
-If these effects are properly removed, the resultant
-observations can be considered to be solely the sum of the
-centrifugal acceleration due to the Earth's rotation and
+If these effects are properly removed,
+the resultant observations are considered to be solely
+the sum of the centrifugal acceleration due to the Earth's rotation and
 the gravitational acceleration produced by
 the internal density distribution of the whole Earth.
-The isolation of this particular gravitational component
+The isolation of this particular gravitational component,
 and its subsequent use for estimating density
-distributions related to geological structures in subsurface
+distributions related to geological structures in subsurface,
 are the main goals of applied gravimetry \citep{blakely1996}.
 The computation of the gravitational effect produced by
-the geological structures is commonly known in geophysics as
+the geological structures is known in geophysics as
 \textit{gravity modelling}.
-This term is different from \textit{gravity field modelling}, which
+Not to be confused with \textit{gravity field modelling}, which
 is commonly used in geodesy to denote the mathematical characterization of the

### 2018-05-10 manuscript/letter.tex
review conclusion

--- 2018-05-06 manuscript/letter.tex
@@
 
-We debate the conceptual differences between the gravity
-disturbance and the gravity anomaly.
-Our reasoning suggests that the gravity disturbance is the
-more appropriated quantity for representing the gravity
-effect produced by the gravity sources.
-In summary, we point out that:
+Although the gravity disturbance is a well-known quantity in geodesy,
+its use in the applied geophysics is uncommon. 
+For years, geophysicists have been using gravity anomaly for modelling
+the gravitational effects due to geological bodies.
+The question "Should geophysicists use gravity anomaly or gravity
+disturbance for modelling the gravitational effect of gravity sources?"
+is of more than academic interest.
 
-\begin{enumerate}
-%\renewcommand{\theenumi}{(\arabic{enumi})}
-\renewcommand{\theenumii}{\alph{enumii}}
+Theoretically, the gravity anomaly is the difference between the
+gravity on the Geoid and the normal gravity on the reference ellipsoid,
+both at the same geodetic latitude and longitude.
+On the other hand, the gravity disturbance is defined by the difference
+between gravity and normal gravity at the same point.
+Because these quantities are at the same point, the gravity disturbance
+is a purely gravitational effect due to the gravity sources whereas the
+gravity anomaly is a combination of gravitational and centrifugal
+effects.
+Consequently, the gravity disturbance is a harmonic function whereas the
+gravity anomaly is not.
+Because the gravity disturbance is a purely gravitational effect, it can
+be represented by a harmonic function and satisfies the premise required
+by the most techniques for processing potential-field data (e.g.,
+upward/downward continuation, data processing with equivalent layer,
+conversions between gravity and magnetic data, computation of vertical
+derivatives via Fourier and Hilbert transforms).
+Hence, the gravity disturbance is the more appropriated quantity for
+representing the gravity effect produced by the gravity sources.
 
-\item Almost all interpretation techniques assume, implicitly or
-directly, that the gravity data is harmonic (e.g., 
-upward/downward continuation, 
-data processing with equivalent layer,
-conversions between gravity and magnetic data,
-computation of vertical derivatives via Fourier and Hilbert
-transforms). As a consequence, they implicitly or directly assume that
-the gravity data approximate the gravity disturbance.
+In practical problems of applied geophysics, geophysicists perform the
+gravity-forward modelling by computing the vertical component of the
+gravitational attraction exerted by the sources at the observation
+points, regardless of the Geoid or reference ellipsoid positions,
+for example.
+This means that, actually, they compute gravity disturbance to fit 
+a gravity anomaly (e.g., Bouguer anomaly) and this is a clear inconsistency.
+Moreover, upward or downward continuations, for example, are commonly applied to
+gravity anomalies. In this case, the gravity anomaly is wrongly considered a
+harmonic function and this is another inconsistency.
 
-\item Almost all forward modelling techniques compute
-the vertical component of the gravitational attraction 
-exerted by the gravity sources at the observation points.
-Notice that the gravity anomaly requires the computation of gravity
-on the Geoid, which is generally within the topographic masses.
-Hence, almost all these techniques in fact compute the gravity 
-disturbance.
-
-\item The gravity anomaly is defined as the difference between 
-the gravity on the Geoid and the normal gravity on the reference 
-ellipsoid, both at the same geodetic latitude and longitude.
-Consequently, the gravity anomaly is a function of
-the geodetic latitude and longitude only and cannot
-be calculated at arbitrary heights. On the other hand,
-the gravity disturbance can be computed at arbitrary points
-outside the sources. Then, computing the upward 
-continuation of gravity anomalies implicitly assumes 
-that they approximate the gravity disturbance.
-
-\item In Brazil, the maximum difference between the gravity 
-disturbance (on or close to the Earth's surface) and the Free-air
-anomaly reaches $\approx 9.3$ mGal.
-
-\end{enumerate}
-
-The principal theoretical implication of this study is that,
-although the gravity anomaly may be used as a good approximation of
-the gravitational effect produced by the sources for most practical applications,
-the more appropriated quantity for gravity modelling is the
-gravity disturbance. We stress that, more important than the 
-terminology changes, the researchers must bear in mind the 
-conceptual assumptions used in gravity modelling.
+By adopting these practical procedures, geophysicists are implicitly
+taking the gravity anomaly as a good approximation of the gravity
+disturbance. But why should geophysicists care about this usual
+procedure? Because it may result in wrong interpretations of
+quantitative nature. In Brazil, for example, the maximum difference
+between the gravity disturbance and the free-air anomaly reaches 9.3
+mGal. Such difference on the Earth's surface may lead, for example, to
+an error of about xxx m in the depth-to-center estimate of a sphere-like
+source with true depth-to-center at xxx m, radius xxxx m and density
+contrast of xxx $kg \, m^{-3}$.
 
@@
 
-\end{document}
+\end{document}\grid

### 2018-05-10 manuscript/letter.tex
small changes suggested by Val

@@
 
-In a local or regional study, geophysicists commonly use a topocentric 
+In local or regional studies, geophysicists commonly use a topocentric 
 Cartesian coordinate system (TCS) with origin at a point $P$ on or 
@@
 the gravity disturbance vector $\delta \mitbf{g}_{P}$ represents 
-a purely gravitational (and consequently harmonic) quantity, which is 
-caused by contrasts between the actual internal 
+a purely gravitational effect caused by contrasts between the actual internal 
 density distribution of the Earth and the internal density 
 distribution of the normal Earth.
+ergo, the gravity disturbance vector $\delta \mitbf{g}_{P}$ 
+(eq. \ref{eq:gravity-disturbance-vector}) is harmonic.
 It is important to stress that both density distributions are unknown.
@@
 $\delta \mitbf{g}_{P}$ \citep{barthelmes2013, sanso_sideris2013}.
-Another anomalous quantity is the \textit{gravity anomaly vector}.
-It is defined, at point $P$, as the difference between the gravity
+
+On the other hand, the \textit{gravity anomaly vector} at $P$ 
+is the difference between the gravity
 vector at a point $Q^{\prime}$ on the Geoid (a particular
@@
 on the reference ellipsoid, at the point $Q$ (Fig. \ref{fig:surfaces}).
-Notice that, by definition, the gravity anomaly depends on 
-longitude and latitude only and is not a function of height 
-\citep{barthelmes2013}. As a consequence, it is not possible to compute,
-for example, the upward continuation of a gravity anomaly.
+As properly pointed out by \citet{barthelmes2013} the gravity anomaly,
+by definition, depends on 
+longitude and latitude only and is not a function of height. 
+Because the gravity is on the Geoid and the normal gravity 
+is on reference ellipsoid, the gravity anomaly $\Delta g_{P}$ 
+(eq. \ref{eq:gravity-anomaly}) is not a purely gravitational effect 
+due to the gravity sources, but rather a combination of gravitational 
+and centrifugal effects. Consequently, its upward continuation, 
+for example, could not be rigorously done.
+
 However, many authors in the literature have computed the upward 
@@
 the gravity disturbance, which in turn can be represented by a 
-harmonic function that is allowed to be upward continued.
+harmonic function and can then being upward continued.
 
@@
 the topographic masses above the Geoid.
-It is evident from eqs \ref{eq:free-air-anomaly}, \ref{eq:gravity-disturbance-approx}
-and \ref{eq:difference-disturbance-free-air} that the simple Bouguer
-anomaly $\Delta g_{P}^{B}$ (eq. \ref{eq:simple-bouguer-anomaly}) represents
-the gravity disturbance $\delta g_{P}$ (eq. \ref{eq:gravity-disturbance})
-minus the gravitational effect produced by this homogeneous topographic model.
+It is evident from eqs 
+\ref{eq:free-air-anomaly} to \ref{eq:difference-disturbance-free-air} 
+that the simple Bouguer anomaly $\Delta g_{P}^{B}$ (eq.
+\ref{eq:simple-bouguer-anomaly}) approximates the gravity disturbance
+$\delta g_{P}$ (eq. \ref{eq:gravity-disturbance})
+minus the gravitational effect produced by this homogeneous topographic
+model.
 Although this approximation is valid for most practical applications,
@@
 evaluated on the ellipsoid from the gravity measured 
-on the Earth's surface, the remaining disturbing field will reflect 
+on the Earth's surface, the remaining anomalous field will reflect 
 not only the effect produced by the gravity sources, but also a

### 2018-05-11 manuscript/letter.tex

@@
 
-\author[Hallam et al.]
-{Kristoffer A. T. Hallam$^{1}$\thanks{Emails: kristoffer.hallam@gmail.com,
-vanderlei@on.br}, Vanderlei C. Oliveira Jr$^1$,
-Val\'{e}ria C. F. Barbosa$^1$ \linebreak and Leonardo Uieda$^2$ \\
+\author[Oliveira Jr et al.]
+{Vanderlei C. Oliveira Jr$^1$\thanks{Email: vanderlei@on.br},
+Kristoffer A. T. Hallam$^{1}$, Val\'{e}ria C. F. Barbosa$^1$
+and Leonardo Uieda$^2$ \\
 $^1$ Department of Geophysics, Observat\'{o}rio Nacional, Rio de Janeiro, Brazil \\
-$^2$ Department of Geology and Geophysics, University of Hawaii, Manoa, USA
+$^2$ Department of Geology and Geophysics, SOEST, University of Hawai'i at M\={a}noa, Honolulu, USA
 }
@@
 
-The resultant of gravitational force and centrifugal force acting 
-on a body at rest on the Earth's surface is called 
-\textit{gravity vector} and its intensity is called 
-\textit{gravity} \citep{heiskanen-moritz1967,
-hofmann-wellenhof-moritz2005}.
-In the case of gravimetry on moving platforms (e.g., airplanes,
+The \textit{gravity vector} is the sum of the gravitational and centrifugal
+accelerations felt by a body at rest on the Earth's surface.
+Its intensity is what is geoscientists call \textit{gravity}
+\citep{heiskanen-moritz1967, hofmann-wellenhof-moritz2005}.
+When measured on moving platforms (e.g., airplanes,
 helicopters, marine vessels), there are additional
-non-gravitational accelerations due to the vehicle motion, 
-such as Coriolis acceleration and high-frequency vibrations 
+non-gravitational accelerations due to the motion of the vehicle,
+such as the Coriolis acceleration and high-frequency vibrations
 \citep{glennie-etal2000,nabighian-etal2005-grav,baumann-etal2012}.
-Geophysicists use gravity for estimating the Earth's 
-internal density distribution whereas geodesists use
-gravity to estimate the Geoid \citep{li2001}.
-Hence, geophysicists are usually interested 
-in the gravitational component of the observed gravity, 
-which is produced by the Earth's internal 
-density distribution.
+Geodesists use gravity measurements to estimate the Geoid \citep{li2001}
+whereas geophysicists use gravity to estimate the Earth's
+internal density distribution.
+Hence, geophysicists are usually only interested
+in the gravitational component of the observed gravity
+because that is what that reflects the Earth's internal density distribution.
 
-For isolating this 
-gravitational component consists in removing the non-gravitational 
-effects due to the vehicle motion and also the time variations 
-such as Earth tides, instrumental drift and barometric 
-pressure changes, for example.
-If these effects are properly removed, the resultant 
-gravity data can be considered as the sum of a 
-centrifugal component due to the Earth's rotation and
-a gravitational component produced by the whole Earth's
-internal density distribution.
-The isolation of this particular gravitational component 
+The first step toward isolating the gravitational component
+is to remove the effects of vehicle motion, Earth tides, instrumental drift,
+and barometric pressure changes, among others.
+If these effects are properly removed,
+the resultant observations are considered to be solely
+the sum of the centrifugal acceleration due to the Earth's rotation and
+the gravitational acceleration produced by
+the internal density distribution of the whole Earth.
+The isolation of this particular gravitational component,
 and its subsequent use for estimating density
-distributions related to geological structures in subsurface 
-are the main goals in applied geophysics \citep{blakely1996}.
+distributions related to geological structures in subsurface,
+are the main goals of applied gravimetry \citep{blakely1996}.
 The computation of the gravitational effect produced by
-the geological structures is commonly known in geophysics as 
+the geological structures is known in geophysics as
 \textit{gravity modelling}.
-This term is different from \textit{gravity field modelling}, which
-is commonly used in geodesy to denote the characterization of the 
-gravity field in a local, regional or global scales.
+Not to be confused with \textit{gravity field modelling}, which
+is commonly used in geodesy to denote the mathematical characterization of the
+gravity field in local, regional, or global scales (e.g., spherical harmonic
+analysis).
 
-Based on well-established concepts of the literature,
-we present a discussion aiming at critically examining
-the following question: in geophysical applications,
-should we use the gravity disturbance or gravity anomaly?
-It seems that this theoretical issue has been 
-debated within the scientific community from a 
-more geodetic than geophysical point of view
+We present a discussion with the aim of shedding some light on the question:
+\textit{should we use the gravity disturbance or gravity anomaly in geophysical
+applications}?
+The concepts discussed here are well-established in the literature.
+However, the debate around this theoretical issue has been
+carried about from a more geodetic than a geophysical point of view
 \citep{lafehr1991,chapin1996,li2001,fairhead2003,
 hackney-featherstone2003,hinze2005}.
-Our reasoning suggests that the gravity 
-disturbance is more appropriated than the gravity anomaly 
-for geophysical purposes.
+Our reasoning suggests that the gravity disturbance is more appropriated for
+geophysical purposes than the gravity anomaly.
 
@@
 distribution of the normal Earth.
-ergo, the gravity disturbance vector $\delta \mitbf{g}_{P}$ 
-(eq. \ref{eq:gravity-disturbance-vector}) is harmonic.
-It is important to stress that both density distributions are unknown.
 In applied geophysics, these density differences are generally 
@@
 and can be represented as follows:
+
 \begin{equation}
@@
 \end{equation}
+
 Notice that the gravity disturbance $\delta g_{P}$ is not equivalent 
@@
 
-\end{document}\grid
+\end{document}

### 2018-05-11 manuscript/letter.tex
Fix author line too long using a parbox

@@
 
-\author[Oliveira Jr et al.]
-{Vanderlei C. Oliveira Jr$^1$\thanks{Email: vanderlei@on.br},
-Kristoffer A. T. Hallam$^{1}$, Val\'{e}ria C. F. Barbosa$^1$
-and Leonardo Uieda$^2$ \\
+\author[Hallam et al.]{
+\parbox{\linewidth}{%
+    Kristoffer A. T. Hallam$^{1}$\thanks{Emails: kristoffer.hallam@gmail.com, vanderlei@on.br},
+    Vanderlei C. Oliveira Jr$^1$,
+    Val\'{e}ria C. F. Barbosa$^1$, and \linebreak Leonardo Uieda$^2$
+    \vspace{0.3cm}
+}%
+    \\
     $^1$ Department of Geophysics, Observat\'{o}rio Nacional, Rio de Janeiro, Brazil \\

### 2018-05-11 manuscript/letter.tex
Revised Normal Earth section
I reordered some of the phrases to have the fact that the internal
density is not specified setup the section about geodesy not caring
about that.

Removed the part about "normal density" because it's never mentioned
again in the text.

@@
 
-Traditionally, the Earth's gravity field is approximated
-by the gravity field produced by a reference ellipsoid
-(or level ellipsoid), which is a rigid and geocentric model.
-This reference ellipsoid has the minor axis $b$
-coincident with the mean rotating axis of the Earth $Z$, the
-same total mass (including the atmosphere) and also the
-same angular velocity of the Earth \citep{heiskanen-moritz1967,
+The Earth's gravity field is traditionally approximated
+by the gravity field of a reference ellipsoid (or level ellipsoid).
+This model is rigid and geocentric,
+with a minor axis $b$ which coincides with
+the mean rotation axis of the Earth $Z$.
+The ellipsoid has the same total mass as the Earth (including the atmosphere)
+and also the same angular velocity \citep{heiskanen-moritz1967,
 vanicek1987,hofmann-wellenhof-moritz2005,torge2012}.
-Another characteristic of this model is that its
-limiting surface coincides with a particular equipotential
-of its own gravity field.
-Here, we follow \citep{torge2012} and call this model as
-\textit{normal Earth}.
+Its limiting surface coincides with
+a particular equipotential of its own gravity field.
+Here, we follow \citep{torge2012} and call this reference ellipsoid
+the \textit{normal Earth}.
+Analogously to the Earth, the \textit{normal gravity vector} is
+the sum of the
+gravitational and centrifugal accelerations exerted by the normal
+Earth on a body at rest at a point $P$.
+The intensity of the normal gravity vector is called \textit{normal gravity}.
 
-The resultant of the virtual
-gravitational and centrifugal forces exerted by the normal
-Earth on a body at rest at a point $P$ is called
-\textit{normal gravity vector} and its intensity is called
-\textit{normal gravity}.
-In geodesy, any model used to represent the normal gravity field
-can be arbitrarily defined for the only purpose
-of keeping the difference from the actual gravity field as small
-as possible \citep{vanicek1987}.
-It is worth noting that, although the normal Earth has the
-same total mass (including the atmosphere) of the Earth,
-its internal density distribution is unknown.
-The search for physically meaningful mass distributions
-that generate a required normal gravity field
-has geophysical rather than geodetic motives \citep{marussi1974}.
+It is worth noting that, although the normal Earth has the same total mass as
+the Earth, its internal density distribution is left unspecified.
+The search for geologically meaningful density distributions
+for the interior of the normal Earth has
+geophysical rather than geodetic motives \citep{marussi1974}.
+In physical geodesy, a model for the normal gravity field
+can be arbitrarily defined with the sole purpose of
+keeping the difference from the actual gravity field
+as small as possible \citep{vanicek1987}.
 The only condition imposed on its internal density
@@
 with its limiting surface.
-For convenience, we denote any density distribution
-satisfying this condition as a \textit{normal density distribution}.
 

### 2018-05-15 manuscript/letter.tex
Remove extra "is"
Reword the key question of the article
Include a phrase about how our approach will differ from the ones in the
literature.

@@
 accelerations felt by a body at rest on the Earth's surface.
-Its intensity is what is geoscientists call \textit{gravity}
+Its intensity is what geoscientists call \textit{gravity}
 \citep{heiskanen-moritz1967, hofmann-wellenhof-moritz2005}.
@@
 
-We present a discussion with the aim of shedding some light on the question:
-\textit{should we use the gravity disturbance or gravity anomaly in geophysical
-applications}?
+We present a discussion of whether the gravity disturbance or the gravity
+anomaly should be used in geophysical applications.
 The concepts discussed here are well-established in the literature.
@@
 hackney-featherstone2003,hinze2005}.
+We will approach the topic by comparing the geophysical interpretations of the
+disturbance and the anomaly.
 Our reasoning suggests that the gravity disturbance is more appropriated for

### 2018-05-16 manuscript/letter.tex

@@
 
-For geophysical purposes, there are three important Terrestrial Reference Systems used in gravity modelling.
+For geophysical gravity modelling purposes, there are three important
+Terrestrial Reference Systems.
 They rotate with the Earth and are used for describing
-positions and movements of objects on and close to the Earth’s surface
+positions and movements of objects on and close to the Earth's surface
 \citep{torge2012}.
 
-The first one is a geocentric system of Cartesian coordinates having
-the $Z$-axis coincident with the mean Earth's rotational axis,
-the $X$-axis pointing to the Greenwich meridian and the $Y$-axis
-directed so as to obtain a right-handed system (Fig. \ref{fig:GCS-GGS}).
-This reference system can be found in the literature with different names: Mean Terrestrial System \citep[e.g.,][]{soler1976}, Earth-fixed geocentric Cartesian system \citep[e.g.,][]{torge2012}
-or Earth-centered Earth-fixed system \citep[e.g.,][]{bouman_etal2013}, for example. Here, we opted for simply using the term Geocentric Cartesian System (GCS).
+The first is a geocentric system of Cartesian coordinates
+having the $Z$-axis coincident with the mean rotational axis of the Earth,
+the $X$-axis pointing to the Greenwich meridian,
+and the $Y$-axis completing a right-handed system (Fig. \ref{fig:GCS-GGS}).
+This reference system is called by different names in the literature:
+Mean Terrestrial System \citep[e.g.,][]{soler1976},
+Earth-fixed geocentric Cartesian system \citep[e.g.,][]{torge2012}
+or Earth-centered Earth-fixed system \citep[e.g.,][]{bouman_etal2013},
+for example.
+Here, we opted for using the term Geocentric Cartesian System (GCS).
 
-Another important reference system is a geocentric system of
-geodetic coordinates, which is defined by the reference ellipsoid
-used in the Normal Earth model \citep{heiskanen-moritz1967, soler1976,
-torge2012, bouman_etal2013}.
-In this coordinate system, the position of a point
-is defined by the \textit{geometric height} $h$,
-\textit{geodetic latitude} $\varphi$ and \textit{longitude} $\lambda$ (Fig. \ref{fig:GCS-GGS}).
-For convenience, we call this system Geocentric Geodetic System
-(GGS).
-At a given point $(h, \varphi, \lambda)$, there are three
-mutually-orthogonal unit vectors (Fig. \ref{fig:GCS-GGS}) given by \citep{soler1976}:
+The second is a geocentric system of geodetic coordinates
+(the Geocentric Geodetic System or GGS),
+which is defined by the reference ellipsoid used as the Normal Earth
+\citep{heiskanen-moritz1967, soler1976, torge2012, bouman_etal2013}.
+The position of a point is defined by
+the \textit{geometric height} $h$,
+the \textit{geodetic latitude} $\varphi$,
+and the \textit{longitude} $\lambda$ (Fig. \ref{fig:GCS-GGS}).
+There are also three mutually-orthogonal unit vectors defined at a given point
+\citep{soler1976}:
+
 \begin{equation}
@@
 \end{equation}
-The required equations to convert coordinates $(h, \varphi,
-\lambda)$ referred to the GGS into coordinates $(X, Y, Z)$ referred to
-the GCS (Fig. \ref{fig:GCS-GGS}) and vice versa can be easily found in
+
+\noindent Equations for converting between these systems can be found in
 the literature \citep[e.g.,][]{heiskanen-moritz1967, torge2012,
@@
 
-In local or regional studies, geophysicists commonly use a topocentric
-Cartesian coordinate system (TCS) with origin at a point $P$ on or
-close to the Earth's surface and axes $x$, $y$ and $z$ (Fig.
-\ref{fig:TCS}). In the TCS, the axes $x$ and $y$ are parallel to
-the unit vectors $\hat{\mitbf{v}}_{P}$ $\hat{\mitbf{w}}_{P}$, respectively,
-whereas the $z$-axis is opposite to the unit vector
-$\hat{\mitbf{u}}_{P}$ (eq. \ref{eq:unit-vectors}) and points downward.
-Consider, for example, a TCS (Fig. \ref{fig:TCS}) with origin at a point $P$ with
-coordinates $(X_{P}, Y_{P}, Z_{P})$ referred to the GCS (Fig. \ref{fig:GCS-GGS}).
-In this case, the relationship between the coordinates $(X, Y, Z)$ of a
-point in the GCS and the coordinates $(x, y, z)$ of the same point in the
-TCS is given by:
-\begin{equation}
-\begin{bmatrix}
-x \\
-y \\
-z
-\end{bmatrix} =
-\mitbf{R}^{\top} \begin{bmatrix}
-X - X_{P} \\
-Y - Y_{P} \\
-Z - Z_{P}
-\end{bmatrix} \: ,
-\label{eq:GCS2TCS}
-\end{equation}
-where $\mitbf{R}$ is a $3 \times 3$ orthogonal matrix whose
-first, second and third columns are defined, respectively,
-by the unit vectors $\hat{\mitbf{v}}$, $\hat{\mitbf{w}}$
-and $-\hat{\mitbf{u}}$ (eq. \ref{eq:unit-vectors}),
-evaluated at the point $P$.
+The third is the Topocentric Cartesian System (TCS),
+which is commonly used in local or regional scale geophysical studies.
+The origin of the TCS is located at a point $P$,
+axes $x$ and $y$ are parallel to
+the unit vectors $\hat{\mitbf{v}}$ and $\hat{\mitbf{w}}$
+(eq. \ref{eq:unit-vectors}), respectively,
+and the $z$-axis is opposite to the unit vector $\hat{\mitbf{u}}$
+(Fig. \ref{fig:TCS}).
 

### 2018-05-16 manuscript/letter.tex
Started working on section 4

@@
 
-Let $\mitbf{g}_{P}$ and $\mitbf{\gamma}_{P}$ be, respectively, the
-gravity vector (corrected from non-gravitational effects due to vehicle
-motion and time variations such as Earth tides, instrumental drift and
-barometric pressure changes, for instance) and the normal gravity vector
-at a point $P$.
-In this case, the gravity vector $\mitbf{g}_{P}$ represents the
-gradient of a scalar potential called \textit{gravity potential},
-which is the sum of a scalar gravitational potential and a
-scalar centrifugal potential.
-Similarly, the normal gravity vector $\mitbf{\gamma}_{P}$ represents the
-gradient of a scalar potential called \textit{normal potential},
-which is also the sum of a scalar gravitational potential and a
-scalar centrifugal potential.
-By definition, the centrifugal part of the normal potential is equal
-to that of the gravity potential.
+Let $\mitbf{\gamma}_{P}$ and $\mitbf{g}_{P}$ and be, respectively,
+the \textit{normal gravity vector} and the Earth's \textit{gravity vector}
+(corrected from non-gravitational effects due to vehicle
+motion and time variations such as Earth tides and instrumental drift),
+both located at a point $P$.
+In this case, the gravity vector $\mitbf{g}_{P}$ is the
+gradient of the scalar \textit{gravity potential},
+which is the sum of a gravitational potential and a centrifugal potential.
+Similarly, the normal gravity vector $\mitbf{\gamma}_{P}$ is the
+gradient of the scalar \textit{normal potential},
+which is also the sum of a gravitational and a centrifugal potential.
+By definition, the centrifugal part of
+the normal potential and the gravity potential are equal.
 
-The difference between $\mitbf{g}_{P}$ and $\mitbf{\gamma}_{P}$, at the same
-point $P$, defines a quantity called \textit{gravity disturbance
-vector}, which is given by:
+The difference between $\mitbf{g}_{P}$ and $\mitbf{\gamma}_{P}$,
+which we emphasize are located at the same point $P$,
+is the \textit{gravity disturbance vector} (Fig. \ref{fig:surfaces}):
+
 \begin{equation}
@@
 \end{equation}
-Because the centrifugal part of the normal gravity vector
-is equal to the centrifugal component of the gravity vector,
-the gravity disturbance vector $\delta \mitbf{g}_{P}$ represents
-a purely gravitational effect caused by contrasts between the actual internal
-density distribution of the Earth and the internal density
-distribution of the normal Earth.
-In applied geophysics, these density differences are generally
-called \textit{anomalous masses} (e.g., \citeauthor{hammer1945},
-\citeyear{hammer1945}; \citeauthor{lafehr1965}, \citeyear{lafehr1965}),
-\textit{density anomalies} (e.g., \citeauthor{forsberg1984},
-\citeyear{forsberg1984}) or \textit{gravity sources} (e.g.,
-\citeauthor{blakely1996}, \citeyear{blakely1996}). Here, we opted for
-using the last term.
-Fig. \ref{fig:surfaces} illustrates the gravity vector
-$\mitbf{g}_{P}$, normal gravity vector $\mitbf{\gamma}_{P}$ and
-the gravity disturbance vector $\delta \mitbf{g}_{P}$ at a point $P$
-located on the Earth's surface.
+
+Because the centrifugal parts of both the normal gravity vector
+and the Earth's gravity vector are equal,
+the gravity disturbance vector $\delta \mitbf{g}_{P}$
+represents a purely gravitational effect.
+As a consequence of the superposition principal \citep{blakely1996},
+this effect is caused by contrasts between
+the internal density distributions
+of the actual Earth and the normal Earth.
+In applied geophysics, these density differences are generally called
+\textit{anomalous masses} \citep[e.g.,][]{hammer1945,lafehr1965},
+\textit{density anomalies} \citep[e.g.,][]{forsberg1984},
+or \textit{gravity sources} \citep[e.g.,][]{blakely1996}.
+Here, we opted for using the last term.
 

### 2018-05-16 manuscript/letter.tex
Use the vector symbol to make things clearer
It's very hard to see the bold gamma. I added the `\vec` command to draw
the little vector arrow to make things cleared. Kept the bold, like what
was used for the unit vectors.
I made the `\versor` and `\vector` commands to make it easier to control
the formatting for all the text.

@@
 \newtheorem{theorem}{Theorem}[section]
+
+\newcommand{\versor}[1]{\hat{\mathbf{#1}}}
+\renewcommand{\vector}[1]{\vec{\mathbf{#1}}}
 
@@
 \begin{equation}
-    \hat{\mitbf{u}} =
+    \versor{u} =
     \begin{bmatrix}
@@
     \end{bmatrix}
-    \hat{\mitbf{v}} =
+    \versor{v} =
         \begin{bmatrix}
@@
     \end{bmatrix}
-    \hat{\mitbf{w}} =
+    \versor{w} =
     \begin{bmatrix}
@@
 axes $x$ and $y$ are parallel to
-the unit vectors $\hat{\mitbf{v}}$ and $\hat{\mitbf{w}}$
+the unit vectors $\versor{v}$ and $\versor{w}$
 (eq. \ref{eq:unit-vectors}), respectively,
-and the $z$-axis is opposite to the unit vector $\hat{\mitbf{u}}$
+and the $z$-axis is opposite to the unit vector $\versor{u}$
 (Fig. \ref{fig:TCS}).
@@
 
-Let $\mitbf{\gamma}_{P}$ and $\mitbf{g}_{P}$ and be, respectively,
+Let $\vector{\gamma}_{P}$ and $\vector{g}_{P}$ and be, respectively,
 the \textit{normal gravity vector} and the Earth's \textit{gravity vector}
@@
 both located at a point $P$.
-In this case, the gravity vector $\mitbf{g}_{P}$ is the
+In this case, the gravity vector $\vector{g}_{P}$ is the
 gradient of the scalar \textit{gravity potential},
 which is the sum of a gravitational potential and a centrifugal potential.
-Similarly, the normal gravity vector $\mitbf{\gamma}_{P}$ is the
+Similarly, the normal gravity vector $\vector{\gamma}_{P}$ is the
 gradient of the scalar \textit{normal potential},
@@
 
-The difference between $\mitbf{g}_{P}$ and $\mitbf{\gamma}_{P}$,
+The difference between $\vector{g}_{P}$ and $\vector{\gamma}_{P}$,
 which we emphasize are located at the same point $P$,
@@
 \begin{equation}
-\delta \mitbf{g}_{P} =
-\mitbf{g}_{P} - \mitbf{\gamma}_{P} \: .
+\delta \vector{g}_{P} =
+\vector{g}_{P} - \vector{\gamma}_{P} \: .
 \label{eq:gravity-disturbance-vector}
@@
 and the Earth's gravity vector are equal,
-the gravity disturbance vector $\delta \mitbf{g}_{P}$
+the gravity disturbance vector $\delta \vector{g}_{P}$
 represents a purely gravitational effect.
@@
 The difference between the magnitudes of gravity vector
-$g_{P} = \| \mitbf{g}_{P} \|$ and the normal gravity vector
-$\gamma_{P} = \| \mitbf{\gamma}_{P} \|$, at the
+$g_{P} = \| \vector{g}_{P} \|$ and the normal gravity vector
+$\gamma_{P} = \| \vector{\gamma}_{P} \|$, at the
 same point $P$, is called \textit{gravity disturbance}
@@
 to the magnitude of the gravity disturbance vector
-$\delta \mitbf{g}_{P}$ \citep{barthelmes2013, sanso_sideris2013}.
+$\delta \vector{g}_{P}$ \citep{barthelmes2013, sanso_sideris2013}.
 
@@
 \begin{equation}
-\Delta \mitbf{g}_{P} = \mitbf{g}_{Q^{\prime}} - \mitbf{\gamma}_{Q} .
+\Delta \vector{g}_{P} = \vector{g}_{Q^{\prime}} - \vector{\gamma}_{Q} .
 \label{eq:gravity-anomaly-vector}
@@
 \end{equation}
-where $g_{Q^{\prime} } = \| \mitbf{g}_{Q^{\prime}} \|$ is
-the magnitude of the gravity vector $\mitbf{g}_{Q}$ on the Geoid,
-at $Q^{\prime}$, and $\gamma_{Q} = \| \mitbf{\gamma}_{Q} \|$ is the
-magnitude of the normal gravity vector $\mitbf{\gamma}_{Q}$
+where $g_{Q^{\prime} } = \| \vector{g}_{Q^{\prime}} \|$ is
+the magnitude of the gravity vector $\vector{g}_{Q}$ on the Geoid,
+at $Q^{\prime}$, and $\gamma_{Q} = \| \vector{\gamma}_{Q} \|$ is the
+magnitude of the normal gravity vector $\vector{\gamma}_{Q}$
 on the reference ellipsoid, at the point $Q$ (Fig. \ref{fig:surfaces}).
@@
 
-From eq. \ref{eq:gravity-disturbance-vector}, the gravity vector $\mitbf{g}_{i}$, at a point
+From eq. \ref{eq:gravity-disturbance-vector}, the gravity vector $\vector{g}_{i}$, at a point
 $(x_{i}, y_{i}, z_{i})$ in the TCS (Fig. \ref{fig:TCS}a), can be represented by:
 \begin{equation}
-\mitbf{g}_{i} = \mitbf{\gamma}_{i} + \delta \mitbf{g}_{i} \: ,
+\vector{g}_{i} = \vector{\gamma}_{i} + \delta \vector{g}_{i} \: ,
 \label{eq:gravity-vector-TCS}
 \end{equation}
-where $\mitbf{\gamma}_{i}$ and $\delta \mitbf{g}_{i}$
+where $\vector{\gamma}_{i}$ and $\delta \vector{g}_{i}$
 are, respectively, the normal gravity vector and the
@@
 g_{i} \approx \gamma_{i} +
-\hat{\mitbf{\gamma}}_{i}^{\top} \delta \mitbf{g}_{i} \: ,
+\versor{\gamma}_{i}^{\top} \delta \vector{g}_{i} \: ,
 \label{eq:gobs-approx}
@@
 where $^{\top}$ denotes transposition,
-$\hat{\mitbf{\gamma}}_{i} = -\hat{\mitbf{u}}_{i}$ is a unit
+$\versor{\gamma}_{i} = -\versor{u}_{i}$ is a unit
 vector with the same direction as the normal gravity vector
-$\mitbf{\gamma}_{i}$ at the point $(x_{i}, y_{i}, z_{i})$, in the
+$\vector{\gamma}_{i}$ at the point $(x_{i}, y_{i}, z_{i})$, in the
 topocentric Cartesian coordinate system (Fig. \ref{fig:TCS}a), and
-$\hat{\mitbf{u}}_{i}$ is the unit vector $\hat{\mitbf{u}}$
+$\versor{u}_{i}$ is the unit vector $\versor{u}$
 (eq. \ref{eq:unit-vectors}) evaluated at the corresponding
@@
 Notice that, in local- or regional-gravity studies, the unit
-vector $\hat{\mitbf{\gamma}}_{i}$ (eq. \ref{eq:gobs-approx})
+vector $\versor{\gamma}_{i}$ (eq. \ref{eq:gobs-approx})
 may be considered constant throughout the study area and
@@
 \begin{equation}
-\delta g_{i} \approx \hat{\mitbf{\gamma}}_{P}^{\top} \delta \mitbf{g}_{i} \: ,
+\delta g_{i} \approx \versor{\gamma}_{P}^{\top} \delta \vector{g}_{i} \: ,
 \label{eq:gravity-disturbance-approx-TCS}
 \end{equation}
-where $\hat{\mitbf{\gamma}}_{P}$ represents the unit vector with
-the same direction as the normal gravity vector $\mitbf{\gamma}_{P}$
+where $\versor{\gamma}_{P}$ represents the unit vector with
+the same direction as the normal gravity vector $\vector{\gamma}_{P}$
 at the origin $P$ of the TCS (Fig. \ref{fig:TCS}a).
 This equation shows that the gravity disturbance $\delta g_{i}$ (eq. \ref{eq:gravity-disturbance}) is different from the magnitude
-of the gravity disturbance vector $\delta \mitbf{g}_{i}$
+of the gravity disturbance vector $\delta \vector{g}_{i}$
 produced by the gravity sources. Rather, it represents the
-component of $\delta \mitbf{g}_{i}$ on the direction of the normal
+component of $\delta \vector{g}_{i}$ on the direction of the normal
 gravity vector \citep{hofmann-wellenhof-moritz2005, sanso_sideris2013}.
@@
     $P^{\prime}$ is represented in gray in (a) and (b).
-    The unit vectors $\hat{\mitbf{u}}_{P}$, $\hat{\mitbf{v}}_{P}$ and
-    $\hat{\mitbf{w}}_{P}$ define three mutually orthogonal
+    The unit vectors $\versor{u}_{P}$, $\versor{v}_{P}$ and
+    $\versor{w}_{P}$ define three mutually orthogonal
     directions at $P$ (eq. \ref{eq:unit-vectors}).
@@
     coordinate system (TCS) with origin at a point $P$. The axes $x$
-    and $y$ are parallel to the unit vectors $\hat{\mitbf{v}}_{P}$
-    and $\hat{\mitbf{w}}_{P}$ (eq. \ref{eq:unit-vectors} and
-    Fig. \ref{fig:GCS-GGS}), respectively. On the other hand, the $z$ axis is opposite to the unit vector $\hat{\mitbf{u}}_{P}$ (eq.
+    and $y$ are parallel to the unit vectors $\versor{v}_{P}$
+    and $\versor{w}_{P}$ (eq. \ref{eq:unit-vectors} and
+    Fig. \ref{fig:GCS-GGS}), respectively. On the other hand, the $z$ axis is opposite to the unit vector $\versor{u}_{P}$ (eq.
     \ref{eq:unit-vectors} and Fig. \ref{fig:GCS-GGS}) and points downward. The gray plane is the same shown in Fig. \ref{fig:GCS-GGS}.
     (b) Schematic representation of the gravity vector
-    $\mitbf{g}_{P}$, normal gravity vector $\mitbf{\gamma}_{P}$,
-    gravity disturbance vector $\delta \mitbf{g}_{P}$
+    $\vector{g}_{P}$, normal gravity vector $\vector{\gamma}_{P}$,
+    gravity disturbance vector $\delta \vector{g}_{P}$
     (eq. \ref{eq:gravity-disturbance-vector}) and unit vector
-    $\hat{\mitbf{u}}_{P}$ (eq. \ref{eq:unit-vectors}) at a point
+    $\versor{u}_{P}$ (eq. \ref{eq:unit-vectors}) at a point
     $P$.}
@@
     \caption{Schematic representation of the gravity vector
-        $\mitbf{g}_{P}$, normal gravity vector $\mitbf{\gamma}_{P}$,
-        gravity disturbance vector $\delta \mitbf{g}_{P}$
+        $\vector{g}_{P}$, normal gravity vector $\vector{\gamma}_{P}$,
+        gravity disturbance vector $\delta \vector{g}_{P}$
         (eq. \ref{eq:gravity-disturbance-vector}), unit vector
-        $\hat{\mitbf{u}}_{P}$ (eq. \ref{eq:unit-vectors}) at a point $P$
-        on the surface of the Earth, gravity vector $\mitbf{g}_{Q^{\prime}}$
+        $\versor{u}_{P}$ (eq. \ref{eq:unit-vectors}) at a point $P$
+        on the surface of the Earth, gravity vector $\vector{g}_{Q^{\prime}}$
         at a point $Q^{\prime}$ on the Geoid, normal gravity vector
-        $\mitbf{\gamma}_{Q}$ at a point $Q$ on the reference ellipsoid,
+        $\vector{\gamma}_{Q}$ at a point $Q$ on the reference ellipsoid,
         geometric height $h$, orthometric height $H$ and geoidal undulation

### 2018-05-16 manuscript/letter.tex
Changes to definition of anomaly and disturbance
Simplified some of the text and added some missing italics.
Removed the P subscript from the gravity anomaly because it is not
defined on the point P (which is where the disturbance is defined).
The anomaly can't be defined on a point in 3D space, as we comment right
below that. So it's best to just say Delta g or Delta g(lambda, phi).

Added a part about not being able to upward continue Delta g because
it's not harmonic. This was implied by a sentence below saying that we
can upcontinue the disturbance bc it is harmonic.

Rephrased the part ab

@@
 
-\newcommand{\versor}[1]{\hat{\mathbf{#1}}}
-\renewcommand{\vector}[1]{\vec{\mathbf{#1}}}
+\newcommand{\versor}[1]{\hat{\mitbf{#1}}}
+\renewcommand{\vector}[1]{\vec{\mitbf{#1}}}
 
@@
 
-\section{Relationship between gravity disturbance and gravity anomaly}
+\section{gravity disturbance and gravity anomaly}
 
@@
 
-The difference between the magnitudes of gravity vector
+The difference between the magnitudes of the gravity vector
 $g_{P} = \| \vector{g}_{P} \|$ and the normal gravity vector
-$\gamma_{P} = \| \vector{\gamma}_{P} \|$, at the
-same point $P$, is called \textit{gravity disturbance}
-\citep{heiskanen-moritz1967, hofmann-wellenhof-moritz2005}
-and can be represented as follows:
+$\gamma_{P} = \| \vector{\gamma}_{P} \|$,
+at the same point $P$, is the \textit{gravity disturbance}
+\citep{heiskanen-moritz1967, hofmann-wellenhof-moritz2005}:
 
@@
 
-Notice that the gravity disturbance $\delta g_{P}$ is not equivalent
+\noindent
+Notice that $\delta g_{P}$ is not equivalent
 to the magnitude of the gravity disturbance vector
@@
 
-On the other hand, the \textit{gravity anomaly vector} at $P$
+The \textit{gravity anomaly vector}
 is the difference between the gravity
-vector at a point $Q^{\prime}$ on the Geoid (a particular
-equipotential surface of the gravity potential) and the normal
-gravity vector at a point $Q$ on the ellipsoid surface, both at the
-same geodetic latitude and longitude (Fig. \ref{fig:surfaces}) as in
+vector at a point $Q^{\prime}$ on the geoid
+(a particular equipotential surface of the gravity potential)
+and the normal gravity vector at a point $Q$ on the surface of the ellipsoid,
+both located at the same geodetic latitude and longitude
+(Fig. \ref{fig:surfaces}):
+
 \begin{equation}
-\Delta \vector{g}_{P} = \vector{g}_{Q^{\prime}} - \vector{\gamma}_{Q} .
+\Delta \vector{g} = \vector{g}_{Q^{\prime}} - \vector{\gamma}_{Q} .
 \label{eq:gravity-anomaly-vector}
 \end{equation}
-Similarly to the gravity disturbance expression (eq.
-\ref{eq:gravity-disturbance}), the gravity anomaly is given by
+
+Similarly to the gravity disturbance (eq. \ref{eq:gravity-disturbance}),
+the \textit{gravity anomaly} is given by
+
 \begin{equation}
-\Delta g_{P} = g_{Q^{\prime}} - \gamma_{Q} ,
+\Delta g = g_{Q^{\prime}} - \gamma_{Q} ,
 \label{eq:gravity-anomaly}
 \end{equation}
-where $g_{Q^{\prime} } = \| \vector{g}_{Q^{\prime}} \|$ is
-the magnitude of the gravity vector $\vector{g}_{Q}$ on the Geoid,
-at $Q^{\prime}$, and $\gamma_{Q} = \| \vector{\gamma}_{Q} \|$ is the
-magnitude of the normal gravity vector $\vector{\gamma}_{Q}$
-on the reference ellipsoid, at the point $Q$ (Fig. \ref{fig:surfaces}).
-As properly pointed out by \citet{barthelmes2013} the gravity anomaly,
-by definition, depends on
-longitude and latitude only and is not a function of height.
-Because the gravity is on the Geoid and the normal gravity
-is on reference ellipsoid, the gravity anomaly $\Delta g_{P}$
-(eq. \ref{eq:gravity-anomaly}) is not a purely gravitational effect
-due to the gravity sources, but rather a combination of gravitational
-and centrifugal effects. Consequently, its upward continuation,
-for example, could not be rigorously done.
 
-However, many authors in the literature have computed the upward
-continuation of gravity anomalies. All these authors have
-implicitly considered that the gravity anomaly is an approximation of
-the gravity disturbance, which in turn can be represented by a
-harmonic function and can then being upward continued.
+\noindent
+where $g_{Q^{\prime}}$ is the magnitude of the gravity vector on the geoid,
+and $\gamma_{Q}$ is the magnitude of the normal gravity vector
+on the reference ellipsoid (Fig. \ref{fig:surfaces}).
+
+The gravity anomaly is not defined at a particular point.
+Instead, as properly pointed out by \citet{barthelmes2013},
+it depends only on longitude and latitude and is not a function of height.
+Because gravity and normal gravity are located at different points,
+the gravity anomaly is a combination of gravitational and centrifugal effects,
+rather than purely gravitational effects like the gravity disturbance.
+Consequently, its upward continuation, for example, cannot be rigorously done
+because it is not a harmonic function.
+However, upward continuation of gravity anomalies is commonly seen in the
+literature.
+In doing so, an implicit assumption is being made that the gravity anomaly is
+an approximation of the gravity disturbance,
+which in turn can be represented by a harmonic function and can be upward
+continued.
 

### 2018-05-16 manuscript/letter.tex
Changes to diff between disturbance and anomaly
When defining the free-air anomaly, the reduction is applied to g, not
gamma. It's reducing g from P to Q', not raising gamma.

I really don't like the term "geoidal undulation" (it has nothing to do
with waves) or, even worse, "geoid anomaly". N is a distance so "geoidal
height" is better and is what is used in the Physical Geodesy book.

@@
 
-Different gravity anomalies can be calculated,
+Different approximations for the gravity anomaly can be calculated
 depending on the corrections applied to them.
-These corrections are usually called \textit{gravity reductions}.
-The Free-air anomaly, for example, may be defined as
-follows \citep{blakely1996, hofmann-wellenhof-moritz2005}:
+These corrections are usually called \textit{gravity reductions}
+and their aim is to approximate the magnitude of gravity at the geoid
+$g_{Q^\prime}$ from the magnitude of gravity $g_P$ measured at a point $P$
+on or above the surface of the Earth.
+The \textit{free-air anomaly}, for example, can be defined as
+\citep{blakely1996, hofmann-wellenhof-moritz2005}:
+
 \begin{equation}
-\Delta g_{P}^{F}
-= g_{P} - \left( \gamma_{Q} + \frac{\partial \gamma}{\partial h} \, H_{P} \right) \: ,
+\Delta g^{F}
+= \left(g_{P} - \frac{\partial \gamma}{\partial h} \, H_{P} \right)
+- \gamma_{Q} \: ,
 \label{eq:free-air-anomaly}
 \end{equation}
-where $\frac{\partial \gamma}{\partial h} \approx -0.3086$ mGal/m is the
+
+\noindent
+where $\partial \gamma/\partial h \approx -0.3086$ mGal/m is the
 derivative of the normal gravity with respect to the geometric height $h$
-and $H_{P}$ is the orthometric height $H$ (Fig. \ref{fig:surfaces}) at the
-point $P$.
-Notice that the term between parenthesis represents an
-approximation of the normal gravity at the point $Q^{\prime}$ on the
-Geoid. By using a similar approach, we can approximate the normal gravity
-$\gamma_{P}$, at a point $P$ on the Earth's surface, so that the
-gravity disturbance $\delta g_{P}$ (eq. \ref{eq:gravity-disturbance})
-can be rewritten as follows:
+and $H_{P}$ is the orthometric height (i.e., with respect to the geoid)
+of point $P$ (Fig. \ref{fig:surfaces}).
+The term between parenthesis represents an
+approximation of the gravity at the point $Q^{\prime}$ on the geoid.
+Similarly, we can approximate the normal gravity $\gamma_{P}$
+at a point $P$ on the Earth's surface
+from the normal gravity at the ellipsoid $\gamma_{Q}$,
+so that the gravity disturbance $\delta g_{P}$
+(eq. \ref{eq:gravity-disturbance}) can be rewritten as:
+
 \begin{equation}
 \delta g_{P} \approx g_{P} -
-\left( \gamma_{Q} + \frac{\partial \gamma}{\partial h} h \right) \: .
+\left( \gamma_{Q} + \frac{\partial \gamma}{\partial h} h_P \right) \: .
 \label{eq:gravity-disturbance-approx}
 \end{equation}
-By inspection, one easily concludes that the absolute difference
-between the approximated gravity disturbance $\delta g_{P}$
-(eq. \ref{eq:gravity-disturbance-approx}), at a point $P$ on or close to
-the Earth's surface, and the Free-air anomaly
-$\Delta g_{P}^{F}$ (eq. \ref{eq:free-air-anomaly}), is given by:
+
+\noindent
+where $h_P$ is the geometric height of point P.
+Thus, the absolute difference between
+the approximated gravity disturbance (eq. \ref{eq:gravity-disturbance-approx}),
+at a point $P$ on or close to the Earth's surface,
+and the Free-air anomaly (eq. \ref{eq:free-air-anomaly}) is:
+
 \begin{equation}
-\vert \delta g_{P} - \Delta g^{F}_{P} \vert \approx
-\vert \frac{\partial \gamma}{\partial h} N \vert \: ,
+\left\vert \delta g_{P} - \Delta g^{F} \right\vert \approx
+\left\vert \frac{\partial \gamma}{\partial h} N \right\vert \: ,
 \label{eq:difference-disturbance-free-air}
 \end{equation}
-where $N \approx h - H$ is the geoidal undulation (Fig. \ref{fig:surfaces}).
-This approximation assumes that the Geoid and the surface of
-the reference ellipsoid are approximately parallel at $P$
+
+\noindent
+where $N \approx h - H$ is the geoidal height (Fig. \ref{fig:surfaces}).
+This approximation assumes that the geoid and the surface of
+the reference ellipsoid are close to being parallel at $P$
 and the surrounding area.
-We know empirically that the geoidal undulation $N$ in the world
-is $\approx \pm 1$ m on the oceans and reaches a
-maximum absolute value of $\approx 120$ m \citep[e.g.,][]{torge2012,
-sanso_sideris2013}. In Brazil, for example, the geoidal undulation
-reaches $\approx \pm 30$ m \citep{ibge_mapgeo2015}. Consequently, the
-maximum absolute differences between gravity disturbance and
-Free-air anomaly, according to eq. \ref{eq:difference-disturbance-free-air},
-is $\approx 9.258$ mGal.
-The approximations defined by eqs \ref{eq:gravity-disturbance-approx}
+Eqs \ref{eq:gravity-disturbance-approx}
 and \ref{eq:difference-disturbance-free-air} are commonly used
@@
 geodesy} \citep{hofmann-wellenhof-moritz2005}.
+
+We know empirically that $N$ is $\approx \pm 1$ m on the oceans
+and reaches a maximum absolute value of $\approx 120$ m
+\citep[e.g.,][]{torge2012, sanso_sideris2013}.
+In Brazil, for example, the geoidal height
+varies between $\approx \pm 30$ m \citep{ibge_mapgeo2015}.
+Consequently,
+the maximum absolute difference between the gravity disturbance and
+the free-air anomaly in Brazil,
+according to eq. \ref{eq:difference-disturbance-free-air},
+is $\approx 9.258$ mGal.
 

### 2018-05-17 code/helpers.py
Notebook to calculate disturbance - anomaly
The notebook contains all the code for loading the data and doing the
calculation. The data is not yet included.

@@ -1,0 +1,138 @@
+"""
+Helper functions for loading and plotting data.
+"""
+import numpy as np
+import xarray as xr
+import matplotlib.pyplot as plt
+import cartopy.crs as ccrs
+
+
+def minmax(data, fields):
+    """
+    Get the minimum and maximum data values out of all fields given.
+    Returns them in a dictionary with the 'vmin' and 'vmax' keys.
+    """
+    vmin = min(data[field].values.min() for field in fields)
+    vmax = max(data[field].values.max() for field in fields)
+    return dict(vmin=vmin, vmax=vmax)
+
+
+def plot_field(ax, data, field, cmap=None, gridline_spacing=30, cb_pad=0.03,
+               cb_aspect=50, cb_shrink=0.7, title=True, **kwargs):
+    """
+    Make a pcolormesh plot of the given data field.
+    Set's the plot extent and includes ticks in longitude and latitude.
+    """
+    if title:
+        ax.set_title(field)
+    if 'add_colorbar' not in kwargs:
+        kwargs['cbar_kwargs'] = dict(orientation='horizontal',
+                                     aspect=cb_aspect, pad=cb_pad,
+                                     shrink=cb_shrink)
+    data[field].plot.pcolormesh(ax=ax, add_labels=False, cmap=cmap,
+                                transform=ccrs.PlateCarree(), **kwargs)
+    ax.coastlines()
+    w, e, s, n = [data.longitude.min(), data.longitude.max(),
+                  data.latitude.min(), data.latitude.max()]
+    xlocs = np.arange(w, e + 0.01, gridline_spacing)
+    ylocs = np.arange(s, n + 0.01, gridline_spacing)
+    ax.gridlines(color="#cccccc55", xlocs=xlocs, ylocs=ylocs)
+
+
+def load_icgem_gdf(fname, dtype='float32'):
+    """
+    Load data from an ICGEM .gdf file into an xarray.Dataset.
+
+    Reads metdata from the header, like the grid area, number of points,
+    height over the ellipsoid (if it's constant), etc.
+
+    Stores the file header in the ``attrs`` attribute of the Dataset.
+
+    Parameters
+    ----------
+    fname : str
+        The name of the .gdf file.
+    dtype : str or numpy dtype object
+        The data type used when loading the data from the file.
+
+    Returns
+    -------
+    data : xarray.Dataset
+
+    """
+    with open(fname) as gdf_file:
+        # Read the header and extract metadata
+        header = []
[…]

### 2018-05-17 code/helpers.py
Calculate the difference and plot in GMT
Data is from the EIGEN-6c4 model. It had to be stored in tar.gz files
because the ascii were too big for Github.

@@ -1,6 +1,8 @@
 """
 Helper functions for loading and plotting data.
 """
+import tarfile
+
 import numpy as np
 import xarray as xr
 import matplotlib.pyplot as plt
@@ -39,7 +41,7 @@ def plot_field(ax, data, field, cmap=None, gridline_spacing=30, cb_pad=0.03,
     ax.gridlines(color="#cccccc55", xlocs=xlocs, ylocs=ylocs)
 
 
-def load_icgem_gdf(fname, dtype='float32'):
+def load_icgem_gdf(fname, dtype='float64'):
     """
     Load data from an ICGEM .gdf file into an xarray.Dataset.
 
@@ -47,11 +49,13 @@ def load_icgem_gdf(fname, dtype='float32'):
     height over the ellipsoid (if it's constant), etc.
 
     Stores the file header in the ``attrs`` attribute of the Dataset.
+    
+    Assumes that files are in tar files with gz compression.
 
     Parameters
     ----------
     fname : str
-        The name of the .gdf file.
+        The name of the .gdf.tar.gz file.
     dtype : str or numpy dtype object
         The data type used when loading the data from the file.
 
@@ -60,7 +64,8 @@ def load_icgem_gdf(fname, dtype='float32'):
     data : xarray.Dataset
 
     """
-    with open(fname) as gdf_file:
+    with tarfile.open(fname, 'r:gz') as archive:
+        gdf_file = archive.extractfile(archive.getmembers()[0])
         # Read the header and extract metadata
         header = []
         shape = [None, None]

### 2018-05-18 manuscript/letter.tex
The Bouguer correction part deserves a section
This is different enough from the discussion about the disturbance vs
free-air to deserve its own section. Plus, this way we end the previous
section with the part about the magnitude of the error from using the
anomaly.
Reworked the part about Bouguer and its meaning
This part was a bit disjointed and it wasn't address the important
question about the meaning of the corrections. In applied geophysics,
Bouguer is more important that free-air and more attention should be
given to it.
I put in the equation for a Bouguer disturbance to show that we s

@@
 
+\section{The geophysical meaning of gravity reductions}
+
 Different approximations for the gravity anomaly can be calculated
@@
 on or above the surface of the Earth.
+
 The \textit{free-air anomaly}, for example, can be defined as
@@
 The term between parenthesis represents an
-approximation of the gravity at the point $Q^{\prime}$ on the geoid.
-Similarly, we can approximate the normal gravity $\gamma_{P}$
+approximation of the gravity at the point $Q^{\prime}$ on the geoid,
+disregarding the effect of topographic masses.
+Similarly, we can apply the same correction
+to calculate the gravity disturbance by
+approximating the normal gravity $\gamma_{P}$
 at a point $P$ on the Earth's surface
-from the normal gravity at the ellipsoid $\gamma_{Q}$,
-so that the gravity disturbance $\delta g_{P}$
-(eq. \ref{eq:gravity-disturbance}) can be rewritten as:
+from the normal gravity at the ellipsoid $\gamma_{Q}$:
 
@@
 where $h_P$ is the geometric height of point P.
-Thus, the absolute difference between
-the approximated gravity disturbance (eq. \ref{eq:gravity-disturbance-approx}),
-at a point $P$ on or close to the Earth's surface,
-and the Free-air anomaly (eq. \ref{eq:free-air-anomaly}) is:
+Such an approximation is usually not required because
+an analytical expression exists for calculating $\gamma$ at any point
+on or above the ellipsoid \citep{li2001}.
+The same cannot be said for the free-air anomaly because
+there is usually no way of measuring gravity at the geoid ($g_{Q^\prime}$).
+
+Using eqs. \ref{eq:free-air-anomaly} and \ref{eq:gravity-disturbance-approx},
+we can estimate the absolute difference between the gravity disturbance and the
+free-air anomaly:
 
@@
 
-The simple Bouguer anomaly is commonly used by geophysicists as the
-gravitational effect produced by the gravity sources.
-It is defined, over the continents, as follows
+The \textit{Bouguer anomaly}
+has the aim of removing the gravitational effect of the topographic masses
+between the observation point $P$ and the point on the geoid $Q^\prime$.
+It is defined in the continents as
 \citep{blakely1996, hofmann-wellenhof-moritz2005}:
+
 \begin{equation}
-\Delta g_{P}^{B}
-= \Delta g_{P}^{F} - 2 \pi G \rho_{t} \, H_{P} \: ,
+\Delta g^{B} =
+\left(g_{P}
+    - \frac{\partial \gamma}{\partial h} \, H_{P}
+    - 2 \pi G \rho_{t} \, H_{P} \right)
+- \gamma_{Q} \: ,
 \label{eq:simple-bouguer-anomaly}
 \end{equation}
-where $\Delta g_{P}^{F}$ is the Free-air anomaly (eq. \ref{eq:free-air-anomaly})
-and the last term on the right side is the simple Bouguer correction.
-This term approximates the gravitational attraction that all topographic
-masses above the Geoid exerts at the point $P$ by the gravitational
-attraction of a homogeneous, infinitely extended slab of
-constant density $\rho_{t}$ and thickness equal to $H_{P}$, which is
-the orthometric height $H$ (Fig. \ref{fig:surfaces}) at the point $P$.
-By computing the simple Bouguer correction at a set of points
-of a survey around $P$, we approximately remove,
-from the gravity measured at the points of the survey,
-the gravitational effect of a homogeneous model representing
-the topographic masses above the Geoid.
-It is evident from eqs
-\ref{eq:free-air-anomaly} to \ref{eq:difference-disturbance-free-air}
-that the simple Bouguer anomaly $\Delta g_{P}^{B}$ (eq.
-\ref{eq:simple-bouguer-anomaly}) approximates the gravity disturbance
-$\delta g_{P}$ (eq. \ref{eq:gravity-disturbance})
-minus the gravitational effect produced by this homogeneous topographic
-model.
-Although this approximation is valid for most practical applications,
-it is important to bear in mind not only the terminology
-changes, but also the conceptual assumptions.
 
-There was a certain lack of comprehension regarding the
-geophysical meaning of gravity anomalies until the
-mid 90's.
-As properly pointed out by \citet{chapin1996} at that time,
-``although the corrections which bring about a Bouguer
-gravity anomaly are well established, the reasons for doing
-them are not well understood. One cause of this common
-misunderstanding is that the subject has been poorly presented in
-many of the basic texts".
-In his seminal book, \citet{blakely1996} brought some light
-on the geophysical meaning of gravity anomalies from the
-perspective of applied geophysics. \citet{blakely1996} correctly
-defined gravity sources as density contrasts between the actual
-internal density distribution of the Earth and the internal density
-distribution of the normal Earth.
-However, he did not stress that, by removing the normal gravity
-evaluated on the ellipsoid from the gravity measured
-on the Earth's surface, the remaining anomalous field will reflect
-not only the effect produced by the gravity sources, but also a
-small combination of gravitational and centrifugal effects.
-This additional, non-harmonic and undesired effect is
-simply due to the calculation of the normal gravity at a point
-other than that were the gravity is measured.
+\noindent
+where $\rho_t$ is the density, $G$ is the gravitational constant,
+and the rightmost term in parenthesis is the simple Bouguer correction.
+This term approximates the gravitational effect that all topographic
+masses above the geoid exert on point $P$ by the effect of a homogeneous,
+infinitely extended slab of constant density $\rho_{t}$ and thickness $H_{P}$.
+The same correction can be applied to the gravity disturbance
+(using the approximation in eq. \ref{eq:gravity-disturbance-approx})
+with the aim of removing the effect of topographic masses between
+the observation point $P$ and a point on the ellipsoid $Q$.
+We will call this quantity the \textit{Bouguer disturbance}:
+
+\begin{equation}
+\delta g_P^{B} \approx
+g_{P} -
+\left( \gamma_{Q} + \frac{\partial \gamma}{\partial h} h_P \right)
+- 2 \pi G \rho_{t} \, h_{P} \: .
+\label{eq:bouguer-disturbance}
+\end{equation}
+
+At first glance, the only difference between eqs.
+\ref{eq:simple-bouguer-anomaly} and \ref{eq:bouguer-disturbance} is the use of
+orthometric ($H_P$) versus geometric ($h_P$) heights.
+However, their objectives and geophysical interpretations are significantly
+different.
+The corrections applied when calculating the Bouguer anomaly approximate
+gravity at the geoid from measurements made on the Earth's surface by removing
+the effects of the distance and the topographic masses between the two points.
+Apart from the centrifugal effect mentioned in the previous section,
+the gravitational effect of topographic masses between the geoid and ellipsoid
+are still present in the anomaly.
+Conversely, when calculating the Bouguer disturbance one removes from
+the observed gravity the effects of an ellipsoidal reference model (the normal
+Earth) and the known topographic masses between the surface of the ellipsoid
+and observation point.
+Hence, the Bouguer disturbance can be interpreted as being solely the
+gravitational effect produced by the density contrasts between the actual
+internal density distribution of the Earth and that of the normal Earth.
+
+There was a certain lack of comprehension regarding the geophysical meaning of
+gravity anomalies until the mid 90s.
+As properly pointed out by \citet{chapin1996},
+``although the corrections which bring about a Bouguer gravity anomaly are well
+established, the reasons for doing them are not well understood. One cause of
+this common misunderstanding is that the subject has been poorly presented in
+many of the basic texts''.
+In his seminal book, \citet{blakely1996} shed some light on the geophysical
+meaning of gravity anomalies from the perspective of applied geophysics and
+their relationship to the gravity sources.
+We complement this interpretation by stressing the fact that
+gravity anomalies reflect not only the gravity sources,
+but also a combination of gravitational and centrifugal effects.
+These additional, non-harmonic and undesired effects are due to the calculation
+of the normal gravity at a point other than that were gravity is measured.
+In practice, this issue can be avoided entirely by performing the free-air and
+Bouguer corrections using geometric instead of orthometric heights.
 

### 2018-05-18 manuscript/letter.tex
Review of the part about gravity modeling
Renames the section to just "Gravity modeling" because that is what it's
about.
Adds an introduction to the section and removes a lot of repetition to
make the text flow better.
Also change some of the language to avoid blaming people too much.

@@
 different.
-The corrections applied when calculating the Bouguer anomaly approximate
+The corrections applied in eq. \ref{eq:simple-bouguer-anomaly} approximate
 gravity at the geoid from measurements made on the Earth's surface by removing
@@
 
-\section{Mathematical description of the gravity disturbance at the TCS}
+\section{Gravity modelling}
 
-From eq. \ref{eq:gravity-disturbance-vector}, the gravity vector $\vector{g}_{i}$, at a point
-$(x_{i}, y_{i}, z_{i})$ in the TCS (Fig. \ref{fig:TCS}a), can be represented by:
+We have established that the gravity disturbance is due solely to the
+gravitational effect of the gravity sources.
+Now we will discuss how this relates to some assumptions made during gravity
+modelling.
+
+The gravity vector $\vector{g}_i$ (eq. \ref{eq:gravity-disturbance-vector})
+at a point $(x_i, y_i, z_i)$ in the TCS (Fig. \ref{fig:TCS}a) can be
+represented by:
+
 \begin{equation}
-\vector{g}_{i} = \vector{\gamma}_{i} + \delta \vector{g}_{i} \: ,
+\vector{g}_i = \vector{\gamma}_i + \delta \vector{g}_i \: .
 \label{eq:gravity-vector-TCS}
 \end{equation}
-where $\vector{\gamma}_{i}$ and $\delta \vector{g}_{i}$
-are, respectively, the normal gravity vector and the
-gravity disturbance vector produced by the anomalous
-masses at the point $(x_{i}, y_{i}, z_{i})$.
 
-The gravity $g_{i}$ can be approximated by a first order Taylor's
-expansion as follows \citep{sanso_sideris2013}:
+\noindent
+Furthermore, the gravity $g_i$ can be approximated by a first order Taylor's
+expansion \citep{sanso_sideris2013}:
+
 \begin{equation}
-g_{i} \approx \gamma_{i} +
-\versor{\gamma}_{i}^{\top} \delta \vector{g}_{i} \: ,
+g_i \approx \gamma_i +
+\versor{\gamma}_i^{\top} \delta \vector{g}_i \: ,
 \label{eq:gobs-approx}
 \end{equation}
-where $^{\top}$ denotes transposition,
-$\versor{\gamma}_{i} = -\versor{u}_{i}$ is a unit
-vector with the same direction as the normal gravity vector
-$\vector{\gamma}_{i}$ at the point $(x_{i}, y_{i}, z_{i})$, in the
-topocentric Cartesian coordinate system (Fig. \ref{fig:TCS}a), and
-$\versor{u}_{i}$ is the unit vector $\versor{u}$
-(eq. \ref{eq:unit-vectors}) evaluated at the corresponding
-point $(h_{i}, \varphi_{i}, \lambda_{i})$ in the
-GGS (Fig. \ref{fig:GCS-GGS}).
-This approximation can be made because, fortunately,
-the condition $\gamma_{i} \gg \| \delta \mathbf{g}_{i} \|$
-is met at all points located above or on the Earth's surface.
-This approximation is known in geodesy \citep[e.g.,][]{sanso_sideris2013}
-and a similar approximation is largely used in applied geophysics
-for representing total-field anomalies \citep[e.g.,][]{blakely1996}.
-Notice that, in local- or regional-gravity studies, the unit
-vector $\versor{\gamma}_{i}$ (eq. \ref{eq:gobs-approx})
+
+\noindent
+where $^{\top}$ denotes transposition and $\versor{\gamma}_{i}$ is a unit
+vector with the direction of the normal gravity vector.
+This approximation is valid because
+$\gamma_{i} \gg \| \delta \mathbf{g}_{i} \|$ at all points on or above the
+Earth's surface.
+It is known in geodesy \citep[e.g.,][]{sanso_sideris2013}
+and is analogous to the approximation used for
+total-field magnetic anomalies \citep[e.g.,][]{blakely1996}.
+In local or regional scales, the unit
+vector $\versor{\gamma}_{i}$
 may be considered constant throughout the study area and
 parallel to the $z$ axis of the TCS (Fig. \ref{fig:TCS}a).
-By using the approximation defined in eq.
-\ref{eq:gobs-approx}, the gravity disturbance
-(eq. \ref{eq:gravity-disturbance}) can be rewritten as
-follows
+
+Using eq. \ref{eq:gobs-approx} and fixing $\gamma$ at a point $P$ at
+the origin of the TCS (Fig. \ref{fig:TCS}a),
+the gravity disturbance can be rewritten as:
+
 \begin{equation}
-\delta g_{i} \approx \versor{\gamma}_{P}^{\top} \delta \vector{g}_{i} \: ,
+\delta g_{i} \approx \versor{\gamma}_{P}^{\top} \delta \vector{g}_{i} \: .
 \label{eq:gravity-disturbance-approx-TCS}
 \end{equation}
-where $\versor{\gamma}_{P}$ represents the unit vector with
-the same direction as the normal gravity vector $\vector{\gamma}_{P}$
-at the origin $P$ of the TCS (Fig. \ref{fig:TCS}a).
-This equation shows that the gravity disturbance $\delta g_{i}$ (eq. \ref{eq:gravity-disturbance}) is different from the magnitude
-of the gravity disturbance vector $\delta \vector{g}_{i}$
-produced by the gravity sources. Rather, it represents the
-component of $\delta \vector{g}_{i}$ on the direction of the normal
-gravity vector \citep{hofmann-wellenhof-moritz2005, sanso_sideris2013}.
 
-In the TCS (Fig. \ref{fig:TCS}a),
-the gravity disturbance $\delta g_{i}$ (eq. \ref{eq:gravity-disturbance-approx-TCS})
-can be defined as the $z$-component of the gravitational attraction exerted by
-the gravity sources at the point $(x_{i}, y_{i}, z_{i})$.
-As a consequence, the gravity disturbance produced by
-a gravity source can be represented by the following harmonic function:
+\noindent
+This equation shows that the gravity disturbance $\delta g_{i}$ is
+approximately the component of the $\delta \vector{g}_{i}$ in the direction of
+the normal gravity vector $\vector{\gamma}_P$
+\citep{hofmann-wellenhof-moritz2005, sanso_sideris2013}.
+Because $z$-axis of the TCS is parallel to $\vector{\gamma}_P$
+(Fig. \ref{fig:TCS}b), the gravity disturbance can be defined as the
+$z$-component of the gravitational attraction exerted by the gravity sources at
+the point $(x_{i}, y_{i}, z_{i})$.
+As a consequence, the gravity disturbance can be modelled by the harmonic
+function:
+
 \begin{equation}
-d_{i} = \int\int\limits_{v}\int
-\frac{c_{g} \, G \, \Delta\rho(x^{\prime}, y^{\prime}, z^{\prime})
-\, (z_{i} - z^{\prime}) \: dv^{\prime}}
-{\sqrt{(x_{i} - x^{\prime})^{2} +
-(y_{i} - y^{\prime})^{2} + (z_{i} - z^{\prime})^{2}}} \: ,
+d_{i} = \iiint\limits_{v}
+\frac{G \, \Delta\rho(x^{\prime}, y^{\prime}, z^{\prime}) \,
+[z_{i} - z^{\prime}]}
+{\sqrt{[x_{i} - x^{\prime}]^{2} +
+[y_{i} - y^{\prime}]^{2} + [z_{i} - z^{\prime}]^{2}}} \: dv^{\prime} \: ,
 \label{eq:gz-local}
 \end{equation}
-where $G$ is the Newtonian constant of gravitation (in $m^{3} \, kg^{-1} \, s^{-2}$),
-$c_{g} = 10^{5}$ is a constant factor transforming from $m \, s^{-2}$
-to milligal (mGal), $\Delta\rho(x^{\prime}, y^{\prime}, z^{\prime})$
-is the density contrast (in $kg \, m^{-3}$) at a point
-$(x^{\prime}, y^{\prime}, z^{\prime})$ within
-the volume $v$ of the gravity source and the integration is conducted
-over $x^{\prime}$, $y^{\prime}$ and $z^{\prime}$.
-This equation can be easily generalized for the case
-of multiple gravity sources.
-Practically, all articles accomplishing gravity modelling
-use the quantity $d_{i}$ (eq. \ref{eq:gz-local})
-to represent the gravity anomaly produced by
-gravity sources \citep[e.g.,][]{blakely1996}.
-Consequently, almost all geophysicists use the gravity
-anomaly as an approximation of the gravity disturbance
-produced by the gravity sources. Notice that $d_{i}$
-(eq. \ref{eq:gz-local}) does not depend on the
-height with respect to the Geoid (orthometric height).
-Rather, it depends on the relative position
-of the observation points $(x_{i}, y_{i}, z_{i})$ with respect to
-the gravity sources in the TCS (Fig. \ref{fig:TCS}a).
-Finally, it is important to notice that, contrary to what
-is written in some basic texts, gravimeters do not measure
-the quantity $d_{i}$ (eq. \ref{eq:gz-local}).
+
+\noindent
+where $\Delta\rho(x^{\prime}, y^{\prime}, z^{\prime})$
+is the density contrast at a point $(x^{\prime}, y^{\prime}, z^{\prime})$
+within the volume $v$ and the integration is conducted over $x^{\prime}$,
+$y^{\prime}$ and $z^{\prime}$.
+
+In practice, most instances of gravity modelling in the literature use Bouguer
+anomalies and function $d_{i}$ (usually presented as ``$g_z$'') to represent
+the gravitational effect of gravity sources.
+Consequently, these studies are using the gravity
+anomaly as an approximation for the gravity disturbance.
+Notice that $d_{i}$ does not depend on the
+height with respect to the geoid (orthometric height).
+Rather, it depends on the relative position of the gravity sources and the
+observation point $(x_{i}, y_{i}, z_{i})$ in the TCS (Fig. \ref{fig:TCS}a).
+Finally, it is important to mention that gravimeters do not measure $d_{i}$ but
+$\| \vector{g}_i \|$.
 

### 2018-05-18 manuscript/letter.tex
Review the conclusions
Removed a lot of repetition and added a paragraph to the end to end in a
high note with a recommendation of what people can do (just stop using
orthometric heights).

@@
 
-Although the gravity disturbance is a well-known quantity in geodesy,
-its use in the applied geophysics is uncommon.
-For years, geophysicists have been using gravity anomaly for modelling
-the gravitational effects due to geological bodies.
-The question "Should geophysicists use gravity anomaly or gravity
-disturbance for modelling the gravitational effect of gravity sources?"
-is of more than academic interest.
+Although the gravity disturbance is well-known in geodesy,
+its use in applied geophysics is uncommon.
+For years, geophysicists have been using gravity anomalies for modelling
+the gravitational effects of geological bodies.
+Wether geophysicists should use the gravity anomaly or the gravity disturbance
+is of more than mere academic interest.
+It is a question that underpins the very assumptions about what it is that we
+are actually modelling.
 
-Theoretically, the gravity anomaly is the difference between the
-gravity on the Geoid and the normal gravity on the reference ellipsoid,
-both at the same geodetic latitude and longitude.
-On the other hand, the gravity disturbance is defined by the difference
-between gravity and normal gravity at the same point.
-Because these quantities are at the same point, the gravity disturbance
-is a purely gravitational effect due to the gravity sources whereas the
-gravity anomaly is a combination of gravitational and centrifugal
+The gravity anomaly is the difference between
+gravity on the geoid and normal gravity on the reference ellipsoid.
+The gravity disturbance is the difference between gravity and normal gravity at
+the same point.
+Because of this, the gravity disturbance is a purely gravitational effect,
+whereas the gravity anomaly is a combination of gravitational and centrifugal
 effects.
-Consequently, the gravity disturbance is a harmonic function whereas the
+Consequently, the gravity disturbance is a harmonic function while the
 gravity anomaly is not.
-Because the gravity disturbance is a purely gravitational effect, it can
-be represented by a harmonic function and satisfies the premise required
-by the most techniques for processing potential-field data (e.g.,
-upward/downward continuation, data processing with equivalent layer,
+Being a harmonic function is the premise required by most techniques for
+processing potential-field data
+(e.g., upward/downward continuation, operations with the equivalent layer,
 conversions between gravity and magnetic data, computation of vertical
-derivatives via Fourier and Hilbert transforms).
-Hence, the gravity disturbance is the more appropriated quantity for
-representing the gravity effect produced by the gravity sources.
+derivatives).
+Hence, the gravity disturbance is the logical choice for
+representing the gravitational effects produced by the heterogeneous density
+distribution of the Earth.
 
-In practical problems of applied geophysics, geophysicists perform the
-gravity-forward modelling by computing the vertical component of the
-gravitational attraction exerted by the sources at the observation
-points, regardless of the Geoid or reference ellipsoid positions,
-for example.
-This means that, actually, they compute gravity disturbance to fit
-a gravity anomaly (e.g., Bouguer anomaly) and this is a clear inconsistency.
-Moreover, upward or downward continuations, for example, are commonly applied to
-gravity anomalies. In this case, the gravity anomaly is wrongly considered a
-harmonic function and this is another inconsistency.
+More often than not, gravity modelling is done by computing the
+vertical component of the gravitational acceleration exerted by the sources.
+We have shown that, in doing so, one is actually computing the gravity
+disturbance while attempting to fit a gravity anomaly.
+This is a clear inconsistency.
+Doing so may even result in wrong interpretations of a quantitative nature.
+In Brazil, for example, the maximum difference
+between the gravity disturbance and the free-air anomaly reaches 9 mGal.
+Such difference at the Earth's surface may lead to
+an error of XXX m in the depth-to-center estimate of a sphere with
+true depth-to-center at xxx m, radius xxxx m, and density
+contrast of xxx $kg.m^{-3}$.
 
-By adopting these practical procedures, geophysicists are implicitly
-taking the gravity anomaly as a good approximation of the gravity
-disturbance. But why should geophysicists care about this usual
-procedure? Because it may result in wrong interpretations of
-quantitative nature. In Brazil, for example, the maximum difference
-between the gravity disturbance and the free-air anomaly reaches 9.3
-mGal. Such difference on the Earth's surface may lead, for example, to
-an error of about xxx m in the depth-to-center estimate of a sphere-like
-source with true depth-to-center at xxx m, radius xxxx m and density
-contrast of xxx $kg \, m^{-3}$.
+Fortunately, all of these issues are nullified during data processing by
+substituting orthometric heights (i.e., with respect to the geoid) with
+geometric heights (i.e., with respect to the ellipsoid).
+For modern acquisitions, this requires no extra work because GPS already
+provide geometric heights.
+The free-air correction can even be bypassed entirely.
+As a bonus, the gravity disturbance is conceptually simpler to understand and
+doesn't require the concept of the geoid with which learners often struggle.
 

### 2018-05-18 manuscript/letter.tex
If we're gonna be provocative, start with the title
Simplify the abstract and add a solution to the end

@@
 \title[Gravity anomaly or gravity disturbance?]
-      {Should geophysicists use gravity anomaly or gravity disturbance?}
+      {Geophysicists should use the gravity disturbance rather than the gravity anomaly}
 
@@
 \begin{summary}
- In applied geophysics, the gravity anomaly has long been used for representing the
- gravitational effect produced by gravity sources, which are defined as the density
- contrasts between the internal density distribution of the Earth and the normal Earth.
- The gravity anomaly is defined by the difference between the gravity (produced by the
- Earth) on the Geoid and the normal gravity (produced by the normal Earth)
- on reference ellipsoid, both at the same geodetic latitude and longitude.
- Because these quantities are not defined at the same point,
- the gravity anomaly does not represent the gravitational effect due to the gravity
- sources and, consequently, cannot be considered a harmonic function.
- On the contrary, the gravity disturbance is defined by the difference between
- the gravity and normal gravity at the same point. Consequently, the centrifugal
- effects can be neglected, the gravity disturbance can be considered a harmonic
- function representing the gravitational effect due to the gravity sources and then
- it satisfies the premise behind most of the processing techniques of potential-field
- data (e.g., upward/downward continuations). Then, rigorously, the gravity disturbance
- is  more appropriated to be used for estimating density variations in subsurface.
- The use the gravity anomaly for most of the geophysicists carries the
- implicit meaning that it can be considered a good approximation of the gravity
- disturbance. It is important to bear in mind, however, that in Brazil, for example,
- the maximum absolute difference between the gravity disturbance (on or close
- to the Earth’s surface) and the Free-air anomaly reaches 9.3 mGal.
-
+ The gravity anomaly is defined as the difference between the Earth's gravity
+ on the geoid and the gravity of the normal Earth on the reference ellipsoid.
+ Because these quantities are not at the same point, the anomaly contains
+ centrifugal accelerations and cannot be considered a harmonic function.
+ The gravity disturbance is the difference between gravity and normal gravity
+ at the same point.
+ Consequently, the centrifugal effects can be neglected and the disturbance can
+ be considered a harmonic function.
+ This is the premise behind most potential-field data processing techniques
+ (e.g., upward/downward continuation and vertical derivatives).
+ Unlike the anomaly, the disturbance is due solely to the
+ gravitational effects of geologic sources, making it the most appropriate
+ choice for modelling.
+ Use of the gravity anomaly in geophysics carries with it the implicit
+ assumption that it is a good approximation for the gravity disturbance.
+ However, bear in mind that the difference between the gravity disturbance and
+ the free-air anomaly can be larger than 10 mGal worldwide.
+ Fortunately, the solution to these problems is to simply replace
+ heights referenced to the geoid (orthometric) with heights referenced to the
+ ellipsoid (geometric) for geophysical use cases.
 \end{summary}

### 2018-05-18 manuscript/letter.tex
Simplify the figure captions by removing redundancy
A lot of the definitions are in the text already so we don't need to
repeat them.

@@
     \includegraphics{figures/cartesian-geodetic-systems.png}
-    \caption{Schematic representation of the Geocentric Cartesian System (GCS) and the Geocentric Geodetic System (GGS). The GCS has the $Z$-axis coincident with the mean Earth's rotational axis,
-    the $X$-axis pointing to the Greenwich meridian and the $Y$-axis
-    directed so as to obtain a right-handed system. The GGS is      defined by an oblate ellipsoid with semi-minor axis $b$,
-    coincident with the $Z$-axis of GCS, and a semi-major
-    axis $a$. In this coordinate system, the position of a point is
-    determined by the geometric height $h$, geodetic latitude
-    $\varphi$ and longitude $\lambda$.
-    The Earth's center of mass is represented
-    by $O$, $P$ represents a point $(h_{P}, \varphi_{P}, \lambda_{P})$ and $P^{\prime}$ its projection onto the plane $XY$
-    (Equatorial plane). The plane containing $O$, $P$ and
-    $P^{\prime}$ is represented in gray in (a) and (b).
-    The unit vectors $\versor{u}_{P}$, $\versor{v}_{P}$ and
-    $\versor{w}_{P}$ define three mutually orthogonal
-    directions at $P$ (eq. \ref{eq:unit-vectors}).
-    In (b), $Q$ represents a point $(h_{Q}, \varphi_{Q}, \lambda_{Q})$,
-    which is the projection of $P$ onto the reference ellipsoid, at
-    the same latitude and longitude ($\varphi_{Q} = \varphi_{P}$ and $\lambda_{Q} = \lambda_{P}$).}
+    \caption{
+    The Geocentric Cartesian System (GCS) and the Geocentric Geodetic System
+    (GGS).
+    The GGS is defined by an oblate ellipsoid with semi-minor axis $b$ and
+    semi-major axis $a$.
+    In this coordinate system, the position of a point is determined by the
+    geometric height $h$, geodetic latitude $\varphi$, and longitude $\lambda$.
+    $O$ is the Earth's center of mass, $P$ is an observation point, and
+    unit vectors $\versor{u}$, $\versor{v}$, and $\versor{w}$ define mutually
+    orthogonal directions at $P$ (eq. \ref{eq:unit-vectors}).
+    In (b), $Q$ is the projection of $P$ onto the reference ellipsoid at
+    the same latitude and longitude.
+    }
   \label{fig:GCS-GGS}
@@
     \includegraphics{figures/local-system.png}
-    \caption{(a) Schematic representation of a topocentric Cartesian
-    coordinate system (TCS) with origin at a point $P$. The axes $x$
-    and $y$ are parallel to the unit vectors $\versor{v}_{P}$
-    and $\versor{w}_{P}$ (eq. \ref{eq:unit-vectors} and
-    Fig. \ref{fig:GCS-GGS}), respectively. On the other hand, the $z$ axis is opposite to the unit vector $\versor{u}_{P}$ (eq.
-    \ref{eq:unit-vectors} and Fig. \ref{fig:GCS-GGS}) and points downward. The gray plane is the same shown in Fig. \ref{fig:GCS-GGS}.
+    \caption{
+    (a) The topocentric Cartesian coordinate system (TCS) with origin at a
+    point $P$.
+    Axes $x$ and $y$ are parallel to the unit vectors $\versor{v}$ and
+    $\versor{w}$ (eq. \ref{eq:unit-vectors} and Fig. \ref{fig:GCS-GGS}),
+    respectively, and the $z$ axis is opposite to $\versor{u}$.
+    The gray plane is the same shown in Fig. \ref{fig:GCS-GGS}.
     (b) Schematic representation of the gravity vector
@@
     $\versor{u}_{P}$ (eq. \ref{eq:unit-vectors}) at a point
-    $P$.}
+    $P$.
+    }
   \label{fig:TCS}
@@
     \includegraphics{figures/surfaces.png}
-    \caption{Schematic representation of the gravity vector
+    \caption{
+    The gravity vector
     $\vector{g}_{P}$, normal gravity vector $\vector{\gamma}_{P}$,
     gravity disturbance vector $\delta \vector{g}_{P}$
-        (eq. \ref{eq:gravity-disturbance-vector}), unit vector
+    (eq. \ref{eq:gravity-disturbance-vector}), and unit vector
     $\versor{u}_{P}$ (eq. \ref{eq:unit-vectors}) at a point $P$
-        on the surface of the Earth, gravity vector $\vector{g}_{Q^{\prime}}$
-        at a point $Q^{\prime}$ on the Geoid, normal gravity vector
+    on the surface of the Earth.
+    The gravity vector $\vector{g}_{Q^{\prime}}$
+    at a point $Q^{\prime}$ on the geoid, normal gravity vector
     $\vector{\gamma}_{Q}$ at a point $Q$ on the reference ellipsoid,
-        geometric height $h$, orthometric height $H$ and geoidal undulation
-        $N$ \citep{heiskanen-moritz1967}.
+    geometric height $h$, orthometric height $H$, and geoidal height $N$.
     The dashed line passing through
     $Q$, $Q^{\prime}$ and $P$ is normal to the
-        surface of the reference ellipsoid at $Q$. This figure
-        shows a commonly used approximation in which the ellipsoid
-        surface and the Geoid are represented as parallel surfaces,
-        so that $h \approx H + N$.}
+    surface of the reference ellipsoid at $Q$.
+    This figure shows a commonly used approximation in which the ellipsoid and
+    geoid are represented as parallel surfaces so that $h \approx H + N$.
+    }
   \label{fig:surfaces}

### 2018-05-18 manuscript/letter.tex
Remove the mentions of the purpose of reductions
Modifications made after our live discussion.

There is no reason to get into that in this paper. We're already
polemical enough.
Changed the title back to a question.
Part about free-air is a discussion of it being an approximation to the
disturbance so moved it back to the previous section.
Bouguer is now a section on the topographic masses.

@@
 \title[Gravity anomaly or gravity disturbance?]
-      {Geophysicists should use the gravity disturbance rather than the gravity anomaly}
+      {Should geophysicists use the gravity disturbance or the anomaly?}
 
@@
  gravitational effects of geologic sources, making it the most appropriate
- choice for modelling.
+ for geophysical purposes.
  Use of the gravity anomaly in geophysics carries with it the implicit
@@
  the free-air anomaly can be larger than 10 mGal worldwide.
- Fortunately, the solution to these problems is to simply replace
- heights referenced to the geoid (orthometric) with heights referenced to the
- ellipsoid (geometric) for geophysical use cases.
+ SOMETHING ABOUT MODELLING.
 \end{summary}
@@
 
-\section{The geophysical meaning of gravity reductions}
-
 Different approximations for the gravity anomaly can be calculated
 depending on the corrections applied to them.
-These corrections are usually called \textit{gravity reductions}
-and their aim is to approximate the magnitude of gravity at the geoid
-$g_{Q^\prime}$ from the magnitude of gravity $g_P$ measured at a point $P$
-on or above the surface of the Earth.
-
+These corrections are usually called \textit{gravity reductions}.
 The \textit{free-air anomaly}, for example, can be defined as
@@
 \begin{equation}
-\Delta g^{F}
-= \left(g_{P} - \frac{\partial \gamma}{\partial h} \, H_{P} \right)
-- \gamma_{Q} \: ,
+\Delta g^{F} = g_{P}
+- \gamma_{Q} - \frac{\partial \gamma}{\partial h} \, H_{P} \: ,
 \label{eq:free-air-anomaly}
@@
 \noindent
-where $\partial \gamma/\partial h \approx -0.3086$ mGal/m is the
-derivative of the normal gravity with respect to the geometric height $h$
+where the rightmost term is called the \textit{free-air correction},
+$\partial \gamma/\partial h \approx -0.3086$ mGal/m is the
+derivative of the normal gravity with respect to the geometric height $h$,
 and $H_{P}$ is the orthometric height (i.e., with respect to the geoid)
 of point $P$ (Fig. \ref{fig:surfaces}).
-The term between parenthesis represents an
-approximation of the gravity at the point $Q^{\prime}$ on the geoid,
-disregarding the effect of topographic masses.
-Similarly, we can apply the same correction
+The free-air correction can be interpreted as a downward continuation of
+$g_P$ to the geoid or as an upward continuation of $\gamma_Q$ to a height
+$H_P$.
+Regardless, we can also apply the free-air correction
 to calculate the gravity disturbance by
@@
 where $h_P$ is the geometric height of point P.
-Such an approximation is usually not required because
-an analytical expression exists for calculating $\gamma$ at any point
-on or above the ellipsoid \citep{li2001}.
-The same cannot be said for the free-air anomaly because
-there is usually no way of measuring gravity at the geoid ($g_{Q^\prime}$).
+However, using the free-air correction to calculate the disturbance is
+discouraged because an analytical expression exists for calculating $\gamma$ at
+any point on or above the ellipsoid \citep{li2001}.
 
@@
 
-The \textit{Bouguer anomaly}
-has the aim of removing the gravitational effect of the topographic masses
+\section{The effect of topographic masses}
+
+The \textit{Bouguer correction} can be applied to the free-air anomaly
+to remove the gravitational effect of the topographic masses
 between the observation point $P$ and the point on the geoid $Q^\prime$.
-It is defined in the continents as
+The \textit{Bouguer anomaly} is defined in the continents as
 \citep{blakely1996, hofmann-wellenhof-moritz2005}:
@@
 \begin{equation}
-\Delta g^{B} =
-\left(g_{P}
+\Delta g^{B} = g_{P}
+    - \gamma_{Q}
     - \frac{\partial \gamma}{\partial h} \, H_{P}
-    - 2 \pi G \rho_{t} \, H_{P} \right)
-- \gamma_{Q} \: ,
+    - 2 \pi G \rho_{t} \, H_{P}
+     \: ,
 \label{eq:simple-bouguer-anomaly}
@@
 \noindent
-where $\rho_t$ is the density, $G$ is the gravitational constant,
-and the rightmost term in parenthesis is the simple Bouguer correction.
-This term approximates the gravitational effect that all topographic
-masses above the geoid exert on point $P$ by the effect of a homogeneous,
-infinitely extended slab of constant density $\rho_{t}$ and thickness $H_{P}$.
-The same correction can be applied to the gravity disturbance
+where $\rho_t$ is the density of the topographic masses and $G$ is the
+gravitational constant.
+Th Bouguer correction approximates the gravitational effect that all
+topographic masses above the geoid exert on point $P$ by the effect of a
+homogeneous, infinitely extended slab of constant density $\rho_{t}$ and
+thickness $H_{P}$.
+A similar correction can be applied to the gravity disturbance
 (using the approximation in eq. \ref{eq:gravity-disturbance-approx})
-with the aim of removing the effect of topographic masses between
+to remove the effect of topographic masses between
 the observation point $P$ and a point on the ellipsoid $Q$.
@@
 orthometric ($H_P$) versus geometric ($h_P$) heights.
-However, their objectives and geophysical interpretations are significantly
-different.
-The corrections applied in eq. \ref{eq:simple-bouguer-anomaly} approximate
-gravity at the geoid from measurements made on the Earth's surface by removing
-the effects of the distance and the topographic masses between the two points.
+However, their geophysical interpretations are significantly different.
 Apart from the centrifugal effect mentioned in the previous section,
-the gravitational effect of topographic masses between the geoid and ellipsoid
-are still present in the anomaly.
-Conversely, when calculating the Bouguer disturbance one removes from
-the observed gravity the effects of an ellipsoidal reference model (the normal
-Earth) and the known topographic masses between the surface of the ellipsoid
-and observation point.
+the Bouguer anomaly (eq. \ref{eq:simple-bouguer-anomaly}) still contains
+the gravitational effect of topographic masses between the geoid and ellipsoid.
+Conversely, when calculating the Bouguer disturbance one removes the effects of
+an ellipsoidal reference model (the normal Earth) and the known topographic
+masses between the surface of the ellipsoid and observation point.
 Hence, the Bouguer disturbance can be interpreted as being solely the
 gravitational effect produced by the density contrasts between the actual
-internal density distribution of the Earth and that of the normal Earth.
+internal density distribution of the Earth and that of the normal Earth plus a
+homogeneous topography.
 
@@
 of the normal gravity at a point other than that were gravity is measured.
-In practice, this issue can be avoided entirely by performing the free-air and
-Bouguer corrections using geometric instead of orthometric heights.
 
@@
 
-Fortunately, all of these issues are nullified during data processing by
+In practice, most of these issues can be nullified during data processing by
 substituting orthometric heights (i.e., with respect to the geoid) with
@@
 provide geometric heights.
-The free-air correction can even be bypassed entirely.
+The free-air correction can even be bypassed entirely by computing $\gamma_P$
+directly.
 As a bonus, the gravity disturbance is conceptually simpler to understand and

### 2018-05-22 manuscript/letter.tex
Final sentence for the abstract
Use new figures and place them in the text

@@
  The gravity anomaly is defined as the difference between the Earth's gravity
- on the geoid and the gravity of the normal Earth on the reference ellipsoid.
+ on the geoid and the normal gravity on the reference ellipsoid.
  Because these quantities are not at the same point, the anomaly contains
@@
  This is the premise behind most potential-field data processing techniques
- (e.g., upward/downward continuation and vertical derivatives).
+ (e.g., upward/downward continuation).
  Unlike the anomaly, the disturbance is due solely to the
@@
  the free-air anomaly can be larger than 10 mGal worldwide.
- SOMETHING ABOUT MODELLING.
+ In fact, we argue that the assumptions made during gravity forward and inverse
+ modeling imply that the quantity being modelled is the disturbance, not the
+ anomaly.
 \end{summary}
@@
 the $X$-axis pointing to the Greenwich meridian,
-and the $Y$-axis completing a right-handed system (Fig. \ref{fig:GCS-GGS}).
+and the $Y$-axis completing a right-handed system (Fig. \ref{fig:systems}).
 This reference system is called by different names in the literature:
@@
 the \textit{geodetic latitude} $\varphi$,
-and the \textit{longitude} $\lambda$ (Fig. \ref{fig:GCS-GGS}).
+and the \textit{longitude} $\lambda$ (Fig. \ref{fig:systems}).
 There are also three mutually-orthogonal unit vectors defined at a given point
@@
 and the $z$-axis is opposite to the unit vector $\versor{u}$
-(Fig. \ref{fig:TCS}).
+(Fig. \ref{fig:systems}c).
+
+\begin{figure}
+%    \vspace{5.5cm}
+    \includegraphics{figures/cartesian-geodetic-systems.eps}
+    \caption{
+    (a) and (b) The Geocentric Cartesian System (GCS) and the Geocentric
+    Geodetic System (GGS).
+    The GGS is defined by an oblate ellipsoid with semi-minor axis $b$ and
+    semi-major axis $a$.
+    In this coordinate system, the position of a point is determined by the
+    geometric height $h$, geodetic latitude $\varphi$, and longitude $\lambda$.
+    $O$ is the Earth's center of mass, $P$ is an observation point, and
+    unit vectors $\versor{u}$, $\versor{v}$, and $\versor{w}$ define mutually
+    orthogonal directions at $P$ (eq. \ref{eq:unit-vectors}).
+    In (b), $Q$ is the projection of $P$ onto the reference ellipsoid at
+    the same latitude and longitude.
+    (c) The topocentric Cartesian coordinate system (TCS) with origin at a
+    point $P$.
+    Axes $x$ and $y$ are parallel to the unit vectors $\versor{v}$ and
+    $\versor{w}$, respectively, and the $z$ axis is opposite to $\versor{u}$.
+    The gray plane is the same shown in (a) and (b).
+    }
+  \label{fig:systems}
+\end{figure}
+
+\begin{figure}
+    \includegraphics{figures/surfaces.png}
+    \caption{
+    The gravity vector
+    $\vector{g}_{P}$, normal gravity vector $\vector{\gamma}_{P}$,
+    gravity disturbance vector $\delta \vector{g}_{P}$
+    (eq. \ref{eq:gravity-disturbance-vector}), and unit vector
+    $\versor{u}_{P}$ (eq. \ref{eq:unit-vectors}) at a point $P$
+    on the surface of the Earth.
+    The gravity vector $\vector{g}_{Q^{\prime}}$
+    at a point $Q^{\prime}$ on the geoid, normal gravity vector
+    $\vector{\gamma}_{Q}$ at a point $Q$ on the reference ellipsoid,
+    geometric height $h$, orthometric height $H$, and geoidal height $N$.
+    The dashed line passing through
+    $Q$, $Q^{\prime}$ and $P$ is normal to the
+    surface of the reference ellipsoid at $Q$.
+    This figure shows a commonly used approximation in which the ellipsoid and
+    geoid are represented as parallel surfaces so that $h \approx H + N$.
+    }
+  \label{fig:surfaces}
+\end{figure}
 
@@
 gravitational constant.
-Th Bouguer correction approximates the gravitational effect that all
+The Bouguer correction approximates the gravitational effect that all
 topographic masses above the geoid exert on point $P$ by the effect of a
@@
 masses between the surface of the ellipsoid and observation point.
-Hence, the Bouguer disturbance can be interpreted as being solely the
-gravitational effect produced by the density contrasts between the actual
-internal density distribution of the Earth and that of the normal Earth plus a
-homogeneous topography.
 
@@
 The gravity vector $\vector{g}_i$ (eq. \ref{eq:gravity-disturbance-vector})
-at a point $(x_i, y_i, z_i)$ in the TCS (Fig. \ref{fig:TCS}a) can be
+at a point $(x_i, y_i, z_i)$ in the TCS (Fig. \ref{fig:systems}c) can be
 represented by:
@@
 may be considered constant throughout the study area and
-parallel to the $z$ axis of the TCS (Fig. \ref{fig:TCS}a).
+parallel to the $z$ axis of the TCS (Fig. \ref{fig:systems}c).
 
 Using eq. \ref{eq:gobs-approx} and fixing $\gamma$ at a point $P$ at
-the origin of the TCS (Fig. \ref{fig:TCS}a),
+the origin of the TCS (Fig. \ref{fig:systems}c),
 the gravity disturbance can be rewritten as:
@@
 Because $z$-axis of the TCS is parallel to $\vector{\gamma}_P$
-(Fig. \ref{fig:TCS}b), the gravity disturbance can be defined as the
+(Fig. \ref{fig:surfaces}), the gravity disturbance can be defined as the
 $z$-component of the gravitational attraction exerted by the gravity sources at
@@
 Rather, it depends on the relative position of the gravity sources and the
-observation point $(x_{i}, y_{i}, z_{i})$ in the TCS (Fig. \ref{fig:TCS}a).
+observation point $(x_{i}, y_{i}, z_{i})$ in the TCS (Fig. \ref{fig:systems}c).
 Finally, it is important to mention that gravimeters do not measure $d_{i}$ but
@@
 
-\appendix
-%\section{For authors}
-%
-%Table~\ref{authors} is a list of design macros which are unique to GJI. The
-%list displays each macro's name and description.
-%
-%\section{For editors}
-%
-%The additional features shown in Table~\ref{editors} may be used for
-%production purposes.
-%
-%\bsp % ``This paper has been produced using the Blackwell
-%     %   Publishing GJI \LaTeXe\ class file.''
-
-\begin{figure}
-%    \vspace{5.5cm}
-    \includegraphics{figures/cartesian-geodetic-systems.png}
-    \caption{
-    The Geocentric Cartesian System (GCS) and the Geocentric Geodetic System
-    (GGS).
-    The GGS is defined by an oblate ellipsoid with semi-minor axis $b$ and
-    semi-major axis $a$.
-    In this coordinate system, the position of a point is determined by the
-    geometric height $h$, geodetic latitude $\varphi$, and longitude $\lambda$.
-    $O$ is the Earth's center of mass, $P$ is an observation point, and
-    unit vectors $\versor{u}$, $\versor{v}$, and $\versor{w}$ define mutually
-    orthogonal directions at $P$ (eq. \ref{eq:unit-vectors}).
-    In (b), $Q$ is the projection of $P$ onto the reference ellipsoid at
-    the same latitude and longitude.
-    }
-  \label{fig:GCS-GGS}
-\end{figure}
-
-\begin{figure}
-    \includegraphics{figures/local-system.png}
-    \caption{
-    (a) The topocentric Cartesian coordinate system (TCS) with origin at a
-    point $P$.
-    Axes $x$ and $y$ are parallel to the unit vectors $\versor{v}$ and
-    $\versor{w}$ (eq. \ref{eq:unit-vectors} and Fig. \ref{fig:GCS-GGS}),
-    respectively, and the $z$ axis is opposite to $\versor{u}$.
-    The gray plane is the same shown in Fig. \ref{fig:GCS-GGS}.
-    (b) Schematic representation of the gravity vector
-    $\vector{g}_{P}$, normal gravity vector $\vector{\gamma}_{P}$,
-    gravity disturbance vector $\delta \vector{g}_{P}$
-    (eq. \ref{eq:gravity-disturbance-vector}) and unit vector
-    $\versor{u}_{P}$ (eq. \ref{eq:unit-vectors}) at a point
-    $P$.
-    }
-  \label{fig:TCS}
-\end{figure}
-
-\begin{figure}
-    \includegraphics{figures/surfaces.png}
-    \caption{
-    The gravity vector
-    $\vector{g}_{P}$, normal gravity vector $\vector{\gamma}_{P}$,
-    gravity disturbance vector $\delta \vector{g}_{P}$
-    (eq. \ref{eq:gravity-disturbance-vector}), and unit vector
-    $\versor{u}_{P}$ (eq. \ref{eq:unit-vectors}) at a point $P$
-    on the surface of the Earth.
-    The gravity vector $\vector{g}_{Q^{\prime}}$
-    at a point $Q^{\prime}$ on the geoid, normal gravity vector
-    $\vector{\gamma}_{Q}$ at a point $Q$ on the reference ellipsoid,
-    geometric height $h$, orthometric height $H$, and geoidal height $N$.
-    The dashed line passing through
-    $Q$, $Q^{\prime}$ and $P$ is normal to the
-    surface of the reference ellipsoid at $Q$.
-    This figure shows a commonly used approximation in which the ellipsoid and
-    geoid are represented as parallel surfaces so that $h \approx H + N$.
-    }
-  \label{fig:surfaces}
-\end{figure}
-
-\label{lastpage}
-
 \end{document}

### 2018-05-22 manuscript/letter.tex

@@
 \begin{figure}
-    \includegraphics{figures/surfaces.png}
+    \includegraphics{figures/surfaces.eps}
     \caption{
@@
 \citep[e.g.,][]{torge2012, sanso_sideris2013}.
+Fig. \ref{fig:difference} shows the absolute difference between the
+gravity disturbance and the free-anomaly globally calculated
+using eq. \ref{eq:difference-disturbance-free-air} and the spherical harmonic
+model EIGEN-6c4 \citep{forste2014}.
 In Brazil, for example, the geoidal height
@@
 is $\approx 9.258$ mGal.
+
+\begin{figure}
+    \includegraphics[width=\columnwidth]{figures/difference.eps}
+    \caption{
+    Absolute difference between the gravity disturbance and free-air anomaly.
+    Calculated using eq. \ref{eq:difference-disturbance-free-air}, the
+    EIGEN-6c4 \citep{forste2014} spherical harmonic model, and software
+    \textit{Fatiando a Terra} \citep{uieda2013}. The figure was made using GMT
+    \citep{wessel2013}.
+    }
+  \label{fig:difference}
+\end{figure}
 

### 2018-05-24 manuscript/letter.tex
Inclui os valores na conclusao sobre a modelagem da esfera

@@
 \begin{keywords}
- potential fields -- gravity disturbance -- gravity anomaly -- gravity modelling.
+ Geopotential theory -- Gravity anomalies and Earth structure -- Reference systems -- Numerical modelling.
 \end{keywords}
@@
 according to eq. \ref{eq:difference-disturbance-free-air},
-is $\approx 9.258$ mGal.
+is $\approx 9.3$ mGal.
 
@@
 In Brazil, for example, the maximum difference
-between the gravity disturbance and the free-air anomaly reaches 9 mGal.
+between the gravity disturbance and the free-air anomaly reaches $\approx 9.3$ mGal.
 Such difference at the Earth's surface may lead to
-an error of XXX m in the depth-to-center estimate of a sphere with
-true depth-to-center at xxx m, radius xxxx m, and density
-contrast of xxx $kg.m^{-3}$.
+an error of $850$ m in the depth-to-center estimate of a sphere with true depth-to-center at $700$ m, radius $699$ m, and density contrast of $600 kg.m^{-3}$.
 

### 2018-05-24 manuscript/letter.tex
review the sphere code

@@
 Such difference at the Earth's surface may lead to
-an error of $850$ m in the depth-to-center estimate of a sphere with true depth-to-center at $700$ m, radius $699$ m, and density contrast of $600 kg.m^{-3}$.
+an error of $850$ m in the depth-to-center estimate of a sphere with true 
+depth-to-center at $700$ m, radius $699$ m, and density contrast of $600$ kg~$\cdot$~m~$^{-3}$.
 

### 2018-05-29 manuscript/letter.tex
rephrase the sentence about the sphere

--- 2018-05-24 manuscript/letter.tex
@@
 \parbox{\linewidth}{%
-    Kristoffer A. T. Hallam$^{1}$\thanks{Emails: kristoffer.hallam@gmail.com, vanderlei@on.br},
-    Vanderlei C. Oliveira Jr$^1$,
-    Val\'{e}ria C. F. Barbosa$^1$, and \linebreak Leonardo Uieda$^2$
+    Vanderlei C. Oliveira Jr$^1$\thanks{Email: vanderlei@on.br},
+    Leonardo Uieda$^2$, Val\'{e}ria C. F. Barbosa$^1$, and
+    \linebreak Kristoffer A. T. Hallam$^{1}$ 
     \vspace{0.3cm}
@@
 between the gravity disturbance and the free-air anomaly reaches $\approx 9.3$ mGal.
-Such difference at the Earth's surface may lead to
-an error of $850$ m in the depth-to-center estimate of a sphere with true depth-to-center at $700$ m, radius $699$ m, and density contrast of $600 kg.m^{-3}$.
+This is approximately the maximum difference between the vertical gravitational 
+accelerations of a sphere with depth-to-center of $700$ m and one with $1550$ m, 
+assuming both spheres have a radius of $699$ m and density contrast of 
+$600$ kg~$\cdot$~m~$^{-3}$.
 

### 2018-05-29 manuscript/letter.tex
Add Zenodo DOI to README and text
Included a link to the DOI and Github repository in the paper
acknowledgements section.

@@
 \usepackage{amsmath}
+\usepackage[pdftex,colorlinks=true]{hyperref}
+\hypersetup{
+    allcolors=black,
+}
 
@@
 
-\author[Hallam et al.]{
+\author[Oliveira Jr et al.]{
 \parbox{\linewidth}{%
@@
 \begin{acknowledgments}
-The authors would like to thank the editor and all the
-reviewers for their criticisms and corrections.
+The authors would like to thank the editor and reviewers for their criticisms
+and corrections.
+All source code and data used to generate the results presented here are
+available at
+\href{https://doi.org/10.5281/zenodo.1255306}{doi.org/10.5281/zenodo.1255306}
+and
+\href{https://github.com/pinga-lab/use-the-disturbance}{github.com/pinga-lab/use-the-disturbance}.
 \end{acknowledgments}

### 2018-05-30 manuscript/letter.tex
change authors order
small correction

@@
     Vanderlei C. Oliveira Jr$^1$\thanks{Email: vanderlei@on.br},
-    Leonardo Uieda$^2$, Val\'{e}ria C. F. Barbosa$^1$, and
-    \linebreak Kristoffer A. T. Hallam$^{1}$
-    \vspace{0.3cm}
-}%
+    Leonardo Uieda$^2$, Kristoffer A. T. Hallam$^{1}$ and
+    \linebreak Val\'{e}ria C. F. Barbosa$^1$
+    \vspace{0.3cm} }%
     \\
@@
 the gravitational effects of geological bodies.
-Wether geophysicists should use the gravity anomaly or the gravity disturbance
+Whether geophysicists should use the gravity anomaly or the gravity disturbance
 is of more than mere academic interest.

### 2018-06-18 manuscript/letter.tex
change the latex template

@@
-% gjilguid2e.tex
-% V2.0 released 1998 December 18
-% V2.1 released 2003 October 7 -- Gregor Hutton, updated the web address for the style files.
+\documentclass[paper,twocolumn,twoside]{geophysics}
+%\documentclass[manuscript,revised]{geophysics}
 
-%\documentclass[extra,mreferee]{gji}
-\documentclass[extra]{gji}
-\usepackage{timet}
+% An example of defining macros
+\newcommand{\rs}[1]{\mathstrut\mbox{\scriptsize\rm #1}}
+\newcommand{\rr}[1]{\mbox{\rm #1}}
 
-%Extra packages
-\usepackage{graphicx}
+% Extra packages
 \usepackage{amsmath}
+\usepackage{bm}
 \usepackage[pdftex,colorlinks=true]{hyperref}
@@
 
-\title[Gravity anomaly or gravity disturbance?]
-      {Should geophysicists use the gravity disturbance or the anomaly?}
-
-\author[Oliveira Jr et al.]{
-\parbox{\linewidth}{%
-    Vanderlei C. Oliveira Jr$^1$\thanks{Email: vanderlei@on.br},
-    Leonardo Uieda$^2$, Kristoffer A. T. Hallam$^{1}$ and
-    \linebreak Val\'{e}ria C. F. Barbosa$^1$
-    \vspace{0.3cm} }%
-    \\
-    $^1$ Department of Geophysics, Observat\'{o}rio Nacional, Rio de Janeiro, Brazil \\
-    $^2$ Department of Geology and Geophysics, SOEST, University of Hawai'i at M\={a}noa, Honolulu, USA
-}
-
-\date{Received 2018 Month XX; in original form 2018 Month XX}
-\pagerange{\pageref{firstpage}--\pageref{lastpage}}
-\volume{XXX}
-\pubyear{2018}
-
-%\def\LaTeX{L\kern-.36em\raise.3ex\hbox{{\small A}}\kern-.15em
-%    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}
-%\def\LATeX{L\kern-.36em\raise.3ex\hbox{{\Large A}}\kern-.15em
-%    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}
-% Authors with AMS fonts and mssymb.tex can comment out the following
-% line to get the correct symbol for Geophysical Journal International.
-\let\leqslant=\leq
-
-\newtheorem{theorem}{Theorem}[section]
-
-\newcommand{\versor}[1]{\hat{\mitbf{#1}}}
-\renewcommand{\vector}[1]{\vec{\mitbf{#1}}}
+\renewcommand{\figdir}{figures} % figure directory
 
@@
 
-\label{firstpage}
+\title{Should geophysicists use the gravity disturbance or the anomaly?}
+
+\renewcommand{\thefootnote}{\fnsymbol{footnote}}
+
+\ms{GEO-XXXX} % manuscript number
+
+\address{
+\footnotemark[1]Observat\'{o}rio Nacional, \\
+Department of Geophysics, \\
+Rio de Janeiro, Brazil \\
+\footnotemark[2]University of Hawai'i at M\={a}noa, \\
+Department of Geology and Geophysics, SOEST, \\
+Honolulu, USA}
+\author{Vanderlei C. Oliveira Jr\footnotemark[1], Leonardo Uieda\footnotemark[2],
+Kristoffer A. T. Hallam\footnotemark[1] and Val\'{e}ria C. F. Barbosa\footnotemark[1]}
+
+\footer{vanderlei@on.br, leouieda@gmail.com, kristoffer.hallam@gmail.com, valcris@on.br}
+\lefthead{Oliveira Jr et al.}
+\righthead{Gravity anomaly or gravity disturbance?}
 
@@
 
-\begin{summary}
+\begin{abstract}
   The gravity anomaly is defined as the difference between the Earth's gravity
@@
  In fact, we argue that the assumptions made during gravity forward and inverse
- modeling imply that the quantity being modelled is the disturbance, not the
+ modeling imply that the quantity being modeled is the disturbance, not the
  anomaly.
-\end{summary}
-
-\begin{keywords}
- Geopotential theory -- Gravity anomalies and Earth structure -- Reference systems -- Numerical modelling.
-\end{keywords}
+\end{abstract}
 
@@
 the geological structures is known in geophysics as
-\textit{gravity modelling}.
-Not to be confused with \textit{gravity field modelling}, which
+\textit{gravity modeling}.
+Not to be confused with \textit{gravity field modeling}, which
 is commonly used in geodesy to denote the mathematical characterization of the
@@
 anomaly should be used in geophysical applications.
-The concepts discussed here are well-established in the literature.
+The concepts discussed here are well-established in the literature and
+are familiar to most geodesists.
 However, the debate around this theoretical issue has been
@@
 \citep{lafehr1991,chapin1996,li2001,fairhead2003,
-hackney-featherstone2003,hinze2005}.
+hackney-featherstone2003,hinze2005} and maybe that is why it is
+unfamiliar to most exploration geophysicists.
 We will approach the topic by comparing the geophysical interpretations of the
@@
 
-\section{Terrestrial Reference Systems used in gravity modelling}
+\section{Terrestrial Reference Systems used in gravity modeling}
 
-For geophysical gravity modelling purposes, there are three important
+For geophysical gravity modeling purposes, there are three important
 Terrestrial Reference Systems.
@@
 the $X$-axis pointing to the Greenwich meridian,
-and the $Y$-axis completing a right-handed system (Fig. \ref{fig:systems}).
+and the $Y$-axis completing a right-handed system (Figure \ref{fig:systems}).
 This reference system is called by different names in the literature:
@@
 the \textit{geodetic latitude} $\varphi$,
-and the \textit{longitude} $\lambda$ (Fig. \ref{fig:systems}).
+and the \textit{longitude} $\lambda$ (Figure \ref{fig:systems}).
 There are also three mutually-orthogonal unit vectors defined at a given point
@@
 \begin{equation}
-    \versor{u} =
-    \begin{bmatrix}
+    \mathbf{u} =
+    \begin{pmatrix}
         \cos\varphi \, \cos\lambda \\
@@
         \sin\varphi
-    \end{bmatrix}
-    \versor{v} =
-        \begin{bmatrix}
+    \end{pmatrix}
+    \mathbf{v} =
+    \begin{pmatrix}
         -\sin\varphi \, \cos\lambda \\
@@
         \cos\varphi
-    \end{bmatrix}
-    \versor{w} =
-    \begin{bmatrix}
+    \end{pmatrix}
+    \mathbf{w} =
+    \begin{pmatrix}
         -\sin\lambda \\
@@
         0
-    \end{bmatrix} .
+    \end{pmatrix} \: .
     \label{eq:unit-vectors}
@@
 axes $x$ and $y$ are parallel to
-the unit vectors $\versor{v}$ and $\versor{w}$
-(eq. \ref{eq:unit-vectors}), respectively,
-and the $z$-axis is opposite to the unit vector $\versor{u}$
-(Fig. \ref{fig:systems}c).
-
-\begin{figure}
-%    \vspace{5.5cm}
-    \includegraphics{figures/cartesian-geodetic-systems.eps}
-    \caption{
-    (a) and (b) The Geocentric Cartesian System (GCS) and the Geocentric
-    Geodetic System (GGS).
-    The GGS is defined by an oblate ellipsoid with semi-minor axis $b$ and
-    semi-major axis $a$.
-    In this coordinate system, the position of a point is determined by the
-    geometric height $h$, geodetic latitude $\varphi$, and longitude $\lambda$.
-    $O$ is the Earth's center of mass, $P$ is an observation point, and
-    unit vectors $\versor{u}$, $\versor{v}$, and $\versor{w}$ define mutually
-    orthogonal directions at $P$ (eq. \ref{eq:unit-vectors}).
-    In (b), $Q$ is the projection of $P$ onto the reference ellipsoid at
-    the same latitude and longitude.
-    (c) The topocentric Cartesian coordinate system (TCS) with origin at a
-    point $P$.
-    Axes $x$ and $y$ are parallel to the unit vectors $\versor{v}$ and
-    $\versor{w}$, respectively, and the $z$ axis is opposite to $\versor{u}$.
-    The gray plane is the same shown in (a) and (b).
-    }
-  \label{fig:systems}
-\end{figure}
-
-\begin{figure}
-    \includegraphics{figures/surfaces.eps}
-    \caption{
-    The gravity vector
-    $\vector{g}_{P}$, normal gravity vector $\vector{\gamma}_{P}$,
-    gravity disturbance vector $\delta \vector{g}_{P}$
-    (eq. \ref{eq:gravity-disturbance-vector}), and unit vector
-    $\versor{u}_{P}$ (eq. \ref{eq:unit-vectors}) at a point $P$
-    on the surface of the Earth.
-    The gravity vector $\vector{g}_{Q^{\prime}}$
-    at a point $Q^{\prime}$ on the geoid, normal gravity vector
-    $\vector{\gamma}_{Q}$ at a point $Q$ on the reference ellipsoid,
-    geometric height $h$, orthometric height $H$, and geoidal height $N$.
-    The dashed line passing through
-    $Q$, $Q^{\prime}$ and $P$ is normal to the
-    surface of the reference ellipsoid at $Q$.
-    This figure shows a commonly used approximation in which the ellipsoid and
-    geoid are represented as parallel surfaces so that $h \approx H + N$.
-    }
-  \label{fig:surfaces}
-\end{figure}
+the unit vectors $\mathbf{v}$ and $\mathbf{w}$
+(equation \ref{eq:unit-vectors}), respectively,
+and the $z$-axis is opposite to the unit vector $\mathbf{u}$
+(Figure \ref{fig:systems}c).
 
@@
 
-Let $\vector{\gamma}_{P}$ and $\vector{g}_{P}$ and be, respectively,
+Let $\boldsymbol{\gamma}_{P}$ and $\mathbf{g}_{P}$ and be, respectively,
 the \textit{normal gravity vector} and the Earth's \textit{gravity vector}
@@
 both located at a point $P$.
-In this case, the gravity vector $\vector{g}_{P}$ is the
+In this case, the gravity vector $\mathbf{g}_{P}$ is the
 gradient of the scalar \textit{gravity potential},
 which is the sum of a gravitational potential and a centrifugal potential.
-Similarly, the normal gravity vector $\vector{\gamma}_{P}$ is the
+Similarly, the normal gravity vector $\boldsymbol{\gamma}_{P}$ is the
 gradient of the scalar \textit{normal potential},
@@
 
-The difference between $\vector{g}_{P}$ and $\vector{\gamma}_{P}$,
+The difference between $\mathbf{g}_{P}$ and $\boldsymbol{\gamma}_{P}$,
 which we emphasize are located at the same point $P$,
-is the \textit{gravity disturbance vector} (Fig. \ref{fig:surfaces}):
+is the \textit{gravity disturbance vector} (Figure \ref{fig:surfaces}):
 
 \begin{equation}
-\delta \vector{g}_{P} =
-\vector{g}_{P} - \vector{\gamma}_{P} \: .
+\delta \mathbf{g}_{P} =
+\mathbf{g}_{P} - \boldsymbol{\gamma}_{P} \: .
 \label{eq:gravity-disturbance-vector}
@@
 and the Earth's gravity vector are equal,
-the gravity disturbance vector $\delta \vector{g}_{P}$
+the gravity disturbance vector $\delta \mathbf{g}_{P}$
 represents a purely gravitational effect.
@@
 The difference between the magnitudes of the gravity vector
-$g_{P} = \| \vector{g}_{P} \|$ and the normal gravity vector
-$\gamma_{P} = \| \vector{\gamma}_{P} \|$,
+$g_{P} = \| \mathbf{g}_{P} \|$ and the normal gravity vector
+$\gamma_{P} = \| \boldsymbol{\gamma}_{P} \|$,
 at the same point $P$, is the \textit{gravity disturbance}
@@
 to the magnitude of the gravity disturbance vector
-$\delta \vector{g}_{P}$ \citep{barthelmes2013, sanso_sideris2013}.
+$\delta \mathbf{g}_{P}$ \citep{barthelmes2013, sanso_sideris2013}.
 
@@
 both located at the same geodetic latitude and longitude
-(Fig. \ref{fig:surfaces}):
+(Figure \ref{fig:surfaces}):
 
 \begin{equation}
-\Delta \vector{g} = \vector{g}_{Q^{\prime}} - \vector{\gamma}_{Q} .
+\Delta \mathbf{g} = \mathbf{g}_{Q^{\prime}} - \boldsymbol{\gamma}_{Q} .
 \label{eq:gravity-anomaly-vector}
@@
 
-Similarly to the gravity disturbance (eq. \ref{eq:gravity-disturbance}),
+Similarly to the gravity disturbance (equation \ref{eq:gravity-disturbance}),
 the \textit{gravity anomaly} is given by
@@
 and $\gamma_{Q}$ is the magnitude of the normal gravity vector
-on the reference ellipsoid (Fig. \ref{fig:surfaces}).
+on the reference ellipsoid (Figure \ref{fig:surfaces}).
 
@@
 and $H_{P}$ is the orthometric height (i.e., with respect to the geoid)
-of point $P$ (Fig. \ref{fig:surfaces}).
+of point $P$ (Figure \ref{fig:surfaces}).
 The free-air correction can be interpreted as a downward continuation of
@@
 \noindent
-where $N \approx h - H$ is the geoidal height (Fig. \ref{fig:surfaces}).
+where $N \approx h - H$ is the geoidal height (Figure \ref{fig:surfaces}).
 This approximation assumes that the geoid and the surface of
@@
 \citep[e.g.,][]{torge2012, sanso_sideris2013}.
-Fig. \ref{fig:difference} shows the absolute difference between the
+Figure \ref{fig:difference} shows the absolute difference between the
 gravity disturbance and the free-anomaly globally calculated
-using eq. \ref{eq:difference-disturbance-free-air} and the spherical harmonic
+using equation \ref{eq:difference-disturbance-free-air} and the spherical harmonic
 model EIGEN-6c4 \citep{forste2014}.
@@
 the free-air anomaly in Brazil,
-according to eq. \ref{eq:difference-disturbance-free-air},
+according to equation \ref{eq:difference-disturbance-free-air},
 is $\approx 9.3$ mGal.
-
-\begin{figure}
-    \includegraphics[width=\columnwidth]{figures/difference.eps}
-    \caption{
-    Absolute difference between the gravity disturbance and free-air anomaly.
-    Calculated using eq. \ref{eq:difference-disturbance-free-air}, the
-    EIGEN-6c4 \citep{forste2014} spherical harmonic model, and software
-    \textit{Fatiando a Terra} \citep{uieda2013}. The figure was made using GMT
-    \citep{wessel2013}.
-    }
-  \label{fig:difference}
-\end{figure}
 
@@
 A similar correction can be applied to the gravity disturbance
-(using the approximation in eq. \ref{eq:gravity-disturbance-approx})
+(using the approximation in equation \ref{eq:gravity-disturbance-approx})
 to remove the effect of topographic masses between
@@
 Apart from the centrifugal effect mentioned in the previous section,
-the Bouguer anomaly (eq. \ref{eq:simple-bouguer-anomaly}) still contains
+the Bouguer anomaly (equation \ref{eq:simple-bouguer-anomaly}) still contains
 the gravitational effect of topographic masses between the geoid and ellipsoid.
@@
 
-\section{Gravity modelling}
+\section{Gravity modeling}
 
@@
 Now we will discuss how this relates to some assumptions made during gravity
-modelling.
+modeling.
 
-The gravity vector $\vector{g}_i$ (eq. \ref{eq:gravity-disturbance-vector})
-at a point $(x_i, y_i, z_i)$ in the TCS (Fig. \ref{fig:systems}c) can be
+The gravity vector $\mathbf{g}_i$ (equation \ref{eq:gravity-disturbance-vector})
+at a point $(x_i, y_i, z_i)$ in the TCS (Figure \ref{fig:systems}c) can be
 represented by:
@@
 \begin{equation}
-\vector{g}_i = \vector{\gamma}_i + \delta \vector{g}_i \: .
+\mathbf{g}_i = \mathbf{\gamma}_i + \delta \mathbf{g}_i \: .
 \label{eq:gravity-vector-TCS}
@@
 g_i \approx \gamma_i +
-\versor{\gamma}_i^{\top} \delta \vector{g}_i \: ,
+\mathbf{\gamma}_i^{\top} \delta \mathbf{g}_i \: ,
 \label{eq:gobs-approx}
@@
 \noindent
-where $^{\top}$ denotes transposition and $\versor{\gamma}_{i}$ is a unit
+where $^{\top}$ denotes transposition and $\mathbf{\gamma}_{i}$ is a unit
 vector with the direction of the normal gravity vector.
@@
 In local or regional scales, the unit
-vector $\versor{\gamma}_{i}$
+vector $\mathbf{\gamma}_{i}$
 may be considered constant throughout the study area and
-parallel to the $z$ axis of the TCS (Fig. \ref{fig:systems}c).
+parallel to the $z$ axis of the TCS (Figure \ref{fig:systems}c).
 
-Using eq. \ref{eq:gobs-approx} and fixing $\gamma$ at a point $P$ at
-the origin of the TCS (Fig. \ref{fig:systems}c),
+Using equation \ref{eq:gobs-approx} and fixing $\gamma$ at a point $P$ at
+the origin of the TCS (Figure \ref{fig:systems}c),
 the gravity disturbance can be rewritten as:
@@
 \begin{equation}
-\delta g_{i} \approx \versor{\gamma}_{P}^{\top} \delta \vector{g}_{i} \: .
+\delta g_{i} \approx \mathbf{\gamma}_{P}^{\top} \delta \mathbf{g}_{i} \: .
 \label{eq:gravity-disturbance-approx-TCS}
@@
 This equation shows that the gravity disturbance $\delta g_{i}$ is
-approximately the component of the $\delta \vector{g}_{i}$ in the direction of
-the normal gravity vector $\vector{\gamma}_P$
+approximately the component of the $\delta \mathbf{g}_{i}$ in the direction of
+the normal gravity vector $\mathbf{\gamma}_P$
 \citep{hofmann-wellenhof-moritz2005, sanso_sideris2013}.
-Because $z$-axis of the TCS is parallel to $\vector{\gamma}_P$
-(Fig. \ref{fig:surfaces}), the gravity disturbance can be defined as the
+Because $z$-axis of the TCS is parallel to $\mathbf{\gamma}_P$
+(Figure \ref{fig:surfaces}), the gravity disturbance can be defined as the
 $z$-component of the gravitational attraction exerted by the gravity sources at
@@
 d_{i} = \iiint\limits_{v}
-\frac{G \, \Delta\rho(x^{\prime}, y^{\prime}, z^{\prime}) \,
-[z_{i} - z^{\prime}]}
-{\sqrt{[x_{i} - x^{\prime}]^{2} +
-[y_{i} - y^{\prime}]^{2} + [z_{i} - z^{\prime}]^{2}}} \: dv^{\prime} \: ,
+\frac{(z_{i} - z^{\prime}) \, G \, \Delta\rho(x^{\prime}, y^{\prime}, z^{\prime})}
+{\sqrt{(x_{i} - x^{\prime})^{2} +
+		(y_{i} - y^{\prime})^{2} + (z_{i} - z^{\prime})^{2}}} \: dv^{\prime} \: ,
 \label{eq:gz-local}
@@
 
-In practice, most instances of gravity modelling in the literature use Bouguer
+In practice, most instances of gravity modeling in the literature use Bouguer
 anomalies and function $d_{i}$ (usually presented as ``$g_z$'') to represent
@@
 Rather, it depends on the relative position of the gravity sources and the
-observation point $(x_{i}, y_{i}, z_{i})$ in the TCS (Fig. \ref{fig:systems}c).
+observation point $(x_{i}, y_{i}, z_{i})$ in the TCS (Figure \ref{fig:systems}c).
 Finally, it is important to mention that gravimeters do not measure $d_{i}$ but
-$\| \vector{g}_i \|$.
+$\| \mathbf{g}_i \|$.
 
@@
 its use in applied geophysics is uncommon.
-For years, geophysicists have been using gravity anomalies for modelling
+For years, geophysicists have been using gravity anomalies for modeling
 the gravitational effects of geological bodies.
@@
 It is a question that underpins the very assumptions about what it is that we
-are actually modelling.
+are actually modeling.
 
@@
 
-More often than not, gravity modelling is done by computing the
+More often than not, gravity modeling is done by computing the
 vertical component of the gravitational acceleration exerted by the sources.
@@
 
-\begin{acknowledgments}
+%%%%%%%%%%%%%%%%%% Figures %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%\plot{cartesian-geodetic-systems}{width=0.6\textwidth}{
+\plot{cartesian-geodetic-systems}{width=\columnwidth}{
+	{(a) and (b) The Geocentric Cartesian System (GCS) and the Geocentric
+		Geodetic System (GGS).
+		The GGS is defined by an oblate ellipsoid with semi-minor axis $b$ and
+		semi-major axis $a$.
+		In this coordinate system, the position of a point is determined by the
+		geometric height $h$, geodetic latitude $\varphi$, and longitude $\lambda$.
+		$O$ is the Earth's center of mass, $P$ is an observation point, and
+		unit vectors $\mathbf{u}$, $\mathbf{v}$, and $\mathbf{w}$ define mutually
+		orthogonal directions at $P$ (equation \ref{eq:unit-vectors}).
+		In (b), $Q$ is the projection of $P$ onto the reference ellipsoid at
+		the same latitude and longitude.
+		(c) The topocentric Cartesian coordinate system (TCS) with origin at a
+		point $P$.
+		Axes $x$ and $y$ are parallel to the unit vectors $\mathbf{v}$ and
+		$\mathbf{w}$, respectively, and the $z$ axis is opposite to $\mathbf{u}$.
+		The gray plane is the same shown in (a) and (b).}
+	\label{fig:systems}
+}
+
+%\plot{surfaces}{width=0.6\textwidth}{
+\plot{surfaces}{width=\columnwidth}{
+	{The gravity vector
+		$\mathbf{g}_{P}$, normal gravity vector $\boldsymbol{\gamma}_{P}$,
+		gravity disturbance vector $\delta \mathbf{g}_{P}$
+		(equation \ref{eq:gravity-disturbance-vector}), and unit vector
+		$\mathbf{u}_{P}$ (equation \ref{eq:unit-vectors}) at a point $P$ 
+		on the surface of the Earth.
+		The gravity vector $\mathbf{g}_{Q^{\prime}}$
+		at a point $Q^{\prime}$ on the geoid, normal gravity vector
+		$\boldsymbol{\gamma}_{Q}$ at a point $Q$ on the reference ellipsoid,
+		geometric height $h$, orthometric height $H$, and geoidal height $N$.
+		The dashed line passing through
+		$Q$, $Q^{\prime}$ and $P$ is normal to the
+		surface of the reference ellipsoid at $Q$.
+		This figure shows a commonly used approximation in which the ellipsoid and
+		geoid are represented as parallel surfaces so that $h \approx H + N$.} 
+	\label{fig:surfaces}
+}
+
+%\plot{difference}{width=0.6\textwidth}{
+\plot{difference}{width=\columnwidth}{
+	{Absolute difference between the gravity disturbance and free-air anomaly.
+		Calculated using equation \ref{eq:difference-disturbance-free-air}, the
+		EIGEN-6c4 \citep{forste2014} spherical harmonic model, and software
+		\textit{Fatiando a Terra} \citep{uieda2013}. The figure was made using GMT
+		\citep{wessel2013}.}
+	\label{fig:difference}
+}
+
+%%%%%%%%%%%%%%%%%% Figures %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{ACKNOWLEDGMENTS}
+
 The authors would like to thank the editor and reviewers for their criticisms
-and corrections.
-All source code and data used to generate the results presented here are
-available at
+and corrections. All source code and data used to generate the results presented 
+here are available at
 \href{https://doi.org/10.5281/zenodo.1255306}{doi.org/10.5281/zenodo.1255306}
@@
 \href{https://github.com/pinga-lab/use-the-disturbance}{github.com/pinga-lab/use-the-disturbance}.
-\end{acknowledgments}
 
-\bibliographystyle{gji}
-\bibliography{bib-file}
+%\newpage
+
+\bibliographystyle{seg}  % style file is seg.bst
+
+\bibliography{bib-file.bib}
 

### 2018-06-20 manuscript/letter.tex
two-column and manuscript versions

@@
-\documentclass[paper,twocolumn,twoside]{geophysics}
+%\documentclass[paper,twocolumn,twoside]{geophysics}
 %\documentclass[manuscript,revised]{geophysics}
+\documentclass[manuscript]{geophysics}
 
@@
 
-%\plot{cartesian-geodetic-systems}{width=0.6\textwidth}{
-\plot{cartesian-geodetic-systems}{width=\columnwidth}{
+\plot{cartesian-geodetic-systems}{width=0.6\textwidth}{
+%\plot{cartesian-geodetic-systems}{width=\columnwidth}{
 	{(a) and (b) The Geocentric Cartesian System (GCS) and the Geocentric
@@
 
-%\plot{surfaces}{width=0.6\textwidth}{
-\plot{surfaces}{width=\columnwidth}{
+\plot{surfaces}{width=0.6\textwidth}{
+%\plot{surfaces}{width=\columnwidth}{
 	{The gravity vector
@@
 
-%\plot{difference}{width=0.6\textwidth}{
-\plot{difference}{width=\columnwidth}{
+\plot{difference}{width=0.6\textwidth}{
+%\plot{difference}{width=\columnwidth}{
 	{Absolute difference between the gravity disturbance and free-air anomaly.

### 2018-06-20 manuscript/letter.tex
insert missing boldsymbol gamma

@@
 \begin{equation}
-\mathbf{g}_i = \mathbf{\gamma}_i + \delta \mathbf{g}_i \: .
+\mathbf{g}_i = \boldsymbol{\gamma}_i + \delta \mathbf{g}_i \: .
 \label{eq:gravity-vector-TCS}
@@
 \noindent
-Furthermore, the gravity $g_i$ can be approximated by a first order Taylor's
+Furthermore, the gravity $g_i = \| \mathbf{g}_i \|$ can be approximated by a first order Taylor's
 expansion \citep{sanso_sideris2013}:
@@
 g_i \approx \gamma_i +
-\mathbf{\gamma}_i^{\top} \delta \mathbf{g}_i \: ,
+\hat{\boldsymbol{\gamma}}_i^{\top} \delta \mathbf{g}_i \: ,
 \label{eq:gobs-approx}
@@
 \noindent
-where $^{\top}$ denotes transposition and $\mathbf{\gamma}_{i}$ is a unit
+where $^{\top}$ denotes transposition, $\gamma_i = \| \boldsymbol{\gamma}_i \|$ 
+and $\hat{\boldsymbol{\gamma}}_{i}$ is a unit
 vector with the direction of the normal gravity vector.
@@
 In local or regional scales, the unit
-vector $\mathbf{\gamma}_{i}$
+vector $\hat{\boldsymbol{\gamma}}_{i}$
 may be considered constant throughout the study area and
@@
 
-Using equation \ref{eq:gobs-approx} and fixing $\gamma$ at a point $P$ at
+Using equation \ref{eq:gobs-approx} and fixing 
+$\hat{\boldsymbol{\gamma}}_{i}$ at a point $P$ at
 the origin of the TCS (Figure \ref{fig:systems}c),
@@
 \begin{equation}
-\delta g_{i} \approx \mathbf{\gamma}_{P}^{\top} \delta \mathbf{g}_{i} \: .
+\delta g_{i} \approx \hat{\boldsymbol{\gamma}}_{P}^{\top} \delta \mathbf{g}_{i} \: .
 \label{eq:gravity-disturbance-approx-TCS}
@@
 approximately the component of the $\delta \mathbf{g}_{i}$ in the direction of
-the normal gravity vector $\mathbf{\gamma}_P$
+the normal gravity vector $\boldsymbol{\gamma}_P$
 \citep{hofmann-wellenhof-moritz2005, sanso_sideris2013}.
-Because $z$-axis of the TCS is parallel to $\mathbf{\gamma}_P$
+Because $z$-axis of the TCS is parallel to $\boldsymbol{\gamma}_P$
 (Figure \ref{fig:surfaces}), the gravity disturbance can be defined as the

### 2018-07-03 manuscript/letter.tex
remove author identification and repository links

@@
 
-\address{
-\footnotemark[1]Observat\'{o}rio Nacional, \\
-Department of Geophysics, \\
-Rio de Janeiro, Brazil \\
-\footnotemark[2]University of Hawai'i at M\={a}noa, \\
-Department of Geology and Geophysics, SOEST, \\
-Honolulu, USA}
-\author{Vanderlei C. Oliveira Jr\footnotemark[1], Leonardo Uieda\footnotemark[2],
-Kristoffer A. T. Hallam\footnotemark[1] and Val\'{e}ria C. F. Barbosa\footnotemark[1]}
+%remove author identification for double-blind review process
+%\address{
+%\footnotemark[1]Observat\'{o}rio Nacional, \\
+%Department of Geophysics, \\
+%Rio de Janeiro, Brazil \\
+%\footnotemark[2]University of Hawai'i at M\={a}noa, \\
+%Department of Geology and Geophysics, SOEST, \\
+%Honolulu, USA}
+%\author{Vanderlei C. Oliveira Jr\footnotemark[1], Leonardo Uieda\footnotemark[2],
+%Kristoffer A. T. Hallam\footnotemark[1] and Val\'{e}ria C. F. Barbosa\footnotemark[1]}
+%\footer{vanderlei@on.br, leouieda@gmail.com, kristoffer.hallam@gmail.com, valcris@on.br}
+%\lefthead{Oliveira Jr et al.}
 
-\footer{vanderlei@on.br, leouieda@gmail.com, kristoffer.hallam@gmail.com, valcris@on.br}
-\lefthead{Oliveira Jr et al.}
 \righthead{Gravity anomaly or gravity disturbance?}
@@
 represents a purely gravitational effect.
-As a consequence of the superposition principal \citep{blakely1996},
+As a consequence of the superposition principle,
 this effect is caused by contrasts between
@@
 
-\section{ACKNOWLEDGMENTS}
-
-The authors would like to thank the editor and reviewers for their criticisms
-and corrections. All source code and data used to generate the results presented 
-here are available at
-\href{https://doi.org/10.5281/zenodo.1255306}{doi.org/10.5281/zenodo.1255306}
-and
-\href{https://github.com/pinga-lab/use-the-disturbance}{github.com/pinga-lab/use-the-disturbance}.
+%\section{ACKNOWLEDGMENTS}
+%
+%The authors would like to thank the editor and reviewers for their criticisms
+%and corrections. All source code and data used to generate the results presented 
+%here are available at
+%remove author identification for double-blind review process
+%\href{https://doi.org/10.5281/zenodo.1255306}{doi.org/10.5281/zenodo.1255306}
+%\textbf{link for online repository 1}
+%and
+%remove author identification for double-blind review process
+%\href{https://github.com/pinga-lab/use-the-disturbance}{github.com/pinga-lab/use-the-disturbance}
+%\textbf{link for online repository 2}.
 

### 2018-08-29 manuscript/letter.tex
include references of Vajda

@@
 \citep{lafehr1991,chapin1996,li2001,fairhead2003,
-hackney-featherstone2003,hinze2005} and maybe that is why it is
+hackney-featherstone2003,hinze2005, vajda-etal2006, vajda-etal2007, 
+vajda-etal2008} and maybe that is why it is
 unfamiliar to most exploration geophysicists.

### 2018-08-29 code/coord.py
start simulation of gravity anomaly

@@ -1,0 +1,481 @@
+from __future__ import division, absolute_import
+import numpy as np
+
+def geodetic2cartesian(height, latitude, longitude,
+                       major_semiaxis, minor_semiaxis):
+    '''
+    Transform geocentric geodetic coordinates into geocentric
+    Cartesian coordinates.
+
+    input
+
+    height: numpy array 1D - vector containing the geometric height (in meters).
+    latitude: numpy array 1D - vector containing the latitude (in degrees).
+    longitude: numpy array 1D - vector containing the longitude (in degrees).
+    major_semiaxis: float - major semiaxis of the reference
+                    ellipsoid (in meters).
+    minor_semiaxis: float - minor semiaxis of the reference
+                    ellipsoid (in meters).
+
+    output
+
+    x: numpy array 1D - vector containing the x component of the Cartesian
+       coordinates (in meters).
+    y: numpy array 1D - vector containing the y component of the Cartesian
+       coordinates (in meters).
+    z: numpy array 1D - vector containing the z component of the Cartesian
+       coordinates (in meters).
+    '''
+
+    h = np.asarray(height)
+    lat = np.asarray(latitude)
+    lon = np.asarray(longitude)
+
+    assert (h.size == lat.size == lon.size), 'height, latitude \
+and longitude must have the same number of elements'
+    assert (major_semiaxis > minor_semiaxis), 'major_semiaxis must be greater \
+than the minor_semiaxis'
+
+
+    #Prime vertical radius of curvature
+    N = prime_vertical_curv(major_semiaxis, minor_semiaxis, lat)
+
+    # convert degrees to radians
+    lat = np.deg2rad(lat)
+    lon = np.deg2rad(lon)
+
+    aux = N + height
+
+    # squared first eccentricity
+    e2 = (major_semiaxis**2. - minor_semiaxis**2.)/(major_semiaxis**2.)
+
+    clat = np.cos(lat)
+    slat = np.sin(lat)
+    clon = np.cos(lon)
+    slon = np.sin(lon)
+
+    x = aux*clat*clon
+    y = aux*clat*slon
+    z = (N*(1 - e2) + height)*slat
+
+    return x, y, z
+
[…]

### 2018-08-29 code/test_coord.py

@@ -1,0 +1,582 @@
+from __future__ import division, absolute_import
+import numpy as np
+from code.cartography import coord
+from numpy.testing import assert_almost_equal
+from pytest import raises
+
+
+def test_prime_radius_bad_semiaxes():
+    'major semiaxis lower than minor semiaxis'
+    latitude = np.ones(100)
+    raises(AssertionError, coord.prime_vertical_curv, 10, 23, latitude)
+
+
+def test_prime_radius_negative_semiaxes():
+    'null semiaxis'
+    latitude = np.ones(100)
+    raises(AssertionError, coord.prime_vertical_curv, -10, -23, latitude)
+
+
+def test_meridian_radius_bad_semiaxes():
+    'major semiaxis lower than minor semiaxis'
+    latitude = np.ones(100)
+    raises(AssertionError, coord.meridian_curv, 10, 23, latitude)
+
+
+def test_meridian_radius_negative_semiaxes():
+    'null semiaxis'
+    latitude = np.ones(100)
+    raises(AssertionError, coord.meridian_curv, -10, -23, latitude)
+
+
+def test_relationship_curvatures():
+    'verify relationship between the curvatures'
+    latitude = np.linspace(-90, 90, 181)
+    a = 6378137.0
+    f = 1.0/298.257223563
+    b = a*(1-f)
+    N = coord.prime_vertical_curv(a, b, latitude)
+    M = coord.meridian_curv(a, b, latitude)
+    e2 = (a*a - b*b)/(a*a)
+    sin2lat = np.sin(np.deg2rad(latitude))
+    sin2lat *= sin2lat
+    M_relationship = ((1 - e2)/(1 - e2*sin2lat))*N
+    assert_almost_equal(M, M_relationship, decimal=8)
+
+
+def test_prime_radius_known_input():
+    'verify results obtained for known input'
+    a = 6378137.0
+    f = 1.0/298.257223563
+    b = a*(1-f)
+    e2 = (a*a - b*b)/(a*a)
+    N_true_0 = a
+    N_true_90 = a/np.sqrt(1 - e2)
+    N_calc_0 = coord.prime_vertical_curv(a, b, 0)
+    N_calc_90 = coord.prime_vertical_curv(a, b, 90)
+    assert_almost_equal(N_true_0, N_calc_0, decimal=15)
+    assert_almost_equal(N_true_90, N_calc_90, decimal=15)
+
+
[…]

### 2018-08-30 code/coord.py
transform geodetic coords into topoc Cart coords

@@ -1,11 +1,113 @@
-from __future__ import division, absolute_import
 import numpy as np
+import warnings
 
-def geodetic2cartesian(height, latitude, longitude,
+
+def distance(xi, yi, zi, x0, y0, z0):
+    '''
+    Compute the Euclidean distances between a point (xi, yi, zi) and a set of
+    points whose Cartesian coordinates are stored in the vectors x0, y0 and z0.
+    input
+    xi, yi, zi: floats - Cartesian coordinates of a point.
+    x0, y0, z0: numpy arrays 1D - vectors containing Cartesian coordinates.
+    output
+    r: numpy array 1D - Euclidean distances.
+    '''
+
+    assert x0.size == y0.size == z0.size, 'x0, y0 and z0 must have the same \
+number of elements'
+
+    assert x0.shape == y0.shape == z0.shape, 'x0, y0 and z0 must have the same \
+shape'
+
+    #the asserts below do not work if xi, yi and/or zi is int
+    #assert isinstance(xi, (int, float)), 'xi must be a float'
+    #assert isinstance(yi, (int, float)), 'yi must be a float'
+    #assert isinstance(zi, (int, float)), 'zi must be a float'
+    assert isinstance(1.*xi, float), 'xi must be a float'
+    assert isinstance(1.*yi, float), 'yi must be a float'
+    assert isinstance(1.*zi, float), 'zi must be a float'
+
+    X = xi - x0
+    Y = yi - y0
+    Z = zi - z0
+    r = np.sqrt(X*X + Y*Y + Z*Z)
+    return r
+
+
+def squared_distance(xi, yi, zi, x0, y0, z0):
+    '''
+    Compute the squared Euclidean distances between a point (xi, yi, zi) and a
+    set of points whose Cartesian coordinates are stored in the vectors
+    x0, y0 and z0.
+    input
+    xi, yi, zi: floats - Cartesian coordinates of a point.
+    x0, y0, z0: numpy arrays 1D - vectors containing Cartesian coordinates.
+    output
+    r2: numpy array 1D - squared Euclidean distances.
+    '''
+
+    assert x0.size == y0.size == z0.size, 'x0, y0 and z0 must have the same \
+number of elements'
+
+    assert x0.shape == y0.shape == z0.shape, 'x0, y0 and z0 must have the same \
+shape'
+
+    #the asserts below do not work if xi, yi and/or zi is int
+    #assert isinstance(xi, (int, float)), 'xi must be a float'
[…]

### 2018-08-30 code/elipsoids.py

@@ -1,0 +1,61 @@
+def WGS84():
+    '''
+    This function returns the following parameters defining the 
+    reference elipsoid WGS84:
+    a = semimajor axis [m]
+    f = flattening
+    
+    output:
+    a, f
+    '''
+    a = 6378137.0
+    f = 1.0/298.257223563
+    
+    return a, f
+    
+def SAD69():
+    '''
+    This function returns the following parameters defining the 
+    reference elipsoid SAD69:
+    a = semimajor axis [m]
+    f = flattening
+    
+    output:
+    a, f
+    '''
+    a = 6378160.0
+    f = 1.0/298.25
+    
+    return a, f
+    
+def Hayford1924():
+    '''
+    This function returns the following parameters defining the 
+    International (Hayford's) reference elipsoid of 1924:
+    a = semimajor axis [m]
+    f = flattening
+    
+    output:
+    a, f
+    '''
+    a = 6378388.0
+    f = 1.0/297.0
+    
+    return a, f
+    
+def SAD69_WGS84():
+    '''
+    Transformation parameters from local geodetic system
+    SAD69 to WGS84.
+    
+    output
+    dx: float - origin translation along the x-axis (in meters).
+    dy: float - origin translation along the y-axis (in meters).
+    dz: float - origin translation along the z-axis (in meters).
+    '''
+    
+    dx = -57.
+    dy = 1.
+    dz = -41.
+    
+    return dx, dy, dz

### 2018-08-30 code/test_coord.py

@@ -1,582 +1,0 @@
-from __future__ import division, absolute_import
-import numpy as np
-from code.cartography import coord
-from numpy.testing import assert_almost_equal
-from pytest import raises
-
-
-def test_prime_radius_bad_semiaxes():
-    'major semiaxis lower than minor semiaxis'
-    latitude = np.ones(100)
-    raises(AssertionError, coord.prime_vertical_curv, 10, 23, latitude)
-
-
-def test_prime_radius_negative_semiaxes():
-    'null semiaxis'
-    latitude = np.ones(100)
-    raises(AssertionError, coord.prime_vertical_curv, -10, -23, latitude)
-
-
-def test_meridian_radius_bad_semiaxes():
-    'major semiaxis lower than minor semiaxis'
-    latitude = np.ones(100)
-    raises(AssertionError, coord.meridian_curv, 10, 23, latitude)
-
-
-def test_meridian_radius_negative_semiaxes():
-    'null semiaxis'
-    latitude = np.ones(100)
-    raises(AssertionError, coord.meridian_curv, -10, -23, latitude)
-
-
-def test_relationship_curvatures():
-    'verify relationship between the curvatures'
-    latitude = np.linspace(-90, 90, 181)
-    a = 6378137.0
-    f = 1.0/298.257223563
-    b = a*(1-f)
-    N = coord.prime_vertical_curv(a, b, latitude)
-    M = coord.meridian_curv(a, b, latitude)
-    e2 = (a*a - b*b)/(a*a)
-    sin2lat = np.sin(np.deg2rad(latitude))
-    sin2lat *= sin2lat
-    M_relationship = ((1 - e2)/(1 - e2*sin2lat))*N
-    assert_almost_equal(M, M_relationship, decimal=8)
-
-
-def test_prime_radius_known_input():
-    'verify results obtained for known input'
-    a = 6378137.0
-    f = 1.0/298.257223563
-    b = a*(1-f)
-    e2 = (a*a - b*b)/(a*a)
-    N_true_0 = a
-    N_true_90 = a/np.sqrt(1 - e2)
-    N_calc_0 = coord.prime_vertical_curv(a, b, 0)
-    N_calc_90 = coord.prime_vertical_curv(a, b, 90)
-    assert_almost_equal(N_true_0, N_calc_0, decimal=15)
-    assert_almost_equal(N_true_90, N_calc_90, decimal=15)
-
-
[…]

### 2018-08-30 code/gamma.py
algorithm for computing a synthetic Geoid

@@ -1,0 +1,101 @@
+import numpy as np
+
+def gamma_somigliana(a, f, GM, omega, phi):
+    '''
+    This function calculates the normal gravity by using
+    the Somigliana's formula.
+    
+    input:
+    a: float containing the semimajor axis [m]
+    f: float containing the flattening
+    GM: float containing the geocentric gravitational constant 
+        of the Earth (including the atmosphere) [m**3/s**2]
+    omega: float containing the angular velocity [rad/s]
+    phi: array containing the geodetic latitudes [degree]
+    
+    output:
+    gamma: array containing the values of normal gravity
+           on the surface of the elipsoid for each geodetic
+           latitude [mGal]
+    '''
+    b = a*(1.0-f)
+    a2 = a**2
+    b2 = b**2
+    E = np.sqrt(a2 - b2)
+    elinha = E/b
+    bE = b/E
+    Eb = E/b
+    atg = np.arctan(Eb)
+    q0 = 0.5*((1+3*(bE**2))*atg - (3*bE))
+    q0linha = 3.0*(1+(bE**2))*(1-(bE*atg)) - 1
+    m = (omega**2)*(a2)*b/GM
+    aux = elinha*q0linha/q0
+    gammaa = (GM/(a*b))*(1-m-(m/6.0)*aux)
+    gammab = (GM/a2)*(1+(m/3.0)*aux)
+    aux = np.deg2rad(phi)
+    s2 = np.sin(aux)**2
+    c2 = np.cos(aux)**2
+    # the 10**5 converts from m/s**2 to mGal
+    gamma = (10**5)*((a*gammaa*c2) + (b*gammab*s2))/np.sqrt((a2*c2) + (b2*s2))
+    return gamma
+    
+def gamma_closedform(a, f, GM, omega, phi, h):
+    '''
+    This function calculates the normal gravity by using
+    a closed-form formula.
+    
+    input:
+    a: float containing the semimajor axis [m]
+    f: float containing the flattening
+    GM: float containing the geocentric gravitational constant 
+        of the Earth (including the atmosphere) [m**3/s**-2]
+    omega: float containing the angular velocity [rad/s]
[…]
